### 2014-05-08 writeup/paper.tex
added writeup tex

% TEMPLATE for Usenix papers, specifically to meet requirements of
%  USENIX '05
% originally a template for producing IEEE-format articles using LaTeX.
%   written by Matthew Ward, CS Department, Worcester Polytechnic Institute.
% adapted by David Beazley for his excellent SWIG paper in Proceedings,
%   Tcl 96
% turned into a smartass generic template by De Clarke, with thanks to
%   both the above pioneers
% use at your own risk.  Complaints to /dev/null.
% make it two column with no page numbering, default is 10 point

% Munged by Fred Douglis <douglis@research.att.com> 10/97 to separate
% the .sty file from the LaTeX source template, so that people can
% more easily include the .sty file into an existing document.  Also
% changed to more closely follow the style guidelines as represented
% by the Word sample file.

% Note that since 2010, USENIX does not require endnotes. If you want
% foot of page notes, don't include the endnotes package in the 
% usepackage command, below.

% This version uses the latex2e styles, not the very ancient 2.09 stuff.
\documentclass[letterpaper,10pt]{article}
\usepackage{epsfig,graphicx,usenix,fullpage,float,hyperref}
\usepackage{mathtools,amssymb,setspace, pbox}
\usepackage[center, labelfont=bf]{caption}

\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}
\def\bE{\mathbb{E}}

%\usepackage{endnotes}
\begin{document}

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf 6.824 Final Project}
%for single author (just remove % characters)
\author{
{\rm Colleen Josephson}\\
cjoseph@mit.edu
\and
{\rm Joseph DelPreto}\\
delpreto@mit.ed
\and
{\rm Pranjal Vachaspati}\\
pranjal@mit.edu
\and
{\rm Steven Valdez}\\
dvorak42@mit.edu
% copy the following lines to add more authors
% \and
% {\rm Name}\\
%Name Institution
} % end author

\date{May 11, 2014}

\maketitle

% Use the following at camera-ready time to suppress page numbers.
% Comment it out when you first submit the paper for review.
%\thispagestyle{empty}

\section{Introduction}
This document is the writeup of our 6.824 final project.

The goals of the project are:

\section{Design} \label{sec:design}

\subsection{Paxos Optimizations}

\subsection{Persistent Storage}

\subsubsection{Recovery from failures}

\section{Performance}

We created an extensive system to try and accurately measure performance. In addition to writing many benchmarks, we also modified the original RPC system so that we could run the servers in an AWS cluster and observe the performance on a real network.

\subsection{Throughput and Latency}

\subsection{Bottlenecks}
Although we made significant improvements performance-wise, we identified a number of bottlenecks.

\subsubsection{Paxos}

\subsubsection{Disk Recovery}

\section{Conclusion}
We did a project.

%{\footnotesize \bibliographystyle{acm}
%\bibliography{paper}}
%\newpage
%\section{\\Appendix A: Alternative Emergency Alert Scheme}
% the \\ insures the section title is centered below the phrase: Appendix B
% See Appendix A for  (NOTE: perhaps the base station could note the empty 
%space, and then take advantage of the periodicity to send a broadcast packet 
%that would allow sensors to opportunistically fill in the blank, thus 
%increasing the success rate).(NOTE: opportunistic make up listed in last note
%could allow for the two sensors to randomly switch to open slots, potentially
% getting rid of the collision state). (NOTE: unless the fill-in-the-blank
%scheme is uses, in which case the base station will be fully aware of any 
%re-mappings as it dictated them in the first place!).
%\newpage
\end{document}

### 2014-05-08 src/paxos/test_test.go
Added some behavioral flags for shardkv

@@ -1085,7 +1085,7 @@ func TestFileRPCCountPrePrepare(test *testing.T) {
 
 	// TODO figure out what expected count should be
 	// will probably depend on chosen implementation
-	fmt.Printf("RPC count: %v (expected max ?)", rpcCount)
+	fmt.Printf("\n\tRPC count: %v (expected max ?)", rpcCount)
 
 	fmt.Printf("\n\tPassed")
 }

### 2014-05-08 src/shardkv/server.go

@@ -8,6 +8,7 @@ import "time"
 import "paxos"
 import "sync"
 import "os"
+
 //import "io"
 import "syscall"
 import "encoding/gob"
@@ -22,11 +23,17 @@ import "strings"
 const Debug = 1
 const DebugPersist = 1
 const printRPCerrors = false
-const Log = 1
+const Log = 0
+
 var logfile *os.File
 
 const persistent = true
 const recovery = true
+const writeToMemory = true    // Whether responses/store should be written to memory (as well as disk / disk cache)
+const dbUseCache = true       // Whether database should use a built-in cache
+const dbUseCompression = true // Whether database should compress entries
+
+//const memoryLimit = 1000000000 // Memory limit in bytes (not used yet)
 
 func DPrintf(format string, a ...interface{}) (n int, err error) {
 	if Debug > 0 {
@@ -84,13 +91,14 @@ type ShardKV struct {
 	recovering     bool
 }
 
-// Get the desired response, either from memory or disk
-func (kv *ShardKV) getResponse(id int64) (string, bool) {
-	response, exists := kv.response[id]
-	if !exists {
-		response, exists = kv.dbGetResponse(id)
+// Write the desired key/value to memory and/or disk
+func (kv *ShardKV) putValue(key string, value string) {
+	// Write to memory if using memory
+	if writeToMemory {
+		kv.store[key] = value
 	}
-	return response, exists
+	// Write to disk if persistent is enabled
+	kv.dbPut(key, value)
 }
 
 // Get the desired value, either from memory or disk
@@ -100,6 +108,25 @@ func (kv *ShardKV) getValue(key string) (string, bool) {
 		value, exists = kv.dbGet(key)
 	}
 	return value, exists
+}
+
+// Write the desired response to memory and/or disk
+func (kv *ShardKV) putResponse(id int64, value string) {
+	// Write to memory if using memory
+	if writeToMemory {
+		kv.response[id] = value
+	}
+	// Write to disk if persistent is enabled
+	kv.dbWriteResponse(id, value)
+}
+
+// Get the desired response, either from memory or disk
[…]

### 2014-05-08 src/shardmaster/server.go

@@ -12,14 +12,16 @@ import "encoding/gob"
 import "math/rand"
 import "time"
 import "strconv"
+
 //import "io"
 import "github.com/jmhodges/levigo"
 import "bytes"
 
-const Debug = 1
-const DebugPersist = 1
-const printRPCerrors = true
-const Log = 1
+const Debug = 0
+const DebugPersist = 0
+const printRPCerrors = false
+const Log = 0
+
 var logfile *os.File
 
 // Note: if persistent and recovery are not enabled,
@@ -397,12 +399,11 @@ func (sm *ShardMaster) Move(args *MoveArgs, reply *MoveReply) error {
 
 // Respond to a query about a particular configuration
 func (sm *ShardMaster) Query(args *QueryArgs, reply *QueryReply) error {
-	DPrintf("%d) Query: %d\n", sm.me, args.Num) 
+	DPrintf("%d) Query: %d\n", sm.me, args.Num)
 	for sm.recovering && !sm.dead {
 		time.Sleep(10 * time.Millisecond)
 	}
 	sm.mu.Lock()
-
 
 	newOp := Op{1, int64(args.Num), nil, 0}
 
@@ -669,7 +670,7 @@ func (sm *ShardMaster) dbInit() {
 	DPrintfPersist("\n\t%v: DB Name: %s", sm.me, sm.dbName)
 	var err error
 	sm.db, err = levigo.Open(sm.dbName, sm.dbOpts)
-	enableLog()//needs to be here, otherwise logging stops working after
+	enableLog() //needs to be here, otherwise logging stops working after
 	if err != nil {
 		DPrintfPersist("\n\t%v: Error opening database! \n\t%s", sm.me, fmt.Sprint(err))
 	} else {
@@ -826,7 +827,7 @@ func StartServer(servers []string, me int, network bool) *ShardMaster {
 	if Log == 1 {
 		//set up logging
 		os.Remove("shardmaster.log")
-		logfile, err = os.OpenFile("shardmaster.log", os.O_RDWR | os.O_CREATE | os.O_APPEND | os.O_SYNC, 0666)
+		logfile, err = os.OpenFile("shardmaster.log", os.O_RDWR|os.O_CREATE|os.O_APPEND|os.O_SYNC, 0666)
 
 		if err != nil {
 			log.Fatalf("error opening file: %v", err)
@@ -860,7 +861,7 @@ func StartServer(servers []string, me int, network bool) *ShardMaster {
 	if !printRPCerrors {
[…]

### 2014-05-08 src/paxos/paxos.go
Paxos fixes

@@ -44,7 +44,7 @@ const recovery = false
 const Debug = 0
 const DebugPersist = 0
 
-const enableLeader = 1
+const enableLeader = 0
 
 func DPrintf(format string, a ...interface{}) (n int, err error) {
 	if Debug > 0 {
@@ -72,6 +72,7 @@ type Proposal struct {
 
 type Paxos struct {
 	mu        sync.Mutex
+	fu        sync.Mutex
 	l         net.Listener
 	dead      bool
 	dbDeleted bool
@@ -332,10 +333,18 @@ func (px *Paxos) Prepare(args *PrepareArgs, reply *PrepareReply) error {
 
 	// Check if proposal number is high enough
 	if args.PID > prop.Prepare {
-    px.instances[args.Instance] = Proposal{args.PID, prop.Accept, prop.Value, prop.Decided}
+    if prop.Decided {
+      px.instances[args.Instance] = Proposal{args.PID + 1, prop.Accept, prop.Value, prop.Decided}
+    } else {
+      px.instances[args.Instance] = Proposal{args.PID, prop.Accept, prop.Value, prop.Decided}
+    }
     px.dbWriteInstance(args.Instance, px.instances[args.Instance])
     reply.Err = false
-    reply.PID = prop.Accept
+    if prop.Decided {
+      reply.PID = args.PID + 1
+    } else {
+      reply.PID = prop.Accept
+    }
     reply.Value = prop.Value
     reply.Leader = args.Server
     reply.Decided = prop.Decided

### 2014-05-09 src/paxos/paxos.go
Fixing paxos opts

@@ -44,7 +44,7 @@ const recovery = false
 const Debug = 0
 const DebugPersist = 0
 
-const enableLeader = 1
+const enableLeader = 0
 
 func DPrintf(format string, a ...interface{}) (n int, err error) {
 	if Debug > 0 {
@@ -367,7 +367,7 @@ func (px *Paxos) Accept(args *AcceptArgs, reply *AcceptReply) error {
 		newDone[dk] = dv
 	}
 
-  if prop.Decided && args.Value != prop.Value && enableLeader > 0 {
+  if enableLeader > 0 && prop.Decided && args.Value != prop.Value {
     DPrintf("\n%v (L%v): Received accept for dead", px.me, px.leader)
     reply.PID = args.PID
   } else if args.PID >= prop.Prepare && (args.Server == px.leader[args.Instance] || enableLeader == 0) {
@@ -505,7 +505,7 @@ func (px *Paxos) Propose(args *ProposeArgs, reply *ProposeReply) error {
               hValue = reply.Value
             }
           } 
-          if reply.PID == hPID && !hDecided {
+          if reply.PID == hPID && !hDecided && enableLeader > 0 {
             if _,ok := hValuePrime[reply.Value]; ok {
               hValuePrime[reply.Value] += 1
             } else {
@@ -516,14 +516,16 @@ func (px *Paxos) Propose(args *ProposeArgs, reply *ProposeReply) error {
       }
     }
     
+    if enableLeader > 0 {
     hValueCount := -1
     for k,v := range hValuePrime {
-      if v > hValueCount {
+      if v > hValueCount && k != nil {
         if !hDecided {
           hValue = k
         }
         hValueCount = v
       }
+    }
     }
 
 		// If prepare was rejected, start over with new proposal value
@@ -728,6 +730,7 @@ func (px *Paxos) Status(seq int) (bool, interface{}) {
 	px.mu.Lock()
 	prop := px.getInstance(seq)
 	px.mu.Unlock()
+
 	return prop.Decided, prop.Value
 }
 

### 2014-05-09 src/paxos/paxos.go
changed paxos port...localhost tests probably no longer work.

@@ -1195,7 +1195,7 @@ func Make(peers []string, me int, rpcs *rpc.Server, network bool, tag string) *P
 		// prepare to receive connections from clients.
 		// change "unix" to "tcp" to use over a network.
 		if px.network {
-			l, e := net.Listen("tcp", ":"+strconv.Itoa(startport+me))
+			l, e := net.Listen("tcp", ":"+strconv.Itoa(startport))
 			if e != nil {
 				log.Fatal("listen error: ", e)
 			}

### 2014-05-09 src/shardkv/server.go
disabled debug

@@ -20,8 +20,8 @@ import "github.com/jmhodges/levigo"
 import "bytes"
 import "strings"
 
-const Debug = 1
-const DebugPersist = 1
+const Debug = 0
+const DebugPersist = 0
 const printRPCerrors = false
 const Log = 0
 

### 2014-05-09 src/test/test_test.go
added battery test

@@ -338,3 +338,44 @@ func BenchmarkClientLatencyManyShard(benchmark *testing.B) {
 		kvClerk.Put(strconv.Itoa(rand.Int()), strconv.Itoa(rand.Int()))
 	}
 }
+
+
+func TestManyClientOneShard(t *testing.T) {
+	numGroups := 3
+	numReplicas := 3
+	nclients := 45
+	nseconds := 10
+	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
+	//defer clean()
+
+	fmt.Printf("\nBenchmark: many clients, one shards...\n")
+
+	
+	smClerk := shardmaster.MakeClerk(smPorts, true)
+	for i := 0; i < len(gids); i++ {
+		smClerk.Join(gids[i], kvPorts[i])
+	}
+	
+	counts := make([]int, nclients)
+	for i := 0; i < nclients; i++ {
+		go func(i int) {
+			ck := shardkv.MakeClerk(smPorts, true)
+			tStart := time.Now()
+			count := 0
+			for time.Since(tStart).Seconds() < float64(nseconds) {
+				ck.Put(strconv.Itoa(rand.Int()), 
+					strconv.Itoa(rand.Int()))
+				count++
+			}
+			counts[i] = count
+		}(i)
+	}
+	toWait := 12 * time.Second
+	time.Sleep(toWait)
+	
+	tot := 0
+	for _,c := range(counts) {
+		tot += c
+	}
+	fmt.Printf("%f operations per second\n", float64(tot)/float64(nseconds))
+}

### 2014-05-09 writeup/paper.tex
added some benchmark results to writeup (unformatted)

@@
 \subsection{Throughput and Latency}
+Avg client latency w/ out paxos optimizations:
+one shard, 3 groups w/ 3 replicas = 15.4ms/op
+many shards, 3 groups w/ 3 replicas = 21.8ms/op
+
+Avg client throughput w/ out paxos optimizations:
+one shard, 3 groups w/ 3 replicas = 65 requests/second
+many shards, 3 groups w/ 3 replicas = 46 requets/second
+
+Client requests per second:
+3 groups w/ 3 replicas:
+~179 requests per second (saturates at around 50 users)
 
@@
 Although we made significant improvements performance-wise, we identified a number of bottlenecks.
+
+One notable bottleneck is disk writes. The database library we used,
+levedb, uses a cache and is quite fast. However, in the process of
+debugging networked RPCs, we implemented simple logging. This logging
+slowed down the operations by nearly a factor of ten. This was a good
+demonstration of how care must be taken in even the most mundane
+aspects, such as logging.
 
@@
 %\newpage
-%\section{\\Appendix A: Alternative Emergency Alert Scheme}
-% the \\ insures the section title is centered below the phrase: Appendix B
-% See Appendix A for  (NOTE: perhaps the base station could note the empty 
-%space, and then take advantage of the periodicity to send a broadcast packet 
-%that would allow sensors to opportunistically fill in the blank, thus 
-%increasing the success rate).(NOTE: opportunistic make up listed in last note
-%could allow for the two sensors to randomly switch to open slots, potentially
-% getting rid of the collision state). (NOTE: unless the fill-in-the-blank
-%scheme is uses, in which case the base station will be fully aware of any 
-%re-mappings as it dictated them in the first place!).
-%\newpage
 \end{document}

### 2014-05-09 src/paxos/paxos.go
Fixing paxos

@@ -44,7 +44,7 @@ const recovery = true
 const Debug = 0
 const DebugPersist = 0
 
-const enableLeader = 0
+const enableLeader = 1
 
 func DPrintf(format string, a ...interface{}) (n int, err error) {
 	if Debug > 0 {
@@ -138,6 +138,7 @@ type AcceptArgs struct {
 	Instance int
 	PID      int
 	Value    interface{}
+  Decided  bool
 	Done     map[int]int
 	Leader   int
 }
@@ -357,7 +358,7 @@ func (px *Paxos) Accept(args *AcceptArgs, reply *AcceptReply) error {
 		px.recordDone(dk, dv)
 	}
 	px.mu.Lock()
-	DPrintf("\n%v (L%v): Received accept for sequence %v (%v)", px.me, px.leader, args.Instance, args.Value)
+	DPrintf("\n%v (L%v): Received accept for sequence %v (%v)", px.me, px.leader[args.Instance], args.Instance, args.Value)
 	// Create instance if needed (note prepare request may have been lost in network)
 	prop := px.getInstance(args.Instance)
 	reply.Err = true
@@ -367,10 +368,10 @@ func (px *Paxos) Accept(args *AcceptArgs, reply *AcceptReply) error {
 		newDone[dk] = dv
 	}
 
-	if enableLeader > 0 && prop.Decided && args.Value != prop.Value {
-		DPrintf("\n%v (L%v): Received accept for dead", px.me, px.leader)
-		reply.PID = args.PID
-	} else if args.PID >= prop.Prepare && (args.Server == px.leader[args.Instance] || enableLeader == 0) {
+  if prop.Decided && !args.Decided && args.Server == px.leader[args.Instance] {
+    DPrintf("\nDIDNT HEAR")
+    px.leader[args.Instance] = -1
+  } else if args.PID >= prop.Prepare && (args.Server == px.leader[args.Instance] || enableLeader == 0) {
 		px.instances[args.Instance] = Proposal{args.PID, args.PID, args.Value, prop.Decided}
 		px.dbWriteInstance(args.Instance, px.instances[args.Instance])
 		reply.Err = false
@@ -391,7 +392,7 @@ func (px *Paxos) Accept(args *AcceptArgs, reply *AcceptReply) error {
 // Respond to a Decide request
 func (px *Paxos) Decide(args *DecideArgs, reply *DecideReply) error {
 	px.mu.Lock()
[…]

### 2014-05-09 src/paxos/paxos.go
Fix one last one

@@ -38,8 +38,8 @@ import "bytes"
 const startport = 2100
 const printRPCerrors = false
 
-const persistent = true
-const recovery = true
+const persistent = false
+const recovery = false
 
 const Debug = 0
 const DebugPersist = 0
@@ -472,8 +472,6 @@ func (px *Paxos) Propose(args *ProposeArgs, reply *ProposeReply) error {
 		hValue := v
 		ok := 0
 
-    hValuePrime := make(map[interface{}]int)
-    
     if enableLeader == 2 {
       px.leader[seq] = -1
     }
@@ -501,36 +499,16 @@ func (px *Paxos) Propose(args *ProposeArgs, reply *ProposeReply) error {
 
           // Record highest prepare number / value among responses
           if reply.PID > hPID {
-            hValuePrime = make(map[interface{}]int)
             hPID = reply.PID
             if !hDecided {
               hValue = reply.Value
             }
           } 
-          if reply.PID == hPID && !hDecided && enableLeader > 0 {
-            if _,ok := hValuePrime[reply.Value]; ok {
-              hValuePrime[reply.Value] += 1
-            } else {
-              hValuePrime[reply.Value] = 1
-            }
-          }
         }
       }
     }
     DPrintf("\n%v (L%v): Has %v for sequence %v", px.me, px.leader[seq], v, seq)
     
-    if enableLeader > 0 {
-    hValueCount := -1
-    for k,v := range hValuePrime {
-      if v > hValueCount && k != nil {
-        if !hDecided {
-          hValue = k
-        }
-        hValueCount = v
-      }
-    }
-    }
-
 		// If prepare was rejected, start over with new proposal value
 		if ok <= total/2 {
 			px.mu.Lock()

### 2014-05-09 src/shardmaster/server.go

@@ -26,8 +26,8 @@ var logfile *os.File
 
 // Note: if persistent and recovery are not enabled,
 // some of the persistence tests fail by panic (divide-by-zero in balance)
-const persistent = true
-const recovery = true
+const persistent = false
+const recovery = false
 
 func DPrintf(format string, a ...interface{}) (n int, err error) {
 	if Debug > 0 {

### 2014-05-09 src/paxos/paxos.go
More work on Paxos

@@ -68,6 +68,7 @@ type Proposal struct {
 	Accept  int
 	Value   interface{}
 	Decided bool
+  Accepted bool
 }
 
 type Paxos struct {
@@ -120,6 +121,7 @@ type PrepareArgs struct {
 	Server   int
 	Instance int
 	PID      int
+  Decided  bool
 	Done     map[int]int
 	Leader   int
 }
@@ -304,7 +306,7 @@ func (px *Paxos) doneCollector() {
 			px.mu.Lock()
 			for k := range px.instances {
 				if k < min {
-					px.instances[k] = Proposal{-1, -1, nil, false}
+					px.instances[k] = Proposal{-1, -1, nil, false, false}
 					delete(px.instances, k)
 					px.dbDeleteInstance(k)
 				}
@@ -322,7 +324,7 @@ func (px *Paxos) Prepare(args *PrepareArgs, reply *PrepareReply) error {
 		px.recordDone(dk, dv)
 	}
 	px.mu.Lock()
-	DPrintf("\n%v (L%v): Received prepare for sequence %v", px.me, px.leader, args.Instance)
+	DPrintf("\n%v (L%v): Received prepare for sequence %v", px.me, px.leader[args.Instance], args.Instance)
 	prop := px.getInstance(args.Instance)
 	reply.Err = true
 
@@ -333,13 +335,16 @@ func (px *Paxos) Prepare(args *PrepareArgs, reply *PrepareReply) error {
 
 	// Check if proposal number is high enough
 	if args.PID > prop.Prepare {
-		px.instances[args.Instance] = Proposal{args.PID, prop.Accept, prop.Value, prop.Decided}
+		px.instances[args.Instance] = Proposal{args.PID, prop.Accept, prop.Value, prop.Decided, prop.Accepted}
 		px.dbWriteInstance(args.Instance, px.instances[args.Instance])
 		reply.Err = false
 		reply.PID = prop.Accept
 		reply.Value = prop.Value
 		reply.Leader = args.Server
-		reply.Decided = prop.Decided
+		reply.Decided = prop.Accepted
+    if args.Decided {
+      reply.Decided = prop.Decided
+    }
 		px.leader[args.Instance] = args.Server
 	}
 	px.mu.Unlock()
@@ -372,7 +377,7 @@ func (px *Paxos) Accept(args *AcceptArgs, reply *AcceptReply) error {
     DPrintf("\nDIDNT HEAR")
[…]

### 2014-05-09 src/paxos/test_test.go

@@ -11,7 +11,7 @@ import "sync"
 
 const onlyBenchmarks = false
 const runOldTests = true
-const runNewTests = true
+const runNewTests = false
 
 // Make a port using the given tag and host number
 func makePort(tag string, host int) string {
@@ -832,7 +832,7 @@ func TestFileForgetMem(test *testing.T) {
 	if onlyBenchmarks || !runOldTests {
 		return
 	}
-  return
+
 	runtime.GOMAXPROCS(4)
 
 	fmt.Printf("\nTest: Paxos frees forgotten instance memory ...")

### 2014-05-09 src/paxos/paxos.go
Behavioral flags for memory work better
Flags such as writeToMemory, dbCacheSize, etc. should be working now.
Remember to set these variables in each package when doing tests.

The exception is memoryLimit, which isn't really implemented yet.

@@ -38,13 +38,23 @@ import "bytes"
 const startport = 2100
 const printRPCerrors = false
 
-const persistent = true
-const recovery = true
-
 const Debug = 0
 const DebugPersist = 0
 
 const enableLeader = 0
+
+const persistent = true
+const recovery = true
+const writeToMemory = true     // Whether responses/store should be written to memory (as well as disk / disk cache)
+const dbUseCompression = true  // Whether database should compress entries
+const dbUseCache = true        // Whether database should use a built-in cache
+const dbCacheSize = 1000000000 // Size of database cache (ignored if dbUseCache is false)
+const memoryLimit = 2000000000 // Memory limit in bytes
+
+// Will use these to check that dbCacheSize doesn't overflow an int
+// (int size is either 32 or 64 bits depending on implementation)
+const MaxUint = ^uint(0)
+const MaxInt = int(^uint(0) >> 1)
 
 func DPrintf(format string, a ...interface{}) (n int, err error) {
 	if Debug > 0 {
@@ -103,6 +113,17 @@ type Paxos struct {
 	dbLock         sync.Mutex
 	dbMaxInstance  int
 	recovering     bool
+
+	// Behavioral Options
+	// NOTE: If shardkv/shardmaster is set to send settings to paxos, the below
+	// options will be overwritten
+	persistent       bool
+	recovery         bool
+	dbUseCompression bool
+	dbUseCache       bool
+	dbCacheSize      int
+	writeToMemory    bool
+	memoryLimit      int64
 }
 
 type RecoverArgs struct {
@@ -274,20 +295,32 @@ func (px *Paxos) callAcceptor(index int, name string, args interface{}, reply in
 	}
 }
 
+// Writes the given proposal to memory and/or disk
+func (px *Paxos) putInstance(seq int, proposal Proposal) {
+	if px.writeToMemory {
+		px.instances[seq] = proposal
+	}
+	px.dbWriteInstance(seq, proposal)
+}
+
 // Get the instance of the given sequence number
 // If haven't heard about it, creats a new instance for that sequence
[…]

### 2014-05-09 src/paxos/test_test.go

@@ -290,9 +290,9 @@ func TestFilePersistenceBasic(test *testing.T) {
 	fmt.Printf("\nTest Persistence, single failure, save disk ...")
 	// Put all servers in the same partition
 	partitionServers(test, tag, numServers, []int{0, 1, 2, 3, 4}, []int{}, []int{})
-	// Get agreement on an instance (only wait for majority)
+	// Get agreement on an instance
 	paxosServers[0].Start(0, 0)
-	waitForDecisionMajority(test, paxosServers, 0)
+	waitForDecision(test, paxosServers, 0, numServers)
 	// Kill one server, make sure it stops
 	paxosServers[0].KillSaveDisk()
 	time.Sleep(1 * time.Second)
@@ -893,7 +893,7 @@ func TestFileForgetMem(test *testing.T) {
 	runtime.ReadMemStats(&m2)
 
 	if m2.Alloc > (m1.Alloc / 2) {
-		test.Fatalf("memory use did not shrink enough")
+		test.Fatalf("memory use did not shrink enough (m2: %v, m1: %v", m2.Alloc, m1.Alloc)
 	}
 
 	fmt.Printf("\n\tPassed")

### 2014-05-09 src/shardkv/server.go

@@ -1,6 +1,5 @@
 package shardkv
 
-// ABOUT TO CHANGE RESPONSE STUFF
 import "net"
 import "fmt"
 import "net/rpc"
@@ -21,6 +20,9 @@ import "github.com/jmhodges/levigo"
 import "bytes"
 import "strings"
 
+import "runtime"
+import "os/exec"
+
 const Debug = 0
 const DebugPersist = 0
 const printRPCerrors = false
@@ -28,13 +30,23 @@ const Log = 0
 
 var logfile *os.File
 
+// DATABASE / MEMORY CONFIGURATION
+// Remember to set the corresponding variables in shardmaster and paxos
+// for accurate testing!
 const persistent = true
 const recovery = true
-const writeToMemory = true    // Whether responses/store should be written to memory (as well as disk / disk cache)
-const dbUseCache = true       // Whether database should use a built-in cache
-const dbUseCompression = true // Whether database should compress entries
+const writeToMemory = true     // Whether responses/store should be written to memory (as well as disk / disk cache)
+const dbUseCompression = true  // Whether database should compress entries
+const dbUseCache = true        // Whether database should use a built-in cache
+const dbCacheSize = 1000000000 // Size of database cache (ignored if dbUseCache is false)
+const memoryLimit = 2000000000 // Memory limit in bytes
+const copySettingsToPaxos = true
+const copySettingsToShardmaster = true
 
-//const memoryLimit = 1000000000 // Memory limit in bytes (not used yet)
+// Will use these to check that dbCacheSize doesn't overflow an int
+// (int size is either 32 or 64 bits depending on implementation)
+const MaxUint = ^uint(0)
+const MaxInt = int(^uint(0) >> 1)
 
 func DPrintf(format string, a ...interface{}) (n int, err error) {
 	if Debug > 0 {
@@ -1172,10 +1184,15 @@ func (kv *ShardKV) dbInit() {
 
 	DPrintfPersist("\n%v-%v: Initializing database", kv.gid, kv.me)
 
-	// Open database (create it if it doesn't exist)
+	// Set up database options
 	kv.dbOpts = levigo.NewOptions()
 	if dbUseCache {
[…]

### 2014-05-09 src/shardkv/test_test.go

@@ -166,6 +166,11 @@ func TestFileBasic(t *testing.T) {
 	}
 
 	rebootDone = 1
+	fmt.Printf("\n\tMemory usage          : %v", getMemoryUsage())
+	fmt.Printf("\n\tPaxos Disk usage      : %v", getPaxosDiskUsage())
+	fmt.Printf("\n\tShardmaster Disk usage: %v", getShardMasterDiskUsage())
+	fmt.Printf("\n\tShardKV Disk usage    : %v", getShardKVDiskUsage())
+	fmt.Printf("\n\tTotal Disk usage      : %v", getDiskUsage())
 	fmt.Printf("\n\tPassed\n")
 	time.Sleep(2 * time.Second)
 }

### 2014-05-09 src/shardmaster/server.go

@@ -24,10 +24,18 @@ const Log = 0
 
 var logfile *os.File
 
-// Note: if persistent and recovery are not enabled,
-// some of the persistence tests fail by panic (divide-by-zero in balance)
 const persistent = true
 const recovery = true
+const writeToMemory = true     // Whether responses/store should be written to memory (as well as disk / disk cache)
+const dbUseCompression = true  // Whether database should compress entries
+const dbUseCache = true        // Whether database should use a built-in cache
+const dbCacheSize = 1000000000 // Size of database cache (ignored if dbUseCache is false)
+const memoryLimit = 2000000000 // Memory limit in bytes
+
+// Will use these to check that dbCacheSize doesn't overflow an int
+// (int size is either 32 or 64 bits depending on implementation)
+const MaxUint = ^uint(0)
+const MaxInt = int(^uint(0) >> 1)
 
 func DPrintf(format string, a ...interface{}) (n int, err error) {
 	if Debug > 0 {
@@ -77,6 +85,17 @@ type ShardMaster struct {
 	dbLock         sync.Mutex
 	dbMaxConfig    int
 	recovering     bool
+
+	// Behavioral Options
+	// NOTE: If shardkv is set to send settings to shardmaster, the below
+	// options will be overwritten
+	persistent       bool
+	recovery         bool
+	dbUseCompression bool
+	dbUseCache       bool
+	dbCacheSize      int
+	writeToMemory    bool
+	memoryLimit      int64
 }
 
 type Op struct {
@@ -84,6 +103,14 @@ type Op struct {
 	GID     int64
 	Servers []string
 	Shard   int
+}
+
+// Store a config to memory and/or disk
+func (sm *ShardMaster) putConfig(configNum int, newConfig Config) {
+	sm.dbWriteConfig(configNum, newConfig)
+	if sm.writeToMemory {
+		sm.configs[configNum] = &newConfig
+	}
 }
 
 // Get the desired configuration from memory or disk
@@ -97,7 +124,9 @@ func (sm *ShardMaster) getConfig(configNum int) Config {
 	// Read from memory if possible, otherwise from disk
[…]

### 2014-05-09 src/test/test_test.go
committing to push multipaxos=true to other AWS instances

@@ -12,6 +12,10 @@ import "sync"
 import "math/rand"
 import "log"
 
+const numGroups = 1
+const numReplicas = 10
+const numMasters = 1
+
 // Use for checking PutHash                                                                 
 func NextValue(hprev string, val string) string {
         h := hash(hprev + val)
@@ -23,7 +27,6 @@ func setup(tag string, unreliable bool, numGroups int, numReplicas int) ([]strin
 	runtime.GOMAXPROCS(4)
 
 	//number of masters per group
-	const numMasters = 1
 	kvPorts, smPorts, gids := GetShardkvs(numReplicas, numMasters, numGroups)
 	fmt.Printf("kvports = %v, smports = %v\n", kvPorts, smPorts)
 	//clean := func() { shardkvCleanup(kvServers); smCleanup(smServers) }
@@ -31,8 +34,6 @@ func setup(tag string, unreliable bool, numGroups int, numReplicas int) ([]strin
 }
 
 func TestBasic(t *testing.T) {
-	numGroups := 3
-	numReplicas := 3
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 
@@ -95,8 +96,6 @@ func TestBasic(t *testing.T) {
 }
 
 func TestMove(t *testing.T) {
-	numGroups := 3
-	numReplicas := 3
 	smPorts, gids, kvPorts := setup("move", false, numGroups, numReplicas)
 
 	fmt.Printf("\nTest: Shards really move...")
@@ -163,8 +162,6 @@ func TestMove(t *testing.T) {
 
 
 func TestLimp(t *testing.T) {
-	numGroups := 3
-	numReplicas := 3
 	smPorts, gids, kvPorts := setup("limp", false, numGroups, numReplicas)
 
 	fmt.Printf("\nTest: Reconfiguration with some dead replicas...")
@@ -236,9 +233,8 @@ func TestLimp(t *testing.T) {
 }
 
 func doConcurrent(t *testing.T, unreliable bool) {
-	numGroups := 3
-	numReplicas := 3
-	smPorts, gids, kvPorts := setup("conc"+strconv.FormatBool(unreliable), unreliable, numGroups, numReplicas)
+	smPorts, gids, kvPorts := setup("conc"+strconv.FormatBool(unreliable), 
[…]

### 2014-05-09 associate-eip-large.sh
add large cluster

@@ -1,0 +1,1 @@
+ec2-associate-address -a eipalloc-b4f8eed6 -i $1 --region us-west-1

### 2014-05-09 connect-large.sh

@@ -1,0 +1,2 @@
+#ssh-keygen -f "/home/pranjal/.ssh/known_hosts" -R 54.86.104.17
+ssh -A -i mexos-keypair-1.pem ubuntu@54.183.9.25	

### 2014-05-09 launch-aws-headnode-large.sh

@@ -1,0 +1,1 @@
+ec2-run-instances ami-ee4f77ab --private-ip-address 10.0.0.100 -t m3.large -g sg-3a30dd5f --subnet subnet-4a21cc2f -k mexos-keypair-1 --region us-west-1 --associate-public-ip-address true

### 2014-05-09 launch-nodes-large.sh

@@ -1,0 +1,4 @@
+for i in {101..112}
+do
+    ec2-run-instances ami-ee4f77ab --private-ip-address 10.0.0.$i -t m3.large -g sg-3a30dd5f --subnet subnet-4a21cc2f -k mexos-keypair-1 --associate-public-ip-address true --region us-west-1
+done

### 2014-05-09 bringdown-large.sh
bringup and bringdown scripts

@@ -1,0 +1,1 @@
+ec2-stop-instances i-7f8bde23 i-728bde2e i-998adfc5 i-9a8adfc6 i-698bde35 i-2888dd74 i-2a88dd76	i-7d8bde21 i-3a8adf66 i-13b5e04f i-1eb5e042 i-9b8adfc7 i-338adf6f i-708bde2c

### 2014-05-09 bringup-large.sh

@@ -1,0 +1,1 @@
+ec2-start-instances i-7f8bde23 i-728bde2e i-998adfc5 i-9a8adfc6 i-698bde35 i-2888dd74 i-2a88dd76 i-7d8bde21 i-3a8adf66 i-13b5e04f i-1eb5e042 i-9b8adfc7 i-338adf6f i-708bde2c

### 2014-05-10 src/main/start.go
modified startup script

@@ -16,7 +16,7 @@ const network = true
 func main() {
 	var npaxos = flag.Int("npaxos", 3, "number of paxos instances")
 	var ngroups = flag.Int("ngroups", 3, "number of shard groups")
-	var nmasters = flag.Int("nmasters", 1, "number of shardmasters per shard group")
+	var nmasters = flag.Int("nmasters", 3, "number of shardmasters per shard group")
 	var nreplicas = flag.Int("nreplicas", 3, "number of kvshard replicas per group")
 	var clean =  flag.Bool("clean", false, "clean the db")
 	

### 2014-05-10 src/test/common.go

@@ -38,21 +38,21 @@ func GetPaxos(npaxos int) []string {
 }
 
 func GetShardmasters(nmasters int, ngroups int) ([]string, []int64) {
-	if nmasters*ngroups > 12 {
+	if nmasters > 12 {
 		fmt.Printf("Invalid configuration parameters!\n" +
-			"nmasters*ngroups >12 (= %d)\n", nmasters*ngroups)
+			"nmasters >12 (= %d)\n", nmasters)
 		os.Exit(1)
 	}
 	
 	gids := make([]int64, ngroups)    // each group ID                                 
-	var smhosts []string = make([]string, nmasters*ngroups)
+	var smhosts []string = make([]string, nmasters)
 	
 	//make GIDS
 	for gid := 0; gid < ngroups; gid++ {
 		gids[gid] = int64(gid + 100)
 	}
 	//make shardmasters
-	for i := 1; i <= ngroups*nmasters; i++ {
+	for i := 1; i <= nmasters; i++ {
 		smhosts[i-1] = "10.0.0." + strconv.Itoa(100+i) + ":" + strconv.Itoa(smport)
 	}
 	return smhosts, gids
@@ -61,7 +61,7 @@ func GetShardmasters(nmasters int, ngroups int) ([]string, []int64) {
 func GetShardkvs(nreplicas int, nmasters int, ngroups int) ([][]string, []string,
 	[]int64){
 	
-	if (nreplicas+nmasters)*ngroups > 12  {
+	if (nreplicas*ngroups +nmasters) > 12  {
 		fmt.Printf("Invalid configuration parameters!\n"+
 			"(nmasters+nreplicas)*ngroups >12 (= %d)\n", 
 			(nmasters+nreplicas)*ngroups)
@@ -71,7 +71,7 @@ func GetShardkvs(nreplicas int, nmasters int, ngroups int) ([][]string, []string
 
 	smhosts, gids := GetShardmasters(nmasters, ngroups)
 
-	offset := nmasters*ngroups+1
+	offset := nmasters+1
 	for grp := 0; grp < ngroups; grp++ {
 		kvhosts[grp] = make([]string, nreplicas)
 		for j := 0; j < nreplicas; j++ {

### 2014-05-10 src/test/test_test.go

@@ -335,7 +335,7 @@ func BenchmarkClientLatencyManyShard(benchmark *testing.B) {
 
 
 func TestManyClientOneShard(t *testing.T) {
-	nclients := 10
+	nclients := 15
 	nseconds := 10
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()

### 2014-05-10 src/paxos/paxos.go
turning off leaders

@@ -41,7 +41,7 @@ const printRPCerrors = false
 const Debug = 0
 const DebugPersist = 0
 
-const enableLeader = 1
+const enableLeader = 0
 
 const persistent = true
 const recovery = true

### 2014-05-10 src/test/test_test.go

@@ -12,9 +12,9 @@ import "sync"
 import "math/rand"
 import "log"
 
-const numGroups = 1
-const numReplicas = 10
-const numMasters = 1
+const numGroups = 4
+const numReplicas = 2
+const numMasters = 2
 
 // Use for checking PutHash                                                                 
 func NextValue(hprev string, val string) string {
@@ -301,7 +301,7 @@ func BenchmarkClientLatencyOneShard(benchmark *testing.B) {
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 
-	fmt.Printf("\nBenchmark: client latency, one shard...")
+	fmt.Printf("\nBenchmark: client latency, one shard...\n")
 
 	smClerk := shardmaster.MakeClerk(smPorts, true)
 	smClerk.Join(gids[0], kvPorts[0])
@@ -318,7 +318,7 @@ func BenchmarkClientLatencyManyShard(benchmark *testing.B) {
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 
-	fmt.Printf("\nBenchmark: client latency, many shards...")
+	fmt.Printf("\nBenchmark: client latency, many shards...\n")
 
 	smClerk := shardmaster.MakeClerk(smPorts, true)
 	for i := 0; i < len(gids); i++ {
@@ -335,7 +335,7 @@ func BenchmarkClientLatencyManyShard(benchmark *testing.B) {
 
 
 func TestManyClientOneShard(t *testing.T) {
-	nclients := 15
+	nclients := 35
 	nseconds := 10
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()

### 2014-05-10 src/shardkv/server.go
turned on logging

@@ -23,10 +23,10 @@ import "strings"
 import "runtime"
 import "os/exec"
 
-const Debug = 0
-const DebugPersist = 0
+const Debug = 1
+const DebugPersist = 1
 const printRPCerrors = false
-const Log = 0
+const Log = 1
 
 var logfile *os.File
 

### 2014-05-10 src/test/test_test.go

@@ -12,7 +12,7 @@ import "sync"
 import "math/rand"
 import "log"
 
-const numGroups = 4
+const numGroups = 3
 const numReplicas = 2
 const numMasters = 2
 

### 2014-05-10 src/paxos/paxos.go
Added multi-RPC shard transfers

@@ -45,11 +45,10 @@ const enableLeader = 1
 
 const persistent = true
 const recovery = true
-const writeToMemory = true     // Whether responses/store should be written to memory (as well as disk / disk cache)
-const dbUseCompression = true  // Whether database should compress entries
-const dbUseCache = true        // Whether database should use a built-in cache
-const dbCacheSize = 1000000000 // Size of database cache (ignored if dbUseCache is false)
-const memoryLimit = 2000000000 // Memory limit in bytes
+const writeToMemory = false   // Whether responses/store should be written to memory (as well as disk / disk cache)
+const dbUseCompression = true // Whether database should compress entries
+const dbUseCache = true       // Whether database should use a built-in cache
+const dbCacheSize = 100       // Size of database cache in MB (ignored if dbUseCache is false)
 
 // Will use these to check that dbCacheSize doesn't overflow an int
 // (int size is either 32 or 64 bits depending on implementation)
@@ -124,7 +123,6 @@ type Paxos struct {
 	dbUseCache       bool
 	dbCacheSize      int
 	writeToMemory    bool
-	memoryLimit      int64
 }
 
 type RecoverArgs struct {
@@ -1035,11 +1033,11 @@ func (px *Paxos) dbInit(tag string) {
 	// Open database (create it if it doesn't exist)
 	px.dbOpts = levigo.NewOptions()
 	if px.dbUseCache {
-		if px.dbCacheSize > MaxInt {
-			fmt.Printf("\nDesired cache size %v is too large... using %v instead\n", px.dbCacheSize, MaxInt)
+		if px.dbCacheSize*1000000 > MaxInt {
+			fmt.Printf("\nDesired cache size %v is too large... using %v instead\n", px.dbCacheSize*1000000, MaxInt)
 			px.dbOpts.SetCache(levigo.NewLRUCache(MaxInt))
 		} else {
-			px.dbOpts.SetCache(levigo.NewLRUCache(px.dbCacheSize))
+			px.dbOpts.SetCache(levigo.NewLRUCache(px.dbCacheSize * 1000000))
 		}
 	}
 	if px.dbUseCompression {
@@ -1231,7 +1229,6 @@ func Make(peers []string, me int, rpcs *rpc.Server, network bool, tag string) *P
 	px.dbUseCache = dbUseCache
 	px.dbCacheSize = dbCacheSize
 	px.writeToMemory = writeToMemory
-	px.memoryLimit = memoryLimit
 
 	// Network stuff
 	px.peers = peers

### 2014-05-10 src/shardkv/server.go

@@ -35,13 +35,12 @@ var logfile *os.File
 // for accurate testing!
 const persistent = true
 const recovery = true
-const writeToMemory = true     // Whether responses/store should be written to memory (as well as disk / disk cache)
-const dbUseCompression = true  // Whether database should compress entries
-const dbUseCache = true        // Whether database should use a built-in cache
-const dbCacheSize = 1000000000 // Size of database cache (ignored if dbUseCache is false)
-const memoryLimit = 2000000000 // Memory limit in bytes
-const copySettingsToPaxos = true
-const copySettingsToShardmaster = true
+const writeToMemory = false                    // Whether responses/store should be written to memory (as well as disk / disk cache)
+const dbUseCompression = true                  // Whether database should compress entries
+const dbUseCache = true                        // Whether database should use a built-in cache
+const dbCacheSize = 250                        // Size of database cache in MB (ignored if dbUseCache is false)
+const memoryLimit = 1000                       // Memory limit in MB
+const memoryThreshold = memoryLimit * 80 / 100 // When to stop filling memory (when to abort a Fetch RPC and use multiple messages)
 
 // Will use these to check that dbCacheSize doesn't overflow an int
 // (int size is either 32 or 64 bits depending on implementation)
@@ -433,21 +432,7 @@ func (kv *ShardKV) fetchHandler(args *FetchArgs, reply *FetchReply) error {
 		return nil
 	}
 
-	shardStore := make(map[string]string)
-	keysInMemory := make(map[string]bool)
-	// Copy key/value pairs for desired shard from memory
-	for k, v := range kv.store {
-		if key2shard(k) == args.Shard {
-			shardStore[k] = v
-			keysInMemory[k] = true
-		}
-	}
-	// Copy key/value pairs for desired shard from disk if not in memory
-	for k, v := range kv.dbGetShard(args.Shard, keysInMemory) {
-		DPrintfPersist("\n\t%v-%v: got shard %v data (%v, %v)", kv.gid, kv.me, args.Shard, k, v)
[…]

### 2014-05-10 src/shardkv/test_test.go

@@ -560,3 +560,78 @@ func TestFilePersistenceOriginals(t *testing.T) {
 	done = true
 	fmt.Printf("\n\n\tPassed!!\n")
 }
+
+func TestFileMemory(t *testing.T) {
+	smPorts, gids, kvPorts, kvServers, clean := setup("persistBasic", false, 3, 3)
+	defer clean()
+	smClerk := shardmaster.MakeClerk(smPorts, false)
+	kvClerk := MakeClerk(smPorts, false)
+
+	targetSize := 100 // Target size in MB
+	valueSize := 2000 // Size of each value in KB
+
+	// Periodically check that too much memory isn't being used
+	done := false
+	go func() {
+		for !done {
+			runtime.GC()
+			time.Sleep(50 * time.Millisecond) // Not sure if this is needed
+			usage := getMemoryUsage() / 1000
+			if usage > memoryLimit {
+				t.Fatalf("Too much memory is being used! Using %v MB, limit is %v MB", usage, memoryLimit)
+			}
+			time.Sleep(2000 * time.Millisecond)
+		}
+	}()
+
+	// Make a big value
+	big := make([]byte, valueSize*1000)
+	for j := 0; j < len(big); j++ {
+		big[j] = byte('a' + rand.Int()%26)
+	}
+
+	// Join all groups
+	for i := 0; i < len(gids); i++ {
+		smClerk.Join(gids[i], kvPorts[i])
+	}
+
+	time.Sleep(3 * time.Second)
+
+	fmt.Printf("\nPutting %v MB worth of stuff ...\t", targetSize)
+	// Many keys with large values
+	for i := targetSize * 1000 / valueSize; i >= 0; i-- {
+		if (i == 99) && (targetSize*1000/valueSize > 99) {
+			fmt.Printf(" ")
+		}
+		fmt.Printf("%v", i)
+
+		kvClerk.Put(strconv.FormatInt(nrand(), 10), string(big))
+
+		fmt.Printf("\b")
+		if i > 9 {
+			fmt.Printf("\b")
+		}
+		if i > 99 {
+			fmt.Printf("\b")
+		}
+	}
+	fmt.Printf("\nDone!")
+
+	// Kill a server (destroy disk)
+	kvServers[0][0].Kill()
+
+	time.Sleep(3 * time.Second)
+
+	// Reboot server to cause recovery
[…]

### 2014-05-10 src/shardmaster/server.go

@@ -26,11 +26,10 @@ var logfile *os.File
 
 const persistent = true
 const recovery = true
-const writeToMemory = true     // Whether responses/store should be written to memory (as well as disk / disk cache)
-const dbUseCompression = true  // Whether database should compress entries
-const dbUseCache = true        // Whether database should use a built-in cache
-const dbCacheSize = 1000000000 // Size of database cache (ignored if dbUseCache is false)
-const memoryLimit = 2000000000 // Memory limit in bytes
+const writeToMemory = false   // Whether responses/store should be written to memory (as well as disk / disk cache)
+const dbUseCompression = true // Whether database should compress entries
+const dbUseCache = true       // Whether database should use a built-in cache
+const dbCacheSize = 100       // Size of database cache in MB (ignored if dbUseCache is false)
 
 // Will use these to check that dbCacheSize doesn't overflow an int
 // (int size is either 32 or 64 bits depending on implementation)
@@ -95,7 +94,6 @@ type ShardMaster struct {
 	dbUseCache       bool
 	dbCacheSize      int
 	writeToMemory    bool
-	memoryLimit      int64
 }
 
 type Op struct {
@@ -689,11 +687,11 @@ func (sm *ShardMaster) dbInit() {
 	// Open database (create it if it doesn't exist)
 	sm.dbOpts = levigo.NewOptions()
 	if sm.dbUseCache {
-		if sm.dbCacheSize > MaxInt {
-			fmt.Printf("\nDesired cache size %v is too large... using %v instead\n", sm.dbCacheSize, MaxInt)
+		if sm.dbCacheSize*1000000 > MaxInt {
+			fmt.Printf("\nDesired cache size %v is too large... using %v instead\n", sm.dbCacheSize*1000000, MaxInt)
 			sm.dbOpts.SetCache(levigo.NewLRUCache(MaxInt))
 		} else {
-			sm.dbOpts.SetCache(levigo.NewLRUCache(sm.dbCacheSize))
+			sm.dbOpts.SetCache(levigo.NewLRUCache(sm.dbCacheSize * 1000000))
 		}
 	}
 	if sm.dbUseCompression {
@@ -885,7 +883,6 @@ func StartServer(servers []string, me int, network bool) *ShardMaster {
 	sm.dbUseCache = dbUseCache
 	sm.dbCacheSize = dbCacheSize
 	sm.writeToMemory = writeToMemory
-	sm.memoryLimit = memoryLimit
 
 	// Network stuff
 	sm.me = me

### 2014-05-10 writeup/paper.tex
paper edits

@@
 \usepackage{mathtools,amssymb,setspace, pbox}
+\usepackage[table,xcdraw]{xcolor}
 \usepackage[center, labelfont=bf]{caption}
@@
 \section{Introduction}
-This document is the writeup of our 6.824 final project.
-
-The goals of the project are:
+This document is the writeup of our 6.824 final project. We did the
+default project, and built a persistent, fault-tolerant
+high-performance key/value store. For persistence, we used the go
+wrapper for the high-performance C++ leveldb key/value libray, and
+extended our code to recover from data loss upon server restart. In
+order to improve the performance of our system, we implemented two
+paxos optimizations. We used a multi-paxos like design to avoid the
+dueling leaders problem, and also reduce the number of RPCs for paxos
+agreeement by having the leader send out prepare messages ahead of
+time. To measure performance, we extended our existing tests, and
+added new ones. We also modified the RPC system to work on a real
+network and deployed our project on an AWS cluster to get realistic
+throughput and latency measures, as well as information about the
+speed of disk recovery.
 
 \section{Design} \label{sec:design}
+Our starting point was lab 4. Before begining, we ran battery tests to
+figure out which combination of our prior labs created the fastest and
+most reliable starting point.
 
@@
 
+\subsection{Testing}
+We added about 25 new tests, and extended existing ones to work with
+persistent storage and run on TCP sockets instead of unix file
+sockets. We use a few flags and configurations to run different modes
+(e.g. with and without persistence, RPC using unix file sockets or
+localhost TCP).
+
+Below we describe the different tests. All tests are located at
+'mexos/src/<name>/test_test.go' unless otherwise noted.
+
+\emph{paxos:} Paxos specific tests
+
+\emph{shardmaster:} Shardmaster specific tests
+
+\emph{shardkv:} Shardkv specific tests
+
+\emph{test:} Tests and benchmarks modified to run on our AWS
+cluster. These tests are the most complicated, and require the user to
+ssh into our cluster leader and start the servers on other cluster
+machines using 'start-on-all.sh' to run
+'mexos/src/main/start.go'. These tests required a surprising amount of
+debugging and glue code. Additionally, to check for correct operation
+under server failure, we had to make a backdoor RPC that told the
+remote server to act deaf or dead.
+
+We present the results of our benchmarks in a later section.
+
 \section{Performance}
 
-We created an extensive system to try and accurately measure performance.
-In addition to writing many benchmarks, we also modified the original RPC system so that we could run the servers in an AWS cluster and observe the performance on a real network.
+We created an extensive system to try and accurately measure
+performance. In addition to writing many benchmarks, we also modified
+the original RPC system so that we could run the servers in an AWS
+cluster and observe the performance on a real network.
 
 \subsection{Throughput and Latency}
-Avg client latency w/ out paxos optimizations:
-one shard, 3 groups w/ 3 replicas = 15.4ms/op
-many shards, 3 groups w/ 3 replicas = 21.8ms/op
 
-Avg client throughput w/ out paxos optimizations:
-one shard, 3 groups w/ 3 replicas = 65 requests/second
-many shards, 3 groups w/ 3 replicas = 46 requets/second
+We compared performance between vanilla paxos and multipaxos under a
+variety of conditions. The tests were run on an actual network in an
+AWS cluster.
 
-Client requests per second:
-3 groups w/ 3 replicas:
-~179 requests per second (saturates at around 50 users)
+\textit{Client latency}: Average time it takes for one client to see a
+request processed.
+
+\textit{Throughput}: Number of requests, on average, the
+client will see processed in a second.
+
+\textit{System throughput}: Number of requests per second the entire
+system can handle. This is different than the client throughput
+because clients always wait to see the response of the current request
+before dispatching another.
+
+System throughput requires a saturation of multiple clients sending
+continuous requests over a ten second period. The system saturated
+from 10 to 50 clients, sepending on the number of
+groups/replicas. After the saturation point, adding more clients
+causes a decrease in throughput occurs because of network traffic
+saturation.
+
+We also ran some tests repeatedly on one key, and others on a random
+set of keys. This allows us to look at how the performance changes
+when requests all go to the same replica group and when they are
+evenly spread across all replica groups (assuming the staff-provided
+key2shard isn't biased).
+
+\begin{table}[h]
+\begin{tabular}{|l|l|l|l|}
+\hline
+\multicolumn{4}{|c|}{\cellcolor[HTML]{C0C0C0}{\color[HTML]{000000} \textbf{Vanilla Paxos}}}                              \\ \hline
+                              & \textbf{Avg. Client Latency} & \textbf{Avg. Client Thruput} & \textbf{System Thruput} \\ \hline
+\multicolumn{4}{|c|}{\textit{PutHash to one key}}                                                                     \\ \hline
+\textbf{1 group, 3 replicas}  & 21.6ms/op                    & 46 requests/s                & 85 requests/s           \\ \hline
+\textbf{1 group, 10 replicas} & 52.7ms/op                    & 19 requests/s                & 19 requests/s           \\ \hline
+\textbf{3 groups, 3 replicas} & 15.4ms/op                    & 65 requests/s                & 70 requests/s           \\ \hline
+\multicolumn{4}{|c|}{\textit{PutHash to many keys}}                                                                 \\ \hline
+\textbf{1 group, 3 replicas}  & 20.5ms/op                    & 49 requests/s                & 78 requests/s           \\ \hline
+\textbf{1 group, 10 replicas} & 74ms/op                      & 14 requests/s                & 19 requests/s           \\ \hline
+\textbf{3 groups, 3 replicas} & 21.8ms/op                    & 46 requests/s                & 179 requests/s          \\ \hline
+\end{tabular}
+\end{table}
+
+%----------------------------------------------------------------------
+
+\begin{table}[h]
+\begin{tabular}{|l|l|l|l|}
+\hline
+\multicolumn{4}{|c|}{\cellcolor[HTML]{C0C0C0}{\color[HTML]{000000} \textbf{Multipaxos}}}                              \\ \hline
+                              & \textbf{Avg. Client Latency} & \textbf{Avg. Client Thruput} & \textbf{System Thruput} \\ \hline
+\multicolumn{4}{|c|}{\textit{PutHash to one key}}                                                                     \\ \hline
+\textbf{1 group, 3 replicas}  & 14.2ms/op                    & 70.4 requests/s              & 88.8 requests/s         \\ \hline
+\textbf{1 group, 10 replicas} & 36.9ms/op                    & 27 requests/s                & 31 requests/s           \\ \hline
+\textbf{3 groups, 3 replicas} & 14.1ms/op                    & 70.9 requests/s              & 86 requests/s           \\ \hline
+\multicolumn{4}{|c|}{\textit{PutHash to many keys}}                                                                   \\ \hline
+\textbf{1 group, 3 replicas}  & 13.3ms/op                    & 75.2 requests/s              & 87.4 requests/s         \\ \hline
+\textbf{1 group, 10 replicas} & 37.7ms/op                    & 26.5 requests/s              & 31 requests/s           \\ \hline
+\textbf{3 groups, 3 replicas} & 15.8ms/op                    & 63.3 requests/s              & 200 requests/s          \\ \hline
+\end{tabular}
+\end{table}
+
+\subsubsection{Paxos RPC Counts}
+
+\subsubsection{Disk Recovery}
+For disk recovery, the bottleneck was...
 
 \subsection{Bottlenecks}
-Although we made significant improvements performance-wise, we identified a number of bottlenecks.
+Although we made significant improvements performance-wise, we
+identified a number of bottlenecks.
 
@@
 
-\subsubsection{Paxos}
+Another bottleneck is network capacity and/or number of messages. As
+shown in the latency and thoughput section, the increasing the number
+of replicas has an adverse effect on performance since each operation
+requires communication with a majority of replicas. Using multipaxos
+improves the performance since it decreases the number of RPCs needed
+to commit an operation.
 
-\subsubsection{Disk Recovery}
+Increasing the number of replica groups increases the performance of
+the system for normal operation, especially under heavy load with
+evenly distributed key requests. This is especially noticable in the
+system throughput metric.  The number of groups roughly corresponds to
+how parallelizable the operations are.
+
+Another benefit of the paxos optimizations is that with
+fewer messages, less disk space is used since less paxos state data
+needs to be saved. We also suspect that using UDP would also improve
+throughput, as servers that do not recieve a response already
+re-transmit their request, making the TCP ack system redundant.
 
 \section{Conclusion}
-We did a project.
 

### 2014-05-10 src/main/start.go
fixed start  bug

@@ -73,6 +73,7 @@ func main() {
 		//group 101: 10.0.0.107, 10.0.0.108, 10.0.0.109
 		//group 102: 10.0.0.110, 10.0.0.111, 10.0.0.112
 		metapeers, masters, groups := test.GetShardkvs(*nreplicas, *nmasters, *ngroups)
+		
 		me := whoami(masters)
 		if me != -1 {
 			fmt.Println("Starting shardmaster instead.")
@@ -90,7 +91,7 @@ func main() {
 		for i, v := range metapeers {
 			peers = v
 			me = whoami(v)
-			gid = groups[i]
+			gid = i
 			if me != -1 {
 				break
 			}

### 2014-05-10 src/test/test_test.go

@@ -12,7 +12,7 @@ import "sync"
 import "math/rand"
 import "log"
 
-const numGroups = 3
+const numGroups = 4
 const numReplicas = 2
 const numMasters = 2
 

### 2014-05-10 src/paxos/paxos.go
enabled leader

@@ -41,7 +41,7 @@ const printRPCerrors = false
 const Debug = 0
 const DebugPersist = 0
 
-const enableLeader = 0
+const enableLeader = 1
 
 const persistent = true
 const recovery = true

### 2014-05-10 src/main/start.go
enabled leader

@@ -72,7 +72,7 @@ func main() {
 		//group 100: 10.0.0.104, 10.0.0.105, 10.0.0.106
 		//group 101: 10.0.0.107, 10.0.0.108, 10.0.0.109
 		//group 102: 10.0.0.110, 10.0.0.111, 10.0.0.112
-		metapeers, masters, groups := test.GetShardkvs(*nreplicas, *nmasters, *ngroups)
+		metapeers, masters, _ := test.GetShardkvs(*nreplicas, *nmasters, *ngroups)
 		
 		me := whoami(masters)
 		if me != -1 {
@@ -91,7 +91,7 @@ func main() {
 		for i, v := range metapeers {
 			peers = v
 			me = whoami(v)
-			gid = i
+			gid = int64(i)
 			if me != -1 {
 				break
 			}

### 2014-05-10 src/main/start.go
groupid bug

@@ -91,7 +91,7 @@ func main() {
 		for i, v := range metapeers {
 			peers = v
 			me = whoami(v)
-			gid = int64(i)
+			gid = int64(100+i)
 			if me != -1 {
 				break
 			}

### 2014-05-10 src/shardkv/server.go
blab blah blah

@@ -23,10 +23,10 @@ import "strings"
 import "runtime"
 import "os/exec"
 
-const Debug = 1
-const DebugPersist = 1
+const Debug = 0
+const DebugPersist = 0
 const printRPCerrors = false
-const Log = 1
+const Log = 0
 
 var logfile *os.File
 

### 2014-05-10 src/shardmaster/server.go

@@ -17,10 +17,10 @@ import "strconv"
 import "github.com/jmhodges/levigo"
 import "bytes"
 
-const Debug = 1
-const DebugPersist = 1
+const Debug = 0
+const DebugPersist = 0
 const printRPCerrors = false
-const Log = 1
+const Log = 0
 
 var logfile *os.File
 

### 2014-05-10 src/main/start.go
more numbers

@@ -73,6 +73,8 @@ func main() {
 		//group 101: 10.0.0.107, 10.0.0.108, 10.0.0.109
 		//group 102: 10.0.0.110, 10.0.0.111, 10.0.0.112
 		metapeers, masters, _ := test.GetShardkvs(*nreplicas, *nmasters, *ngroups)
+		fmt.Printf("peers: %v\n", metapeers)
+
 		
 		me := whoami(masters)
 		if me != -1 {

### 2014-05-10 src/test/common.go

@@ -77,7 +77,7 @@ func GetShardkvs(nreplicas int, nmasters int, ngroups int) ([][]string, []string
 		for j := 0; j < nreplicas; j++ {
 			//fmt.Println(100+(grp*ngroups)+j+offset)
 			kvhosts[grp][j] = "10.0.0." + 
-				strconv.Itoa(100+(grp*ngroups)+j+offset) +
+				strconv.Itoa(100+(grp*nreplicas)+j+offset) +
 				":" + strconv.Itoa(kvport)
 		}
 	}

### 2014-05-10 src/shardkv/client.go
still debugging

@@ -165,7 +165,8 @@ func (ck *Clerk) PutExt(key string, value string, dohash bool) string {
 		// ask master for a new configuration.
 		prior := ck.config.Num
 		ck.config = ck.sm.Query(-1)
-		DPrintf("Prior config %d new is %d", prior, ck.config.Num)
+		DPrintf("Prior config %d new is %d: %v", prior, ck.config.Num,
+			ck.config.Shards,)
 	}
 	return ""
 }

### 2014-05-10 src/shardkv/server.go

@@ -23,8 +23,8 @@ import "strings"
 import "runtime"
 import "os/exec"
 
-const Debug = 0
-const DebugPersist = 0
+const Debug = 1
+const DebugPersist = 1
 const printRPCerrors = false
 const Log = 0
 

### 2014-05-10 src/shardmaster/server.go

@@ -17,8 +17,8 @@ import "strconv"
 import "github.com/jmhodges/levigo"
 import "bytes"
 
-const Debug = 0
-const DebugPersist = 0
+const Debug = 1
+const DebugPersist = 1
 const printRPCerrors = false
 const Log = 0
 
@@ -291,11 +291,12 @@ func (sm *ShardMaster) processLog(maxSeq int) {
 
 // Accept a Join request
 func (sm *ShardMaster) Join(args *JoinArgs, reply *JoinReply) error {
+	DPrintf("%d) Join: %d -> %s\n", sm.me, args.GID, args.Servers)
 	for sm.recovering && !sm.dead {
 		time.Sleep(10 * time.Millisecond)
 	}
 	sm.mu.Lock()
-	DPrintf("%d) Join: %d -> %s\n", sm.me, args.GID, args.Servers)
+
 
 	newOp := Op{2, args.GID, args.Servers, 0}
 

### 2014-05-10 src/shardkv/server.go
still debugging...

@@ -26,7 +26,7 @@ import "os/exec"
 const Debug = 1
 const DebugPersist = 1
 const printRPCerrors = false
-const Log = 0
+const Log = 1
 
 var logfile *os.File
 

### 2014-05-10 src/shardmaster/server.go

@@ -20,7 +20,7 @@ import "bytes"
 const Debug = 1
 const DebugPersist = 1
 const printRPCerrors = false
-const Log = 0
+const Log = 1
 
 var logfile *os.File
 

### 2014-05-10 src/test/test_test.go
turned off leader

@@ -37,7 +37,7 @@ func TestBasic(t *testing.T) {
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 
-	fmt.Printf("\nTest: Basic Join/Leave...")
+	fmt.Printf("\nTest: Basic Join/Leave...\n")
 
 	smClerk := shardmaster.MakeClerk(smPorts, true)
 	smClerk.Join(gids[0], kvPorts[0])
@@ -51,7 +51,7 @@ func TestBasic(t *testing.T) {
 	if kvClerk.Get("a") != ov {
 		t.Fatalf("Get got wrong value")
 	}
-	log.Printf("got here6")
+	log.Printf("got here6\n")
 	keys := make([]string, 10)
 	vals := make([]string, len(keys))
 	for i := 0; i < len(keys); i++ {
@@ -59,10 +59,11 @@ func TestBasic(t *testing.T) {
 		vals[i] = strconv.Itoa(rand.Int())
 		kvClerk.Put(keys[i], vals[i])
 	}
-
+	log.Printf("got here7\n")
 	// are keys still there after joins?
 	for g := 1; g < len(gids); g++ {
 		smClerk.Join(gids[g], kvPorts[g])
+		fmt.Printf("gid 10%d joined\n",g)
 		time.Sleep(1 * time.Second)
 		for i := 0; i < len(keys); i++ {
 			v := kvClerk.Get(keys[i])
@@ -74,10 +75,11 @@ func TestBasic(t *testing.T) {
 			kvClerk.Put(keys[i], vals[i])
 		}
 	}
-
+	log.Printf("got here8\n")
 	// are keys still there after leaves?
 	for g := 0; g < len(gids)-1; g++ {
 		smClerk.Leave(gids[g])
+		fmt.Printf("gid 10%d left\n",g)
 		time.Sleep(1 * time.Second)
 		for i := 0; i < len(keys); i++ {
 			v := kvClerk.Get(keys[i])

### 2014-05-10 src/shardmaster/server.go
turn on dbugs

@@ -17,10 +17,10 @@ import "strconv"
 import "github.com/jmhodges/levigo"
 import "bytes"
 
-const Debug = 0
-const DebugPersist = 0
+const Debug = 1
+const DebugPersist = 1
 const printRPCerrors = false
-const Log = 0
+const Log = 1
 
 var logfile *os.File
 

### 2014-05-10 src/shardmaster/server.go
ADDED DEBUG AND CAPSLOCK

@@ -875,7 +875,7 @@ func StartServer(servers []string, me int, network bool) *ShardMaster {
 		}
 		enableLog()
 	}
-
+	DPrintf("Shardamaster %s server started with peers",servers[me],servers)
 	sm := new(ShardMaster)
 	// Read memory options
 	sm.persistent = persistent

### 2014-05-10 src/test/test_test.go
more debugging

@@ -78,17 +78,22 @@ func TestBasic(t *testing.T) {
 	log.Printf("got here8\n")
 	// are keys still there after leaves?
 	for g := 0; g < len(gids)-1; g++ {
+		fmt.Printf("gid 10%d wants to leave\n",g)
 		smClerk.Leave(gids[g])
 		fmt.Printf("gid 10%d left\n",g)
 		time.Sleep(1 * time.Second)
 		for i := 0; i < len(keys); i++ {
+			fmt.Printf("getting %v\n",keys[i])
 			v := kvClerk.Get(keys[i])
+			fmt.Printf("got %v\n",keys[i])
 			if v != vals[i] {
 				t.Fatalf("leaving; wrong value; g=%v k=%v wanted=%v got=%v",
 					g, keys[i], vals[i], v)
 			}
 			vals[i] = strconv.Itoa(rand.Int())
+			fmt.Printf("putting %v\n",keys[i])
 			kvClerk.Put(keys[i], vals[i])
+			fmt.Printf("put success %v\n",keys[i])
 		}
 	}
 

### 2014-05-10 src/shardkv/test_test.go
changed shardkv test to show failure

@@ -65,7 +65,7 @@ func shardkvCleanup(kvServers [][]*ShardKV) {
 func setup(tag string, unreliable bool, numGroups int, numReplicas int) ([]string, []int64, [][]string, [][]*ShardKV, func()) {
 	runtime.GOMAXPROCS(4)
 
-	const numMasters = 3
+	const numMasters = 2
 	var smServers []*shardmaster.ShardMaster = make([]*shardmaster.ShardMaster, numMasters)
 	var smPorts []string = make([]string, numMasters)
 	// Start shardmaster servers
@@ -101,12 +101,12 @@ func TestFileBasic(t *testing.T) {
 	if !runOldTests {
 		return
 	}
-	numGroups := 3
-	numReplicas := 3
+	numGroups := 4
+	numReplicas := 2
 	smPorts, gids, kvPorts, kvServers, clean := setup("basic", false, numGroups, numReplicas)
 	defer clean()
 
-	fmt.Printf("\nTest: Basic Join/Leave...")
+	fmt.Printf("\nTest: Basic Join/Leave...\n")
 
 	smClerk := shardmaster.MakeClerk(smPorts, false)
 	smClerk.Join(gids[0], kvPorts[0])
@@ -153,15 +153,20 @@ func TestFileBasic(t *testing.T) {
 	// are keys still there after leaves?
 	for g := 0; g < len(gids)-1; g++ {
 		smClerk.Leave(gids[g])
+		fmt.Printf("group 10%d left\n",g)
 		time.Sleep(1 * time.Second)
 		for i := 0; i < len(keys); i++ {
+			fmt.Printf("getting %v\n",keys[i])
 			v := kvClerk.Get(keys[i])
+			fmt.Printf("got %v",keys[i])
 			if v != vals[i] {
 				t.Fatalf("leaving; wrong value; g=%v k=%v wanted=%v got=%v",
 					g, keys[i], vals[i], v)
 			}
 			vals[i] = strconv.Itoa(rand.Int())
+			fmt.Printf("putting %v\n",keys[i])
 			kvClerk.Put(keys[i], vals[i])
+			fmt.Printf("put success %v\n",keys[i])
 		}
 	}
 

### 2014-05-10 src/shardkv/test_test.go
enabled leader

@@ -65,7 +65,7 @@ func shardkvCleanup(kvServers [][]*ShardKV) {
 func setup(tag string, unreliable bool, numGroups int, numReplicas int) ([]string, []int64, [][]string, [][]*ShardKV, func()) {
 	runtime.GOMAXPROCS(4)
 
-	const numMasters = 2
+	const numMasters = 3
 	var smServers []*shardmaster.ShardMaster = make([]*shardmaster.ShardMaster, numMasters)
 	var smPorts []string = make([]string, numMasters)
 	// Start shardmaster servers

### 2014-05-10 src/paxos/paxos.go
Running gofmt

@@ -1235,7 +1235,7 @@ func Make(peers []string, me int, rpcs *rpc.Server, network bool, tag string) *P
 	px.me = me
 	px.network = network
 	px.reachable = make([]bool, len(px.peers))
-	for i, _ := range px.peers {
+	for i := range px.peers {
 		px.reachable[i] = true
 	}
 	px.deaf = false

### 2014-05-10 writeup/paper.tex
paper edits
added some charts....last commit before bed

@@
 
+These charts need to be re-done, numbers were off because of a code bug.
 \begin{table}[h]
@@
 
+\begin{figure}[H]
+\centering
+\makebox[\textwidth][c]{
+\includegraphics[scale=.66]{replicas.png}}
+\caption{Performance of a one-group configuration rapidly dteriorates as the number of replicas increases}
+\label{fig:replicas}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\makebox[\textwidth][c]{
+\includegraphics[scale=.66]{groups.png}}
+\caption{Performance increases linearly as the number of groups increases. There is probably a saturation point, but we didn't have enough cluster machines to reach it.}
+\label{fig:groups}
+\end{figure}
+
 Increasing the number of replica groups increases the performance of
@@
 \section{Conclusion}
+We did a project.
 

### 2014-05-10 src/shardmaster/server.go
Small shardmaster fix for debug printout
One of the debug printouts had referenced sm.configs[config] which
causes a panic if writeToMemory is false since the map entry is nil.

@@ -297,7 +297,6 @@ func (sm *ShardMaster) Join(args *JoinArgs, reply *JoinReply) error {
 	}
 	sm.mu.Lock()
 
-
 	newOp := Op{2, args.GID, args.Servers, 0}
 
 	for !sm.dead {
@@ -817,7 +816,7 @@ func (sm *ShardMaster) startup(servers []string) {
 				if ok && !reply.Err {
 					replyConfig := reply.RequestedConfig
 					sm.putConfig(config, replyConfig)
-					DPrintfPersist("\n\t\t%v: Got %v for config %v", sm.me, *sm.configs[config], config)
+					DPrintfPersist("\n\t\t%v: Got %v for config %v", sm.me, replyConfig, config)
 					haveConfig = true
 					sm.maxConfig = config
 					break
@@ -875,7 +874,7 @@ func StartServer(servers []string, me int, network bool) *ShardMaster {
 		}
 		enableLog()
 	}
-	DPrintf("Shardamaster %s server started with peers",servers[me],servers)
+	DPrintf("Shardamaster %s server started with peers", servers[me], servers)
 	sm := new(ShardMaster)
 	// Read memory options
 	sm.persistent = persistent

### 2014-05-10 levigo-ex.go
levigo benchmark

@@ -1,8 +1,20 @@
 package main
 import "github.com/jmhodges/levigo"
 import "fmt"
+import "math/rand"
+import "time"
+func trace(s string) (string, time.Time) {
+    fmt.Println("START:", s)
+    return s, time.Now()
+}
 
-func main() {
+func un(s string, startTime time.Time) {
+    endTime := time.Now()
+    fmt.Println("  END:", s, "ElapsedTime in seconds:", endTime.Sub(startTime))
+}
+func db_create() (*levigo.DB) {
+	defer un(trace("db create"))
+
 	opts := levigo.NewOptions()
 	opts.SetCache(levigo.NewLRUCache(3<<30))
 	opts.SetCreateIfMissing(true)
@@ -12,5 +24,37 @@ func main() {
 	db.Put(wo, []byte("key1"), []byte("val1"))
 	data, _ := db.Get(ro, []byte("key1"))
 	fmt.Println(string(data));
+	return db
+}
 
+func gen_data(size uint64) ([]byte) {
+	defer un(trace("gen data"))
+	
+	data := make([]byte, size)
+//	src := new(rand.Rand)
+	for i := range data {
+		data[i] = byte(rand.Int31() & 0xff)
+	}
+	return data
+}
+
+func write_data(data []byte, db *levigo.DB, nitems int, keysize int, valsize int) {
+	defer un(trace("write data"))
+	wo := levigo.NewWriteOptions()	
+	keystart := 0
+	valstart := 0
+	for i:=0; i < nitems; i++ {
+		db.Put(wo, 
+			data[keystart:keystart + keysize], 
+			data[valstart:valstart + valsize])
+		keystart %= (len(data) - keysize)
+		valstart %= (len(data) - valsize)
+	}
+	fmt.Println("Wrote ", nitems, "items with key size", keysize, "and value size", valsize)
+}
+
+func main() {
+	db := db_create()
+	data := gen_data(1 << 24)
+	write_data(data, db, 10000000, 16, 256)
 }

### 2014-05-10 levigo-ex.go
tweak parameters in levigo

@@ -56,5 +56,5 @@ func write_data(data []byte, db *levigo.DB, nitems int, keysize int, valsize int
 func main() {
 	db := db_create()
 	data := gen_data(1 << 24)
-	write_data(data, db, 10000000, 16, 256)
+	write_data(data, db, 1000000, 4, 4)
 }

### 2014-05-10 writeup/paper.tex
added some more stuff to writeup

@@
 Below we describe the different tests. All tests are located at
-'mexos/src/<name>/test_test.go' unless otherwise noted.
+'mexos/src/$<$name$>$/test\_test.go' unless otherwise noted.
 
@@
 
-These charts need to be re-done, numbers were off because of a code bug.
+% Please add the following required packages to your document preamble:
+% \usepackage[table,xcdraw]{xcolor}
+% If you use beamer only pass ``xcolor=table'' option, i.e. \documentclass[xcolor=table]{beamer}
 \begin{table}[h]
@@
 \multicolumn{4}{|c|}{\textit{PutHash to one key}}                                                                     \\ \hline
-\textbf{1 group, 3 replicas}  & 21.6ms/op                    & 46 requests/s                & 85 requests/s           \\ \hline
-\textbf{1 group, 10 replicas} & 52.7ms/op                    & 19 requests/s                & 19 requests/s           \\ \hline
-\textbf{3 groups, 3 replicas} & 15.4ms/op                    & 65 requests/s                & 70 requests/s           \\ \hline
+\textbf{1 group, 3 replicas}  & 40.3 ms/op                   & 24.8 requests/s              & 27 requests/s           \\ \hline
+\textbf{1 group, 10 replicas} & 5370 ms/op                   & 0.18 requests/s              & 2 requests/s            \\ \hline
+\textbf{3 groups, 3 replicas} & 62 ms/op                     &                              & 22 requests/s           \\ \hline
 \multicolumn{4}{|c|}{\textit{PutHash to many keys}}                                                                   \\ \hline
-\textbf{1 group, 3 replicas}  & 20.5ms/op                    & 49 requests/s                & 78 requests/s           \\ \hline
-\textbf{1 group, 10 replicas} & 74ms/op                      & 14 requests/s                & 19 requests/s           \\ \hline
-\textbf{3 groups, 3 replicas} & 21.8ms/op                    & 46 requests/s                & 179 requests/s          \\ \hline
+\textbf{1 group, 3 replicas}  & 40.7 ms/op                   & 24.5 requests/s              & 30 requests/s           \\ \hline
+\textbf{1 group, 10 replicas} & 282 ms/op                    & 3.5 requests/s               & 7 requests/s            \\ \hline
+\textbf{3 groups, 3 replicas} & 214 ms/op                    & 4.6 requests/s               & 67 requests/s           \\ \hline
 \end{tabular}
@@
 
-%----------------------------------------------------------------------
-
+% Please add the following required packages to your document preamble:
+% \usepackage[table,xcdraw]{xcolor}
+% If you use beamer only pass ``xcolor=table'' option, i.e. \documentclass[xcolor=table]{beamer}
 \begin{table}[h]
@@
 \multicolumn{4}{|c|}{\textit{PutHash to one key}}                                                                     \\ \hline
-\textbf{1 group, 3 replicas}  & 14.2ms/op                    & 70.4 requests/s              & 88.8 requests/s         \\ \hline
-\textbf{1 group, 10 replicas} & 36.9ms/op                    & 27 requests/s                & 31 requests/s           \\ \hline
-\textbf{3 groups, 3 replicas} & 14.1ms/op                    & 70.9 requests/s              & 86 requests/s           \\ \hline
+\textbf{1 group, 3 replicas}  & 27 ms/op                     & 37 requests/s                & 47 requests/s           \\ \hline
+\textbf{1 group, 10 replicas} & 2360 ms/op                   & 0.42 requests/s              & 3.4 requests/s          \\ \hline
+\textbf{3 groups, 3 replicas} & 22 ms/op                     & 45 requests/s                & 45 requests/s           \\ \hline
 \multicolumn{4}{|c|}{\textit{PutHash to many keys}}                                                                   \\ \hline
-\textbf{1 group, 3 replicas}  & 13.3ms/op                    & 75.2 requests/s              & 87.4 requests/s         \\ \hline
-\textbf{1 group, 10 replicas} & 37.7ms/op                    & 26.5 requests/s              & 31 requests/s           \\ \hline
-\textbf{3 groups, 3 replicas} & 15.8ms/op                    & 63.3 requests/s              & 200 requests/s          \\ \hline
+\textbf{1 group, 3 replicas}  & 25 ms/op                     & 40 requests/s                & 45 requests/s           \\ \hline
+\textbf{1 group, 10 replicas} & 130 ms/op                    & 7 requests/s                 & 10.3 requests/s         \\ \hline
+\textbf{3 groups, 3 replicas} & 40 ms/op                     & 25 requests/s                & 124 requests/s          \\ \hline
 \end{tabular}
@@
 
+For a single group case, the difference between three replicas and ten is
+striking. The more replicas, the more traffic is needed for a paxos
+agreement. Figure 1 demonstrates how throughput suffers as the number
+of replicas increases. One other interesting effect is that as the
+number of replicas increases, the larger the gap there is between the
+single key and multi-key cases. Because there is one group, all puts
+touch the same servers. We are unsure what causes this.
+
+Adding more groups increases throughput, however. For evenly
+distributed traffic, going from one group to 3 nearly triples the
+throughput. Figure 2 shows that this trend continues, so adding groups
+linearly improves the throughput of the system. The table shows that
+average throughput for a single client may suffer, but this is
+probably because the system throughput benchmark creates many clients,
+while the client throughput test creates a single client. Having
+multiple clients increases the liklihood that a client must wait to
+see its request served, since the replicas may be busy serving another
+client's request.
+
+\begin{figure}[H]
+\centering
+\makebox[\textwidth][c]{
+\includegraphics[scale=.50]{replicas.png}}
+\caption{Performance of a one-group configuration rapidly deteriorates as the number of replicas increases}
+\label{fig:replicas}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\makebox[\textwidth][c]{
+\includegraphics[scale=.5]{groups.png}}
+\caption{Performance increases linearly as the number of groups increases. There is probably a saturation point, but we didn't have enough cluster machines to reach it.}
+\label{fig:groups}
+\end{figure}
+
 \subsubsection{Paxos RPC Counts}
 
+Multipaxos decreases the number of RPCs needed to reach agreement.
+
+% Please add the following required packages to your document preamble:
+% \usepackage[table,xcdraw]{xcolor}
+% If you use beamer only pass ``xcolor=table'' option, i.e. \documentclass[xcolor=table]{beamer}
+\begin{table}[H]
+\centering
+\makebox[\textwidth][c]{
+\begin{tabular}{|l|l|l|}
+\hline
+\multicolumn{3}{|c|}{\cellcolor[HTML]{C0C0C0}\textbf{Paxos RPC Counts}}                  \\ \hline
+                                 & \textbf{Single Proposer} & \textbf{Multiple Proposer} \\ \hline
+\textbf{Normal Paxos}            & 30                       & 89            \\ \hline
+\textbf{Paxos with Leaders}      & 20                       & 27                         \\ \hline
+\textbf{Leaders and Pre-prepare} & 17                       & 20                         \\ \hline
+\end{tabular}
+}
+\end{table}
+
 \subsubsection{Disk Recovery}
-For disk recovery, the bottleneck was...
+Raw time to write 1GB to database, HDD:  21.470750409s
+
+Raw time to write 1GB to database, SSD:  6.1470524260s
+
+Recovery time of a 1GB database using our system: TBD
 
@@
 of replicas has an adverse effect on performance since each operation
-requires communication with a majority of replicas. Using multipaxos
-improves the performance since it decreases the number of RPCs needed
-to commit an operation.
-
-\begin{figure}[H]
-\centering
-\makebox[\textwidth][c]{
-\includegraphics[scale=.66]{replicas.png}}
-\caption{Performance of a one-group configuration rapidly dteriorates as the number of replicas increases}
-\label{fig:replicas}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-\makebox[\textwidth][c]{
-\includegraphics[scale=.66]{groups.png}}
-\caption{Performance increases linearly as the number of groups increases. There is probably a saturation point, but we didn't have enough cluster machines to reach it.}
-\label{fig:groups}
-\end{figure}
+requires communication with a majority of replicas.
 
@@
 
-Another benefit of the paxos optimizations is that with
-fewer messages, less disk space is used since less paxos state data
-needs to be saved. We also suspect that using UDP would also improve
-throughput, as servers that do not recieve a response already
-re-transmit their request, making the TCP ack system redundant.
+Using multipaxos reduces the amount of network communication required,
+which improves the throughput and latency by about a factor of
+two. Multipaxos also reduces the amount of paxos state information,
+since prepares are pre-determined. This saves some disk space.  We
+also suspect that using UDP would also improve throughput, as servers
+that do not recieve a response already re-transmit their request,
+making the TCP ack system redundant.
 

### 2014-05-10 src/shardkv/server.go
added recovery benchmark

@@ -24,9 +24,9 @@ import "runtime"
 import "os/exec"
 
 const Debug = 0
-const DebugPersist = 0
+const DebugPersist = 1
 const printRPCerrors = false
-const Log = 0
+const Log = 1
 
 var logfile *os.File
 
@@ -56,7 +56,7 @@ func DPrintf(format string, a ...interface{}) (n int, err error) {
 
 func DPrintfPersist(format string, a ...interface{}) (n int, err error) {
 	if DebugPersist > 0 {
-		log.Printf(format, a...)
+		fmt.Printf(format, a...)
 	}
 	return
 }
@@ -1301,11 +1301,11 @@ func (kv *ShardKV) dbInit() {
 func (kv *ShardKV) startup(servers []string) {
 	defer func() {
 		kv.recovering = false
-		DPrintfPersist("\n%v-%v Marked recovery false", kv.gid, kv.me)
+		log.Printf("\n%v-%v Marked recovery false", kv.gid, kv.me)
 	}()
 	// Initialize database, check if state is stored
 	kv.recovering = true
-	DPrintfPersist("\n%v-%v Marked recovery true", kv.gid, kv.me)
+	log.Printf("\n%v-%v Marked recovery true", kv.gid, kv.me)
 	kv.dbInit()
 	if !recovery {
 		return

### 2014-05-10 src/test/test_test.go

@@ -400,3 +400,22 @@ func TestManyClientOneShard(t *testing.T) {
 	}
 	fmt.Printf("%f operations per second\n", float64(tot)/float64(nseconds))
 }
+
+
+func TestDiskRecovery(t *testing.T) {
+	numBytes := 1073741824 //1GB
+	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
+	//defer clean()
+	
+	fmt.Printf("\nDisk recovery benchmark...\n")
+
+	smClerk := shardmaster.MakeClerk(smPorts, true)
+	smClerk.Join(gids[0], kvPorts[0])
+	ck := shardkv.MakeClerk(smPorts, true)
+
+	for i := 0; i < numBytes; i+=16 {
+		ck.Put(strconv.Itoa(rand.Int()), strconv.Itoa(rand.Int()))
+	}
+	
+	fmt.Printf("\n1GB of data written to database...\n")
+}

### 2014-05-10 src/shardkv/server.go
changed flags

@@ -24,7 +24,7 @@ import "runtime"
 import "os/exec"
 
 const Debug = 0
-const DebugPersist = 1
+const DebugPersist = 0
 const printRPCerrors = false
 const Log = 1
 

### 2014-05-10 src/test/test_test.go

@@ -12,7 +12,7 @@ import "sync"
 import "math/rand"
 import "log"
 
-const numGroups = 3
+const numGroups = 1
 const numReplicas = 3
 const numMasters = 3
 
@@ -403,7 +403,7 @@ func TestManyClientOneShard(t *testing.T) {
 
 
 func TestDiskRecovery(t *testing.T) {
-	numBytes := 1073741824 //1GB
+	numBytes := 536870912 //0.5GB
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	
@@ -415,6 +415,9 @@ func TestDiskRecovery(t *testing.T) {
 
 	for i := 0; i < numBytes; i+=16 {
 		ck.Put(strconv.Itoa(rand.Int()), strconv.Itoa(rand.Int()))
+		if i%1024 == 0 {
+			fmt.Printf("%v/%v\n", i, numBytes)
+		}
 	}
 	
 	fmt.Printf("\n1GB of data written to database...\n")

### 2014-05-10 src/test/test_test.go
enabld leadre

@@ -403,7 +403,8 @@ func TestManyClientOneShard(t *testing.T) {
 
 
 func TestDiskRecovery(t *testing.T) {
-	numBytes := 536870912 //0.5GB
+	nclients := 2
+	numBytes := 20971520 //20MB
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	
@@ -411,14 +412,32 @@ func TestDiskRecovery(t *testing.T) {
 
 	smClerk := shardmaster.MakeClerk(smPorts, true)
 	smClerk.Join(gids[0], kvPorts[0])
-	ck := shardkv.MakeClerk(smPorts, true)
 
-	for i := 0; i < numBytes; i+=16 {
-		ck.Put(strconv.Itoa(rand.Int()), strconv.Itoa(rand.Int()))
-		if i%1024 == 0 {
-			fmt.Printf("%v/%v\n", i, numBytes)
+	counts := make([]int, nclients)
+	fmt.Printf("\nBenchmark: many clients, many shards...\n")
+	for i := 0; i < nclients; i++ {
+		go func(c int) {
+			ck := shardkv.MakeClerk(smPorts, true)
+			for i := 0; i < numBytes/5; i+=16 {
+				ck.Put(strconv.Itoa(rand.Int()), 
+					strconv.Itoa(rand.Int()))
+				counts[c]+=16
+			}
+		}(i)
+	}
+	fmt.Println("I am here")
+	sum := 0
+	toWait := 22*time.Millisecond
+	for sum < numBytes {
+		time.Sleep(toWait)
+		for _,i := range counts {
+			sum = 0
+			sum += i
+		}
+		if sum%1024 == 0 {
+			fmt.Printf("%v/%v\n", sum, numBytes)
 		}
 	}
-	
+
 	fmt.Printf("\n1GB of data written to database...\n")
 }

### 2014-05-10 src/paxos/paxos.go
Updated shardkv recovery protocol

@@ -1146,7 +1146,6 @@ func (px *Paxos) startup(tag string) {
 					px.maxInstance = reply.MaxInstance
 				}
 				haveState = true
-				break
 			}
 		}
 	}

### 2014-05-10 src/shardkv/common.go

@@ -45,6 +45,7 @@ type FetchArgs struct {
 	Config  int
 	Shard   int
 	Exclude map[string]bool
+	Sender  string
 }
 
 type FetchReply struct {
@@ -59,6 +60,7 @@ type RecoverArgs struct {
 	Config  int
 	Shard   int
 	Exclude map[string]bool
+	Sender  string
 }
 
 type RecoverReply struct {

### 2014-05-10 src/shardkv/server.go

@@ -104,6 +104,8 @@ type ShardKV struct {
 	db             *levigo.DB
 	dbLock         sync.Mutex
 	recovering     bool
+	sending        bool
+	sendingTo      string
 }
 
 // Write the desired key/value to memory and/or disk
@@ -348,7 +350,7 @@ func (kv *ShardKV) addReconfigure(num int, store map[string]string, response map
 
 // Accept a Get request
 func (kv *ShardKV) Get(args *GetArgs, reply *KVReply) error {
-	for kv.recovering && !kv.dead {
+	for (kv.recovering || kv.sending) && !kv.dead {
 		time.Sleep(10 * time.Millisecond)
 	}
 	kv.mu.Lock()
@@ -377,7 +379,7 @@ func (kv *ShardKV) Put(args *PutArgs, reply *KVReply) error {
 	} else {
 		DPrintf("%d.%d.%d) Put: %s -> %s\n", kv.gid, kv.me, kv.config.Num, args.Key, args.Value)
 	}
-	for kv.recovering && !kv.dead {
+	for (kv.recovering || kv.sending) && !kv.dead {
 		time.Sleep(10 * time.Millisecond)
 	}
 	kv.mu.Lock()
@@ -411,7 +413,21 @@ func (kv *ShardKV) Fetch(args *FetchArgs, reply *FetchReply) error {
 	for kv.recovering && !kv.dead {
 		time.Sleep(10 * time.Millisecond)
 	}
+	for kv.sending && args.Sender != kv.sendingTo {
+		time.Sleep(10 * time.Millisecond)
+	}
 	return kv.fetchHandler(args, reply)
+}
+
+// Respond to acknowledgement that Fetch is complete
+func (kv *ShardKV) FetchComplete(args *FetchArgs, reply *FetchReply) error {
+	//if args.Sender == kv.sendingTo {
+	kv.sending = false
+	kv.sendingTo = ""
+	DPrintf("\n%v.%v: Marking sending complete", kv.gid, kv.me)
+	reply.Complete = true
+	//}
+	return nil
 }
 
 // This helper "fetch" method now exists because both Fetch
@@ -420,7 +436,6 @@ func (kv *ShardKV) Fetch(args *FetchArgs, reply *FetchReply) error {
 func (kv *ShardKV) fetchHandler(args *FetchArgs, reply *FetchReply) error {
 	//kv.mu.Lock()
 	defer func() {
[…]

### 2014-05-10 src/shardkv/test_test.go

@@ -101,7 +101,7 @@ func TestFileBasic(t *testing.T) {
 	if !runOldTests {
 		return
 	}
-	numGroups := 4
+	numGroups := 3
 	numReplicas := 2
 	smPorts, gids, kvPorts, kvServers, clean := setup("basic", false, numGroups, numReplicas)
 	defer clean()
@@ -153,20 +153,20 @@ func TestFileBasic(t *testing.T) {
 	// are keys still there after leaves?
 	for g := 0; g < len(gids)-1; g++ {
 		smClerk.Leave(gids[g])
-		fmt.Printf("group 10%d left\n",g)
+		fmt.Printf("group 10%d left\n", g)
 		time.Sleep(1 * time.Second)
 		for i := 0; i < len(keys); i++ {
-			fmt.Printf("getting %v\n",keys[i])
+			fmt.Printf("getting %v\n", keys[i])
 			v := kvClerk.Get(keys[i])
-			fmt.Printf("got %v",keys[i])
+			fmt.Printf("got %v", keys[i])
 			if v != vals[i] {
 				t.Fatalf("leaving; wrong value; g=%v k=%v wanted=%v got=%v",
 					g, keys[i], vals[i], v)
 			}
 			vals[i] = strconv.Itoa(rand.Int())
-			fmt.Printf("putting %v\n",keys[i])
+			fmt.Printf("putting %v\n", keys[i])
 			kvClerk.Put(keys[i], vals[i])
-			fmt.Printf("put success %v\n",keys[i])
+			fmt.Printf("put success %v\n", keys[i])
 		}
 	}
 
@@ -275,7 +275,9 @@ func TestFileLimp(t *testing.T) {
 	}
 
 	for g := 0; g < len(kvServers); g++ {
-		kvServers[g][rand.Int()%len(kvServers[g])].Kill()
+		toKill := rand.Int() % len(kvServers[g])
+		//fmt.Printf("\nKilling %v-%v = %v", g, toKill, g*3+toKill)
+		kvServers[g][toKill].Kill()
 	}
 
 	keys := make([]string, 10)
@@ -304,7 +306,7 @@ func TestFileLimp(t *testing.T) {
 	// are keys still there after leaves?
 	for g := 0; g < len(gids)-1; g++ {
 		smClerk.Leave(gids[g])
-		time.Sleep(2 * time.Second)
+		time.Sleep(5 * time.Second)
[…]

### 2014-05-10 src/shardmaster/server.go

@@ -794,7 +794,6 @@ func (sm *ShardMaster) startup(servers []string) {
 					sm.dbWriteProcessedSeq(sm.processedSeq)
 				}
 				haveState = true
-				break
 			}
 		}
 	}

### 2014-05-10 src/shardkv/server.go
Minor timing changes

@@ -26,7 +26,7 @@ import "os/exec"
 const Debug = 0
 const DebugPersist = 0
 const printRPCerrors = false
-const Log = 1
+const Log = 0
 
 var logfile *os.File
 
@@ -660,7 +660,7 @@ func (kv *ShardKV) tick() {
 						}
 						if !ok {
 							numTries++
-							time.Sleep(5 * time.Millisecond)
+							time.Sleep(50 * time.Millisecond)
 						}
 					}
 				}
@@ -1384,11 +1384,11 @@ func (kv *ShardKV) dbInit() {
 func (kv *ShardKV) startup(servers []string) {
 	defer func() {
 		kv.recovering = false
-		log.Printf("\n%v-%v Marked recovery false", kv.gid, kv.me)
+		DPrintfPersist("\n%v-%v Marked recovery false", kv.gid, kv.me)
 	}()
 	// Initialize database, check if state is stored
 	kv.recovering = true
-	log.Printf("\n%v-%v Marked recovery true", kv.gid, kv.me)
+	DPrintfPersist("\n%v-%v Marked recovery true", kv.gid, kv.me)
 	kv.dbInit()
 	if !recovery {
 		return
@@ -1518,7 +1518,7 @@ func (kv *ShardKV) startup(servers []string) {
 					}
 					if !ok {
 						numTries++
-						time.Sleep(5 * time.Millisecond)
+						time.Sleep(50 * time.Millisecond)
 					}
 				}
 			}

### 2014-05-10 src/shardkv/test_test.go

@@ -306,7 +306,7 @@ func TestFileLimp(t *testing.T) {
 	// are keys still there after leaves?
 	for g := 0; g < len(gids)-1; g++ {
 		smClerk.Leave(gids[g])
-		time.Sleep(5 * time.Second)
+		time.Sleep(10 * time.Second)
 		for i := 0; i < len(kvServers[g]); i++ {
 			kvServers[g][i].Kill()
 		}

### 2014-05-11 src/shardkv/server.go
A few more shardkv timing changes

@@ -41,6 +41,7 @@ const dbUseCache = true                        // Whether database should use a 
 const dbCacheSize = 250                        // Size of database cache in MB (ignored if dbUseCache is false)
 const memoryLimit = 1000                       // Memory limit in MB
 const memoryThreshold = memoryLimit * 80 / 100 // When to stop filling memory (when to abort a Fetch RPC and use multiple messages)
+const recoveryRetryDelay = 500                 // Time in ms to wait before resending acknowledgments
 
 // Will use these to check that dbCacheSize doesn't overflow an int
 // (int size is either 32 or 64 bits depending on implementation)
@@ -618,7 +619,7 @@ func (kv *ShardKV) tick() {
 										ackOK := call(server, "ShardKV.FetchComplete", ackArgs, &ackReply, kv.network)
 										ackSuccess = ackOK && ackReply.Complete
 										if !ackSuccess {
-											time.Sleep(10 * time.Millisecond)
+											time.Sleep(recoveryRetryDelay * time.Millisecond)
 										}
 									}
 									DPrintf("\n%v.%v: Done sending fetch complete to %s", kv.gid, kv.me, server)
@@ -651,7 +652,7 @@ func (kv *ShardKV) tick() {
 									ackOK := call(server, "ShardKV.FetchComplete", ackArgs, &ackReply, kv.network)
 									ackSuccess = ackOK && ackReply.Complete
 									if !ackSuccess {
-										time.Sleep(10 * time.Millisecond)
+										time.Sleep(recoveryRetryDelay * time.Millisecond)
 									}
 								}
 								DPrintf("\n%v.%v: Done sending fetch complete to %s", kv.gid, kv.me, server)
@@ -660,7 +661,7 @@ func (kv *ShardKV) tick() {
 						}
 						if !ok {
 							numTries++
-							time.Sleep(50 * time.Millisecond)
+							time.Sleep(100 * time.Millisecond)
 						}
 					}
 				}
@@ -1476,7 +1477,7 @@ func (kv *ShardKV) startup(servers []string) {
 									ackOK := call(srv, "ShardKV.FetchComplete", ackArgs, &ackReply, kv.network)
 									ackSuccess = ackOK && ackReply.Complete
 									if !ackSuccess {
-										time.Sleep(10 * time.Millisecond)
+										time.Sleep(recoveryRetryDelay * time.Millisecond)
 									}
[…]

### 2014-05-11 src/shardkv/server.go
Another small shardkv recovery update

@@ -454,32 +454,36 @@ func (kv *ShardKV) fetchHandler(args *FetchArgs, reply *FetchReply) error {
 	kv.sending = true
 	kv.sendingTo = args.Sender
 
-	// Assume all responses can fit in memory (only one per client)
 	responses := make(map[int64]string)
-	idsInMemory := make(map[int64]bool)
-	// Copy responses from memory
-	for id, value := range kv.response {
-		responses[id] = value
-		idsInMemory[id] = true
-	}
-	// Copy responses from disk if not in memory
-	for id, value := range kv.dbGetResponses(idsInMemory) {
-		DPrintfPersist("\n\t%v-%v: got response data (%v, %v)", kv.gid, kv.me, id, value)
-		responses[id] = value
-	}
-
-	// Assume all of seenIDs can fit in memory (IDs should be small)
 	seenIDs := make(map[int64]bool)
-	idsInMemory = make(map[int64]bool)
-	// Copy seen IDs from memory
-	for id, _ := range kv.seen {
-		seenIDs[id] = true
-		idsInMemory[id] = true
-	}
-	// Copy seen IDs from disk if not in memory
-	for id, _ := range kv.dbGetSeenIDs(idsInMemory) {
-		DPrintfPersist("\n\t%v-%v: got seen data %v", kv.gid, kv.me, id)
-		seenIDs[id] = true
+
+	// If this is the first message, include responses and seenIDs
+	if len(args.Exclude) == 0 {
+		// Assume all responses can fit in memory (only one per client)
+		idsInMemory := make(map[int64]bool)
+		// Copy responses from memory
+		for id, value := range kv.response {
+			responses[id] = value
+			idsInMemory[id] = true
+		}
+		// Copy responses from disk if not in memory
+		for id, value := range kv.dbGetResponses(idsInMemory) {
+			DPrintfPersist("\n\t%v-%v: got response data (%v, %v)", kv.gid, kv.me, id, value)
+			responses[id] = value
+		}
+
+		// Assume all of seenIDs can fit in memory (IDs should be small)
+		idsInMemory = make(map[int64]bool)
+		// Copy seen IDs from memory
+		for id, _ := range kv.seen {
+			seenIDs[id] = true
+			idsInMemory[id] = true
+		}
+		// Copy seen IDs from disk if not in memory
+		for id, _ := range kv.dbGetSeenIDs(idsInMemory) {
[…]

### 2014-05-11 src/test/test_test.go
bigger keys for recovery test

@@ -401,10 +401,21 @@ func TestManyClientOneShard(t *testing.T) {
 	fmt.Printf("%f operations per second\n", float64(tot)/float64(nseconds))
 }
 
+func paddedRandIntString(size uint64) string {
+	data := make([]byte, size)
+	s := strconv.Itoa(rand.Int())
+	for i := range i {
+		data[i] = s[i % len(s)]
+	}
+	return string(data[:])
+}
 
 func TestDiskRecovery(t *testing.T) {
 	nclients := 2
-	numBytes := 20971520 //20MB
+//	numBytes := 20971520 //20MB
+	nItems = 100000
+	keySize := 32
+	valSize := 512
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	
@@ -418,9 +429,10 @@ func TestDiskRecovery(t *testing.T) {
 	for i := 0; i < nclients; i++ {
 		go func(c int) {
 			ck := shardkv.MakeClerk(smPorts, true)
-			for i := 0; i < numBytes/5; i+=16 {
-				ck.Put(strconv.Itoa(rand.Int()), 
-					strconv.Itoa(rand.Int()))
+			for i := 0; i < nItems; i++ {
+				ck.Put(paddedRandIntString(keySize), 
+					paddedRandIntString(valSize)
+				)
 				counts[c]+=16
 			}
 		}(i)

### 2014-05-11 src/test/test_test.go
typos

@@ -401,10 +401,10 @@ func TestManyClientOneShard(t *testing.T) {
 	fmt.Printf("%f operations per second\n", float64(tot)/float64(nseconds))
 }
 
-func paddedRandIntString(size uint64) string {
+func paddedRandIntString(size int) string {
 	data := make([]byte, size)
 	s := strconv.Itoa(rand.Int())
-	for i := range i {
+	for i := range data {
 		data[i] = s[i % len(s)]
 	}
 	return string(data[:])
@@ -413,7 +413,7 @@ func paddedRandIntString(size uint64) string {
 func TestDiskRecovery(t *testing.T) {
 	nclients := 2
 //	numBytes := 20971520 //20MB
-	nItems = 100000
+	nItems := 100000
 	keySize := 32
 	valSize := 512
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
@@ -430,9 +430,7 @@ func TestDiskRecovery(t *testing.T) {
 		go func(c int) {
 			ck := shardkv.MakeClerk(smPorts, true)
 			for i := 0; i < nItems; i++ {
-				ck.Put(paddedRandIntString(keySize), 
-					paddedRandIntString(valSize)
-				)
+				ck.Put(paddedRandIntString(keySize), paddedRandIntString(valSize))
 				counts[c]+=16
 			}
 		}(i)
@@ -440,14 +438,14 @@ func TestDiskRecovery(t *testing.T) {
 	fmt.Println("I am here")
 	sum := 0
 	toWait := 22*time.Millisecond
-	for sum < numBytes {
+	for sum < nItems {
 		time.Sleep(toWait)
+		sum = 0
 		for _,i := range counts {
-			sum = 0
 			sum += i
 		}
 		if sum%1024 == 0 {
-			fmt.Printf("%v/%v\n", sum, numBytes)
+			fmt.Printf("%v/%v\n", sum, nItems)
 		}
 	}
 

### 2014-05-11 src/test/test_test.go
tweak parameters for test

@@ -415,7 +415,7 @@ func TestDiskRecovery(t *testing.T) {
 //	numBytes := 20971520 //20MB
 	nItems := 100000
 	keySize := 32
-	valSize := 512
+	valSize := 4096
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	
@@ -449,5 +449,5 @@ func TestDiskRecovery(t *testing.T) {
 		}
 	}
 
-	fmt.Printf("\n1GB of data written to database...\n")
+	fmt.Printf("\n", nItems * (keySize + valSize)/1000000, " MB of data written to database...\n")
 }

### 2014-05-11 src/test/test_test.go
more tweaks

@@ -449,5 +449,5 @@ func TestDiskRecovery(t *testing.T) {
 		}
 	}
 
-	fmt.Printf("\n", nItems * (keySize + valSize)/1000000, " MB of data written to database...\n")
+	fmt.Println("\n", nItems * (keySize + valSize) * nclients/1000000, " MB of data written to database...\n")
 }

### 2014-05-11 src/shardkv/server.go
recovery updates, transfer speed test (local)

@@ -38,9 +38,9 @@ const recovery = true
 const writeToMemory = false                    // Whether responses/store should be written to memory (as well as disk / disk cache)
 const dbUseCompression = true                  // Whether database should compress entries
 const dbUseCache = true                        // Whether database should use a built-in cache
-const dbCacheSize = 250                        // Size of database cache in MB (ignored if dbUseCache is false)
-const memoryLimit = 1000                       // Memory limit in MB
-const memoryThreshold = memoryLimit * 80 / 100 // When to stop filling memory (when to abort a Fetch RPC and use multiple messages)
+const dbCacheSize = 20                         // Size of database cache in MB (ignored if dbUseCache is false)
+const memoryLimit = 100                        // Memory limit in MB
+const memoryThreshold = memoryLimit * 75 / 100 // When to stop filling memory (when to abort a Fetch RPC and use multiple messages)
 const recoveryRetryDelay = 500                 // Time in ms to wait before resending acknowledgments
 
 // Will use these to check that dbCacheSize doesn't overflow an int
@@ -107,6 +107,7 @@ type ShardKV struct {
 	recovering     bool
 	sending        bool
 	sendingTo      string
+	shardIterator  *levigo.Iterator
 }
 
 // Write the desired key/value to memory and/or disk
@@ -451,6 +452,8 @@ func (kv *ShardKV) fetchHandler(args *FetchArgs, reply *FetchReply) error {
 	// Mark sending as true so no client requests will
 	// be processed until the transfer is complete
 	DPrintf("\n%v.%v: Marking sending started", kv.gid, kv.me)
+	//fmt.Printf("\n%v.%v: Marking sending started", kv.gid, kv.me)
+	startTime := time.Now()
 	kv.sending = true
 	kv.sendingTo = args.Sender
 
@@ -485,6 +488,7 @@ func (kv *ShardKV) fetchHandler(args *FetchArgs, reply *FetchReply) error {
 			seenIDs[id] = true
 		}
 	}
[…]

### 2014-05-11 src/shardkv/test_test.go

@@ -38,31 +38,33 @@ func NextValue(hprev string, val string) string {
 }
 
 // Kill shardmaster servers
-func smCleanup(smServers []*shardmaster.ShardMaster) {
-	//fmt.Printf("\nsmCleanup")
+func smCleanup(smServers []*shardmaster.ShardMaster, saveDisk bool) {
 	for i := 0; i < len(smServers); i++ {
 		if smServers[i] != nil {
-			//fmt.Printf("\n\tKilling %v", i)
-			smServers[i].Kill()
-		} else {
-			//fmt.Printf("\n\t%v is nil", i)
+			if saveDisk {
+				smServers[i].KillSaveDisk()
+			} else {
+				smServers[i].Kill()
+			}
 		}
 	}
 }
 
 // Kill shardkv servers
-func shardkvCleanup(kvServers [][]*ShardKV) {
-	//fmt.Printf("\nshardkvCleanup")
+func shardkvCleanup(kvServers [][]*ShardKV, saveDisk bool) {
 	for i := 0; i < len(kvServers); i++ {
 		for j := 0; j < len(kvServers[i]); j++ {
-			//fmt.Printf("\n\tKilling %v-%v", i, j)
-			kvServers[i][j].Kill()
+			if saveDisk {
+				kvServers[i][j].KillSaveDisk()
+			} else {
+				kvServers[i][j].Kill()
+			}
 		}
 	}
 }
 
 // Set up and start servers for shardkv groups and shardmaster
-func setup(tag string, unreliable bool, numGroups int, numReplicas int) ([]string, []int64, [][]string, [][]*ShardKV, func()) {
+func setup(tag string, unreliable bool, numGroups int, numReplicas int) ([]string, []int64, [][]string, [][]*ShardKV, func(), func()) {
 	runtime.GOMAXPROCS(4)
 
 	const numMasters = 3
@@ -93,8 +95,9 @@ func setup(tag string, unreliable bool, numGroups int, numReplicas int) ([]strin
 		}
 	}
 
-	clean := func() { shardkvCleanup(kvServers); smCleanup(smServers) }
-	return smPorts, gids, kvPorts, kvServers, clean
+	clean := func() { shardkvCleanup(kvServers, false); smCleanup(smServers, false) }
+	cleanSaveDisk := func() { shardkvCleanup(kvServers, true); smCleanup(smServers, true) }
+	return smPorts, gids, kvPorts, kvServers, clean, cleanSaveDisk
 }
 
[…]

### 2014-05-11 src/shardkv/test_test.go
Performed 500MB transfer test (local)

@@ -710,7 +710,7 @@ func paddedRandIntString(size int) string {
 //	time.Sleep(10 * time.Second)
 //}
 
-func TestDiskRecovery(t *testing.T) {
+func TestDiskTransfer(t *testing.T) {
 	nclients := 3
 	keySize := 32         // Size in bytes
 	valSize := 500 * 1024 // Size in bytes

### 2014-05-11 src/shardkv/server.go
changed debugging

@@ -1409,11 +1409,11 @@ func (kv *ShardKV) dbInit() {
 func (kv *ShardKV) startup(servers []string) {
 	defer func() {
 		kv.recovering = false
-		DPrintfPersist("\n%v-%v Marked recovery false", kv.gid, kv.me)
+		log.Printf("\n%v-%v Marked recovery false", kv.gid, kv.me)
 	}()
 	// Initialize database, check if state is stored
 	kv.recovering = true
-	DPrintfPersist("\n%v-%v Marked recovery true", kv.gid, kv.me)
+	log.Printf("\n%v-%v Marked recovery true", kv.gid, kv.me)
 	kv.dbInit()
 	if !recovery {
 		return

### 2014-05-11 writeup/paper.tex
changed debugging
Persistence section draft (may be too much info)

@@
 {\rm Joseph DelPreto}\\
-delpreto@mit.ed
+delpreto@mit.edu
 \and
@@
 \section{Introduction}
-This document is the writeup of our 6.824 final project. We did the
-default project, and built a persistent, fault-tolerant
-high-performance key/value store. For persistence, we used the go
-wrapper for the high-performance C++ leveldb key/value libray, and
-extended our code to recover from data loss upon server restart. In
-order to improve the performance of our system, we implemented two
-paxos optimizations. We used a multi-paxos like design to avoid the
-dueling leaders problem, and also reduce the number of RPCs for paxos
-agreeement by having the leader send out prepare messages ahead of
-time. To measure performance, we extended our existing tests, and
-added new ones. We also modified the RPC system to work on a real
-network and deployed our project on an AWS cluster to get realistic
-throughput and latency measures, as well as information about the
-speed of disk recovery.
+The presented project implements a persistent, fault-tolerant, high-performance key/value store.
+In order to achieve these goals, the system implements at-most-once semantics, relies upon Paxos for reaching consenus, shards the database across multiple replica groups, persistently stores values to disk, and implements protocols for automatically recovering and updating disk contents upon server reboot.
+In order to improve performance, the system features two paxos optimizations: a protocol similar to multi-paxos to avoid the dueling leaders problem, and the ability for leaders to send out prepare messages in advance to reduce the required number of RPCs for paxos agreeement.
+
+Extensive testing was done to evaluate the system's persistence, recovery, and performance by extending existing tests and creating new ones.
+After modifying the local RPC system to work on a real network, the system was deployed on an Amazon Web Services (AWS) cluster to obtain realistic measurements of throughput, latency, and disk recovery speed.
 
 \section{Design} \label{sec:design}
-Our starting point was lab 4. Before begining, we ran battery tests to
-figure out which combination of our prior labs created the fastest and
-most reliable starting point.
+The design is based upon the code developed throughout the semester for lab 4.
+There are therefore three basic components running on each server: Paxos processes to reach consenus for instances, a shard master to determine which groups are responsible for various shards of the database, and a shard key/value storage process to execute client Put/Get requests.
+The Paxos instances are used to create a consistent log of operations for both shard reconfiguration and client operations.
+
+Battery tests were run to determine which combination of our prior labs created the fastest and most reliable starting point.
+Once a working codebase was compiled, modifications were made to address three  major areas: Paxos optimizations, persistence (with disk recovery), and testing over a real network on AWS.
 
@@
 \subsection{Persistent Storage}
+In order to tolerate servers restarting and to handle databases that cannot fit in memory, all three components of the system store state and data persistently on disk.
+To interface with the disk, a Go wrapper for the high-performance C++ key/value libray LevelDB was utilized.
+An outline of what each system component stores on disk is presented below.
+\begin{itemize}
+\item The Paxos processes write the highest accepted value, the highest accepted proposal number, and the highest acknowledged prepare request to disk for each instance.
+This ensures that the protocol will operate correctly even if a server restarts (for example, a decided value can never be changed).
+In addition to this state, the Paxos processes persistently store the done responses received from peers and the number of the maximum known instance (see the recovery section below).
 
-\subsubsection{Recovery from failures}
+\item The shard master writes every configuration to disk, ensuring that will be able to respond to historical queries even after restarting.
+In addition, it stores the highest executed Paxos log entry and the number of its highest known configuration.
+
+\item The shard key/value service writes its database to disk by writing each value as each Put request is executed.
+In addition, it must preserve at-most-once semantics and consistency across reboots, so it persistently stores which operations have been seen as well as the most recent response sent to each client.
+It also writes the highest executed Paxos log entry and the current configuration number.
+\end{itemize}
+
+In all cases, the specified data is written to disk before returning to the caller, so any response that a caller receives from a peer will not be forgotten by the responder even if it restarts.
+It may be the case that the responder crashes before sending the response, but this is acceptable since the system offers at-most-once semantics (so the caller must be prepared to retry and the responder is prepared to filter duplicates).
+
+\subsubsection{Recovery from Failures}
+In order to tolerate servers restarting with or without their disk contents, the system includes protocols for recovering and catching up on startup.
+These can be summarized by considering what happens when a server starts:
+
+\begin{itemize}
+\item The Paxos process will load any saved done responses and maximum instance number from its disk.
+It will then contact peers to request their done responses and maximum instance number, determining which peer is the most updated out of any peers that respond.
+From these values, both the ones read from disk and the ones received, it can determine which instances it should know about but does not; it then requests these instances from its peers.
+\item The shard master process will load the highest known configuration and highest executed Paxos log instance from disk if it exists.
+It will then request the same information from its peers, determining who among the responders is the most updated.
+From this information it can determine if there are any missed configurations, and if so request them from a peer.
+\item The shard key/value service will load its last known configuration and the highest executed Paxos log instance from disk if it exists.
+It will then request the same information from peers, determining the most updated state that it should have.
+Finally, it can request the actual database key/value entries from a peer (see below for details of how this transfer is performed) along with any client responses and lists of seen operations.
+\end{itemize}
+These protocols allow any number of servers restarting with their disk contents to catch themselves up from any updated peer.
+They also allow any number of servers restarting without their disk contents to recover the needed data from any peer that has the data.
+
+\subsubsection{Memory and Performance Considerations}
+There are a number of design choices which were made in order to address both memory management and overall performance.
+
+Firstly, the service must be able to handle database content which is too large to fit in volatile memory.
+Towards this end, values are written to disk immediately as described above so it does not need to be kept in memory.
+More importantly, however, servers that need to transfer shards (either because the shard master decided a shard needs to move or because a server is recovering) must be able to do so in multiple RPCs since memory is too small to create a reply with the entire database.
+Therefore, a shard transfer occurrs with the following protocol:
+\begin{itemize}
+\item The requester sends a request to a peer, including a request number, and the current configuration number.  If doing a shard transfer, the shard number is also included.
+\item If the receiver is currently at the appropriate configuration, it accepts the request.
+It will then begin populating a reply with client response data, a list of seen operations, and the desired key/value data.
+While doing this, it will periodically check memory usage; if it exceeds a configurable threshold, it will reply with whatever it currently has and mark the reply as incomplete.
+\item The requester copies the data from the reply to its disk.  If the reply was marked as incomplete, it sends another request to the same peer with an incremented request number.
+\item This continues until the reply is marked as complete, at which point the requester launches a goroutine to make sure the sender receives an acknowledgement of the receipt of the final data message.
+\item If at any point the requester decides that the contacted peer is dead, it launches an acknowledgement goroutine as previously described for that peer (to terminate the transfer in case the peer wakes up or is not dead) and moves on to try a different peer, resetting the request number to 0.
+\end{itemize}
+Both the requester and the sender ignore client requests until the transfer is complete - this ensures that the requester will receive a consistent snapshot of the data (and is the reason the requester launches acknowledgement goroutines since otherwise a peer erroneously declared dead would be halted forever).
+
+One aspect which addresses performance is that progress can still be made while recovery is taking place as long as a majority of servers are not recovering.
+For example, if a replica group has three servers and a shard transfer is ocurring between two of them (perhaps because one just restarted), a client can still have its requests processed if it contacts the third peer.
+This is because the Paxos service runs independently of the shard key/value service; although the latter will ignore client requests on the transferring servers, their Paxos processes can still respond and make progress.
+Since the Paxos instances are stored persistently on disk, the two occupied servers can catch themselves up from the log when they complete the transfer.
+
+In addition, effort was made to utilize memory in addition to the disk in order to increase performance.
+The LevelDB database has the option to use a least-recently-used (LRU) cache to store recently used entries in memory.
+In addition, a configuration variable was added to the service to allow values to be written to memory as well as disk until the memory is full.
+These two options, combined with the native cache of the operating system, allow many needed values to be read quickly from memory rather than always consulting the disk.
 

### 2014-05-11 writeup/paper.tex
added to writeup

--- 2014-05-10 writeup/paper.tex
@@
 \subsection{Persistent Storage}
+We use leveldb. To guarantee persistence, we enforce writing to disk
+all operations before returning. This slows performance down, but
+prevents loss of data that the client wants committed. Speed could be
+gained using a Harp-like set of pointers, but this would be
+complicated and the client would have to understand the modified
+persistence model.
 
 \subsubsection{Recovery from failures}
+
+When one server contacts another via the FetchRecovery RPC, both
+servers enter a recovering state. This state means that neither are
+allowed to modify their databases \emph{except} through the recovery
+process. Depending on how many replicas there are, this may render a
+replica group unable to process requests while recovery is underway.
+If there are enough replicas so that paxos majority can still be
+reached, then the recovering servers will catch up once they come back
+online, using the functionality already built in to sync lagging paxos
+servers.
+
+All recovery processes send over both the relevant portions of the k/v
+store, as well as paxos state and any state needed to detect
+duplicates.
+
+\textbf{Single server crashing and restarting, disk in-tact}: The
+recovery process contacts a random replica in the same replica group.
+The recovering server tells the up-to-date server its current state,
+and the up-to-date server responds with any state the recoverer is
+missing. This works because the server only discards old decisions if
+all participants have marked that sequence number as 'done'.
+
+\textbf{Single server crashing and disk lost}: The recovery process
+contacts a random replica in the same replica group, and tells the
+up-to-date server that it has no state. The two servers open an
+auxillary TCP connection to send over the relevant portions of the
+database. This avoids RPC overhead, since the database might be large.
+The database has an in-order iterator, in case the TCP connection
+fails, the recoverer can re-establish the connection and ask the
+sender to pick up where it left off. If the sending server fails, then
+the recovering server asks a new replica. This allows for a failure
+mode where the original server remians locked permanently. We have
+decided that this is preferable to sacrificing consistency, and the
+system administartor can resolve the problem by rebooting the hing
+server.
+
+\textbf{Complete crash of all servers}: If all servers lose their disk
+state, then all will restart and begin like a brand new deployment. If
+the disks are not lost, then the recovery process will read the paxos
+state written to disk and pick up where it left off. Because we write
+to disk before any operation commits, data is not lost.
 
@@
 \section{Conclusion}
-We did a project.
+This project implements, tests, and benchmarks a persistent and
+performant sharded key/value store. While it is not the next Spanner,
+the end result is reasonable considering the time and manpower
+limitations.
 

### 2014-05-11 writeup/paper.tex

--- 2014-05-11 writeup/paper.tex
@@
 These two options, combined with the native cache of the operating system, allow many needed values to be read quickly from memory rather than always consulting the disk.
+
+When one server contacts another via the FetchRecovery RPC, both
+servers enter a recovering state. This state means that neither are
+allowed to modify their databases \emph{except} through the recovery
+process. Depending on how many replicas there are, this may render a
+replica group unable to process requests while recovery is underway.
+If there are enough replicas so that paxos majority can still be
+reached, then the recovering servers will catch up once they come back
+online, using the functionality already built in to sync lagging paxos
+servers.
+
+All recovery processes send over both the relevant portions of the k/v
+store, as well as paxos state and any state needed to detect
+duplicates.
+
+\textbf{Single server crashing and restarting, disk in-tact}: The
+recovery process contacts a random replica in the same replica group.
+The recovering server tells the up-to-date server its current state,
+and the up-to-date server responds with any state the recoverer is
+missing. This works because the server only discards old decisions if
+all participants have marked that sequence number as 'done'.
+
+\textbf{Single server crashing and disk lost}: The recovery process
+contacts a random replica in the same replica group, and tells the
+up-to-date server that it has no state. The two servers open an
+auxillary TCP connection to send over the relevant portions of the
+database. This avoids RPC overhead, since the database might be large.
+The database has an in-order iterator, in case the TCP connection
+fails, the recoverer can re-establish the connection and ask the
+sender to pick up where it left off. If the sending server fails, then
+the recovering server asks a new replica. This allows for a failure
+mode where the original server remians locked permanently. We have
+decided that this is preferable to sacrificing consistency, and the
+system administartor can resolve the problem by rebooting the hing
+server.
+
+\textbf{Complete crash of all servers}: If all servers lose their disk
+state, then all will restart and begin like a brand new deployment. If
+the disks are not lost, then the recovery process will read the paxos
+state written to disk and pick up where it left off. Because we write
+to disk before any operation commits, data is not lost.
+
+We use leveldb. To guarantee persistence, we enforce writing to disk
+all operations before returning. This slows performance down, but
+prevents loss of data that the client wants committed. Speed could be
+gained using a Harp-like set of pointers, but this would be
+complicated and the client would have to understand the modified
+persistence model.
 

### 2014-05-11 writeup/paper.tex
commented some stuff out in the writeup

@@
 
-\subsubsection{Memory and Performance Considerations}
-There are a number of design choices which were made in order to address both memory management and overall performance.
-
-Firstly, the service must be able to handle database content which is too large to fit in volatile memory.
-Towards this end, values are written to disk immediately as described above so it does not need to be kept in memory.
-More importantly, however, servers that need to transfer shards (either because the shard master decided a shard needs to move or because a server is recovering) must be able to do so in multiple RPCs since memory is too small to create a reply with the entire database.
-Therefore, a shard transfer occurrs with the following protocol:
-\begin{itemize}
-\item The requester sends a request to a peer, including a request number, and the current configuration number.  If doing a shard transfer, the shard number is also included.
-\item If the receiver is currently at the appropriate configuration, it accepts the request.
-It will then begin populating a reply with client response data, a list of seen operations, and the desired key/value data.
-While doing this, it will periodically check memory usage; if it exceeds a configurable threshold, it will reply with whatever it currently has and mark the reply as incomplete.
-\item The requester copies the data from the reply to its disk.  If the reply was marked as incomplete, it sends another request to the same peer with an incremented request number.
-\item This continues until the reply is marked as complete, at which point the requester launches a goroutine to make sure the sender receives an acknowledgement of the receipt of the final data message.
-\item If at any point the requester decides that the contacted peer is dead, it launches an acknowledgement goroutine as previously described for that peer (to terminate the transfer in case the peer wakes up or is not dead) and moves on to try a different peer, resetting the request number to 0.
-\end{itemize}
-Both the requester and the sender ignore client requests until the transfer is complete - this ensures that the requester will receive a consistent snapshot of the data (and is the reason the requester launches acknowledgement goroutines since otherwise a peer erroneously declared dead would be halted forever).
-
-One aspect which addresses performance is that progress can still be made while recovery is taking place as long as a majority of servers are not recovering.
-For example, if a replica group has three servers and a shard transfer is ocurring between two of them (perhaps because one just restarted), a client can still have its requests processed if it contacts the third peer.
-This is because the Paxos service runs independently of the shard key/value service; although the latter will ignore client requests on the transferring servers, their Paxos processes can still respond and make progress.
-Since the Paxos instances are stored persistently on disk, the two occupied servers can catch themselves up from the log when they complete the transfer.
-
-In addition, effort was made to utilize memory in addition to the disk in order to increase performance.
-The LevelDB database has the option to use a least-recently-used (LRU) cache to store recently used entries in memory.
-In addition, a configuration variable was added to the service to allow values to be written to memory as well as disk until the memory is full.
-These two options, combined with the native cache of the operating system, allow many needed values to be read quickly from memory rather than always consulting the disk.
-
-When one server contacts another via the FetchRecovery RPC, both
-servers enter a recovering state. This state means that neither are
-allowed to modify their databases \emph{except} through the recovery
-process. Depending on how many replicas there are, this may render a
-replica group unable to process requests while recovery is underway.
-If there are enough replicas so that paxos majority can still be
-reached, then the recovering servers will catch up once they come back
-online, using the functionality already built in to sync lagging paxos
-servers.
-
-All recovery processes send over both the relevant portions of the k/v
-store, as well as paxos state and any state needed to detect
-duplicates.
-
 \textbf{Single server crashing and restarting, disk in-tact}: The
@@
 
-We use leveldb. To guarantee persistence, we enforce writing to disk
-all operations before returning. This slows performance down, but
-prevents loss of data that the client wants committed. Speed could be
-gained using a Harp-like set of pointers, but this would be
-complicated and the client would have to understand the modified
-persistence model.
+%% \subsubsection{Memory and Performance Considerations}
+%% There are a number of design choices which were made in order to address both memory management and overall performance.
+
+%% Firstly, the service must be able to handle database content which is too large to fit in volatile memory.
+Towards this end, values are written to disk immediately as described above so it does not need to be kept in memory.
+More importantly, however, servers that need to transfer shards (either because the shard master decided a shard needs to move or because a server is recovering) must be able to do so in multiple RPCs since memory is too small to create a reply with the entire database.
+Therefore, a shard transfer occurrs with the following protocol:
+%% \begin{itemize}
+%% \item The requester sends a request to a peer, including a request number, and the current configuration number.  If doing a shard transfer, the shard number is also included.
+%% \item If the receiver is currently at the appropriate configuration, it accepts the request.
+It will then begin populating a reply with client response data, a list of seen operations, and the desired key/value data.
+While doing this, it will periodically check memory usage; if it exceeds a configurable threshold, it will reply with whatever it currently has and mark the reply as incomplete.
+%% \item The requester copies the data from the reply to its disk.  If the reply was marked as incomplete, it sends another request to the same peer with an incremented request number.
+%% \item This continues until the reply is marked as complete, at which point the requester launches a goroutine to make sure the sender receives an acknowledgement of the receipt of the final data message.
+%% \item If at any point the requester decides that the contacted peer is dead, it launches an acknowledgement goroutine as previously described for that peer (to terminate the transfer in case the peer wakes up or is not dead) and moves on to try a different peer, resetting the request number to 0.
+%% \end{itemize}
+%% Both the requester and the sender ignore client requests until the transfer is complete - this ensures that the requester will receive a consistent snapshot of the data (and is the reason the requester launches acknowledgement goroutines since otherwise a peer erroneously declared dead would be halted forever).
+
+%% One aspect which addresses performance is that progress can still be made while recovery is taking place as long as a majority of servers are not recovering.
+For example, if a replica group has three servers and a shard transfer is ocurring between two of them (perhaps because one just restarted), a client can still have its requests processed if it contacts the third peer.
+This is because the Paxos service runs independently of the shard key/value service; although the latter will ignore client requests on the transferring servers, their Paxos processes can still respond and make progress.
+Since the Paxos instances are stored persistently on disk, the two occupied servers can catch themselves up from the log when they complete the transfer.
+
+%% In addition, effort was made to utilize memory in addition to the disk in order to increase performance.
+The LevelDB database has the option to use a least-recently-used (LRU) cache to store recently used entries in memory.
+In addition, a configuration variable was added to the service to allow values to be written to memory as well as disk until the memory is full.
+These two options, combined with the native cache of the operating system, allow many needed values to be read quickly from memory rather than always consulting the disk.
+
+%% When one server contacts another via the FetchRecovery RPC, both
+%% servers enter a recovering state. This state means that neither are
+%% allowed to modify their databases \emph{except} through the recovery
+%% process. Depending on how many replicas there are, this may render a
+%% replica group unable to process requests while recovery is underway.
+%% If there are enough replicas so that paxos majority can still be
+%% reached, then the recovering servers will catch up once they come back
+%% online, using the functionality already built in to sync lagging paxos
+%% servers.
+
+%% All recovery processes send over both the relevant portions of the k/v
+%% store, as well as paxos state and any state needed to detect
+%% duplicates.
+
+%% We use leveldb. To guarantee persistence, we enforce writing to disk
+%% all operations before returning. This slows performance down, but
+%% prevents loss of data that the client wants committed. Speed could be
+%% gained using a Harp-like set of pointers, but this would be
+%% complicated and the client would have to understand the modified
+%% persistence model.
 

### 2014-05-11 src/test/test_test.go
modify disk test params

@@ -405,7 +405,11 @@ func paddedRandIntString(size int) string {
 	data := make([]byte, size)
 	s := strconv.Itoa(rand.Int())
 	for i := range data {
-		data[i] = s[i % len(s)]
+		if i < len(s) {
+			data[i] = s[i]
+		} else {
+			data[i] = byte(rand.Int31() & 0xff)
+		}
 	}
 	return string(data[:])
 }
@@ -413,7 +417,7 @@ func paddedRandIntString(size int) string {
 func TestDiskRecovery(t *testing.T) {
 	nclients := 2
 //	numBytes := 20971520 //20MB
-	nItems := 100000
+	nItems := 10000
 	keySize := 32
 	valSize := 4096
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)

### 2014-05-11 writeup/paper.tex
Some paper edits

@@
 \subsection{Testing}
-We added about 25 new tests, and extended existing ones to work with
-persistent storage and run on TCP sockets instead of unix file
-sockets. We use a few flags and configurations to run different modes
-(e.g. with and without persistence, RPC using unix file sockets or
-localhost TCP).
+About 25 new tests were created, and existing ones were extended, in order to evaluate persistent storage and run on TCP sockets instead of unix file sockets.
+A few flags and configuration variables can select different modes (e.g.
+with and without persistence, RPC using unix file sockets or localhost TCP).
 
-Below we describe the different tests. All tests are located at
-'mexos/src/$<$name$>$/test\_test.go' unless otherwise noted.
+Below are summaries of different located at 'mexos/src/\emph{$<$name$>$}/test\_test.go' unless otherwise noted.
 
-\emph{paxos:} Paxos specific tests
+\emph{paxos:} Paxos specific tests which include restarting servers deterministically or randomly.
+These include cases where servers lose their disks, where all servers restart, where partitioning changes, where communication is unreliable, and many combinations of the above.
+There are also benchmarks for measuring performance.
 
-\emph{shardmaster:} Shardmaster specific tests
+\emph{shardmaster:} Shardmaster specific tests which include restarting servers with or without disk to ensure that persistence and recovery work as expected.
+There are also benchmarks for measuring performance.
 
-\emph{shardkv:} Shardkv specific tests
+\emph{shardkv:} Shardkv specific tests which include deterministically restarting servers to ensure that persistence and recovery operate correctly.
+There is also a test which performs the previously written tests but with random reboots of random servers or entire groups.
+There are also benchmarks for measuring shard transfer speed.
 
-\emph{test:} Tests and benchmarks modified to run on our AWS
-cluster. These tests are the most complicated, and require the user to
-ssh into our cluster leader and start the servers on other cluster
-machines using 'start-on-all.sh' to run
-'mexos/src/main/start.go'. These tests required a surprising amount of
-debugging and glue code. Additionally, to check for correct operation
-under server failure, we had to make a backdoor RPC that told the
-remote server to act deaf or dead.
+\emph{test:} Tests and benchmarks modified to run on our AWS cluster.
+Running these tests are the most complicated, and require the user to SSH into the AWS cluster leader and start the servers on other cluster machines using 'start-on-all.sh' to run 'mexos/src/main/start.go'.
+These tests required a surprising amount of debugging and glue code.
+Additionally, to check for correct operation under server failure, backdoor RPCs were required to tell a remote server to act deaf or dead.
 
-We present the results of our benchmarks in a later section.
+The results of these benchmarks are presented below.
 
@@
 
-We created an extensive system to try and accurately measure
-performance. In addition to writing many benchmarks, we also modified
-the original RPC system so that we could run the servers in an AWS
-cluster and observe the performance on a real network.
+An extensive testing system was created to try and accurately measure performance.
+This comprises many new benchmarks, as well as a modified RPC system to run the servers in an AWS cluster and observe the performance on a real network.
 
@@
 
-We compared performance between vanilla paxos and multipaxos under a
-variety of conditions. The tests were run on an actual network in an
-AWS cluster.
+We compared performance between vanilla paxos and multipaxos under a variety of conditions. The tests were run on an actual network in an AWS cluster.
 

### 2014-05-11 src/test/test_test.go
test parameter

@@ -415,11 +415,11 @@ func paddedRandIntString(size int) string {
 }
 
 func TestDiskRecovery(t *testing.T) {
-	nclients := 2
+	nclients := 1
 //	numBytes := 20971520 //20MB
-	nItems := 10000
+	nItems := 68
 	keySize := 32
-	valSize := 4096
+	valSize := 1500 * 1024
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	

### 2014-05-11 src/test/test_test.go
tweak test params

@@ -417,9 +417,9 @@ func paddedRandIntString(size int) string {
 func TestDiskRecovery(t *testing.T) {
 	nclients := 1
 //	numBytes := 20971520 //20MB
-	nItems := 68
+	nItems := 20352
 	keySize := 32
-	valSize := 1500 * 1024
+	valSize := 5 * 1024
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	

### 2014-05-11 src/test/test_test.go
update param

@@ -417,9 +417,9 @@ func paddedRandIntString(size int) string {
 func TestDiskRecovery(t *testing.T) {
 	nclients := 1
 //	numBytes := 20971520 //20MB
-	nItems := 20352
+	nItems := 2046
 	keySize := 32
-	valSize := 5 * 1024
+	valSize := 50 * 1024
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	

### 2014-05-11 src/test/test_test.go
update param

@@ -417,9 +417,9 @@ func paddedRandIntString(size int) string {
 func TestDiskRecovery(t *testing.T) {
 	nclients := 1
 //	numBytes := 20971520 //20MB
-	nItems := 2046
+	nItems := 1023
 	keySize := 32
-	valSize := 50 * 1024
+	valSize := 100 * 1024
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	

### 2014-05-11 src/test/test_test.go
update params

@@ -417,9 +417,9 @@ func paddedRandIntString(size int) string {
 func TestDiskRecovery(t *testing.T) {
 	nclients := 1
 //	numBytes := 20971520 //20MB
-	nItems := 1023
+	nItems := 682
 	keySize := 32
-	valSize := 100 * 1024
+	valSize := 150 * 1024
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	

### 2014-05-11 src/shardkv/common.go
fetch not implemented yet

@@ -57,20 +57,24 @@ type FetchReply struct {
 }
 
 type RecoverArgs struct {
-	Config  int
-	Shard   int
-	Exclude map[string]bool
-	Sender  string
+	Address  string
+	ShardNum int
+	Resume   bool
+	LastKey  string
 }
 
 type RecoverReply struct {
 	MinSeq        int
 	CurrentConfig shardmaster.Config
-	Store         map[string]string
+
 	Response      map[int64]string
 	Seen          map[int64]bool
 	Err           bool
-	Complete      bool
+}
+
+type RecoverDoneArgs struct {
+}
+type RecoverDoneReply struct {
 }
 
 func hash(s string) uint32 {

### 2014-05-11 src/shardkv/server.go

@@ -8,8 +8,11 @@ import "time"
 import "paxos"
 import "sync"
 import "os"
+import "errors"
 
-//import "io"
+import "io"
+import "bufio"
+import "encoding/binary"
 import "syscall"
 import "encoding/gob"
 import "math/rand"
@@ -42,6 +45,7 @@ const dbCacheSize = 20                         // Size of database cache in MB (
 const memoryLimit = 100                        // Memory limit in MB
 const memoryThreshold = memoryLimit * 75 / 100 // When to stop filling memory (when to abort a Fetch RPC and use multiple messages)
 const recoveryRetryDelay = 500                 // Time in ms to wait before resending acknowledgments
+const recoveryPortNum = 3222
 
 // Will use these to check that dbCacheSize doesn't overflow an int
 // (int size is either 32 or 64 bits depending on implementation)
@@ -86,6 +90,7 @@ type ShardKV struct {
 	me         int
 	unreliable bool // for testing
 	network    bool
+	servers    []string
 
 	// ShardKV state
 	sm       *shardmaster.Clerk
@@ -107,7 +112,6 @@ type ShardKV struct {
 	recovering     bool
 	sending        bool
 	sendingTo      string
-	shardIterator  *levigo.Iterator
 }
 
 // Write the desired key/value to memory and/or disk
@@ -418,7 +422,8 @@ func (kv *ShardKV) Fetch(args *FetchArgs, reply *FetchReply) error {
 	for kv.sending && args.Sender != kv.sendingTo {
 		time.Sleep(10 * time.Millisecond)
 	}
-	return kv.fetchHandler(args, reply)
+//	return kv.fetchHandler(args, reply)
+	return nil
 }
 
 // Respond to acknowledgement that Fetch is complete
@@ -429,111 +434,6 @@ func (kv *ShardKV) FetchComplete(args *FetchArgs, reply *FetchReply) error {
 	DPrintf("\n%v.%v: Marking sending complete", kv.gid, kv.me)
 	reply.Complete = true
 	//}
-	return nil
-}
-
-// This helper "fetch" method now exists because both Fetch
-// and FetchRecovery use it, and Fetch must wait for recovery
[…]

### 2014-05-11 src/test/test_test.go
update params

@@ -417,9 +417,9 @@ func paddedRandIntString(size int) string {
 func TestDiskRecovery(t *testing.T) {
 	nclients := 1
 //	numBytes := 20971520 //20MB
-	nItems := 682
+	nItems := 409
 	keySize := 32
-	valSize := 150 * 1024
+	valSize := 250 * 1024
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	

### 2014-05-11 src/test/test_test.go
test params

@@ -417,9 +417,9 @@ func paddedRandIntString(size int) string {
 func TestDiskRecovery(t *testing.T) {
 	nclients := 1
 //	numBytes := 20971520 //20MB
-	nItems := 409
+	nItems := 204
 	keySize := 32
-	valSize := 250 * 1024
+	valSize := 500 * 1024
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	

### 2014-05-11 src/test/test_test.go
params

@@ -417,9 +417,9 @@ func paddedRandIntString(size int) string {
 func TestDiskRecovery(t *testing.T) {
 	nclients := 1
 //	numBytes := 20971520 //20MB
-	nItems := 204
+	nItems := 99
 	keySize := 32
-	valSize := 500 * 1024
+	valSize := 1024 * 1024
 	smPorts, gids, kvPorts := setup("basic", false, numGroups, numReplicas)
 	//defer clean()
 	

### 2014-05-11 src/shardkv/server.go
timing info for recovery

@@ -1406,7 +1406,18 @@ func (kv *ShardKV) dbInit() {
 	}
 }
 
+func trace(s string) (string, time.Time) {
+    fmt.Println("START:", s)
+    return s, time.Now()
+}
+
+func un(s string, startTime time.Time) {
+    endTime := time.Now()
+    fmt.Println("  END:", s, "ElapsedTime in seconds:", endTime.Sub(startTime))
+}
+
 func (kv *ShardKV) startup(servers []string) {
+	defer un(trace("recovery time "))
 	defer func() {
 		kv.recovering = false
 		log.Printf("\n%v-%v Marked recovery false", kv.gid, kv.me)

### 2014-05-12 writeup/paper.tex
Additional paper tweaking

@@
 
-We compared performance between vanilla paxos and multipaxos under a variety of conditions. The tests were run on an actual network in an AWS cluster.
+Performance tests were run using both vanilla paxos and multipaxos under a variety of conditions. The tests were run on an actual network in an AWS cluster.
 
@@
 
-\textit{System throughput}: Number of requests per second the entire
-system can handle. This is different than the client throughput
-because clients always wait to see the response of the current request
-before dispatching another.
+\textit{System throughput}: Number of requests per second the entire system can handle.
+This is different than the client throughput because clients always wait to see the response of the current request before dispatching another.
 
-System throughput requires a saturation of multiple clients sending
-continuous requests over a ten second period. The system saturated
-from 10 to 50 clients, sepending on the number of
-groups/replicas. After the saturation point, adding more clients
-causes a decrease in throughput occurs because of network traffic
-saturation.
+System throughput requires a saturation of multiple clients sending continuous requests for a certain amount of time (here, ten seconds).
+The system saturated from 10 to 50 clients, depending on the number of groups/replicas.
+After the saturation point, adding more clients
+causes a decrease in throughput because of network traffic saturation.
 
-We also ran some tests repeatedly on one key, and others on a random
-set of keys. This allows us to look at how the performance changes
-when requests all go to the same replica group and when they are
-evenly spread across all replica groups (assuming the staff-provided
-key2shard isn't biased).
+Some tests were repeatedly run on one key, and others on a random set of keys.
+This helps indicate how the performance changes when requests all go to the same replica group versus when they are evenly spread across all replica groups.
 
@@
 % If you use beamer only pass ``xcolor=table'' option, i.e. \documentclass[xcolor=table]{beamer}
-\begin{table}[h]
+\begin{table}
 \begin{tabular}{|l|l|l|l|}
@@
 \multicolumn{4}{|c|}{\cellcolor[HTML]{C0C0C0}{\color[HTML]{000000} \textbf{Vanilla Paxos}}}                           \\ \hline
-                              & \textbf{Avg. Client Latency} & \textbf{Avg. Client Thruput} & \textbf{System Thruput} \\ \hline
+                              & \textbf{Avg. Client Latency} & \textbf{Avg. Client Throughput} & \textbf{System Throughput} \\ \hline
 \multicolumn{4}{|c|}{\textit{PutHash to one key}}                                                                     \\ \hline
@@
 \multicolumn{4}{|c|}{\cellcolor[HTML]{C0C0C0}{\color[HTML]{000000} \textbf{Multipaxos}}}                              \\ \hline
-                              & \textbf{Avg. Client Latency} & \textbf{Avg. Client Thruput} & \textbf{System Thruput} \\ \hline
+                              & \textbf{Avg. Client Latency} & \textbf{Avg. Client Throughput} & \textbf{System Throughput} \\ \hline
 \multicolumn{4}{|c|}{\textit{PutHash to one key}}                                                                     \\ \hline
@@
 
-For a single group case, the difference between three replicas and ten is
-striking. The more replicas, the more traffic is needed for a paxos
-agreement. Figure 1 demonstrates how throughput suffers as the number
-of replicas increases. One other interesting effect is that as the
-number of replicas increases, the larger the gap there is between the
-single key and multi-key cases. Because there is one group, all puts
-touch the same servers. We are unsure what causes this.
+It can be seen that, in all cases, the Paxos optimizations had a significant positive impact on both latency and client and system throughput.
+For example, system throughput increased by an average of over 71\%.
 
-Adding more groups increases throughput, however. For evenly
-distributed traffic, going from one group to 3 nearly triples the
-throughput. Figure 2 shows that this trend continues, so adding groups
-linearly improves the throughput of the system. The table shows that
-average throughput for a single client may suffer, but this is
-probably because the system throughput benchmark creates many clients,
-while the client throughput test creates a single client. Having
-multiple clients increases the liklihood that a client must wait to
-see its request served, since the replicas may be busy serving another
-client's request.
+For a single group case, the difference between three replicas and ten is striking.
+With more replicas, more traffic is needed for a paxos agreement and thus latency increases.
+Figure 1 demonstrates how throughput suffers as the number of replicas increases.
+Another interesting effect is that as the number of replicas increases, the gap between the single key and multi-key cases increases dramatically even when there is a single group (and therefore all puts touch the same servers).
 
-\begin{figure}[H]
+\begin{figure}[h]
 \centering
 \makebox[\textwidth][c]{
-\includegraphics[scale=.50]{replicas.png}}
+\includegraphics[scale=.40]{replicas.png}}
 \caption{Performance of a one-group configuration rapidly deteriorates as the number of replicas increases}
@@
 
-\begin{figure}[H]
+Adding more groups, however, increases throughput.
+For evenly distributed traffic, going from one group to 3 nearly triples the throughput when using multiple keys (thereby leveraging the additional groups).
+Figure 2 shows that this trend continues, such that adding groups linearly improves the throughput of the system.
+The table shows that average throughput for a single client may suffer, probably because the system throughput benchmark creates many clients, while the client throughput test creates a single client; having multiple clients increases the liklihood that a client must wait to see its request served since the replicas may be busy serving another client's request.
+
+\begin{figure}[h]
 \centering
 \makebox[\textwidth][c]{
-\includegraphics[scale=.5]{groups.png}}
+\includegraphics[scale=.4]{groups.png}}
 \caption{Performance increases linearly as the number of groups increases. There is probably a saturation point, but we didn't have enough cluster machines to reach it.}
@@
 
-Multipaxos decreases the number of RPCs needed to reach agreement.
+Utilizing optimizations similar to Multipaxos decreases the number of RPCs needed to reach agreement.
 
@@
 % If you use beamer only pass ``xcolor=table'' option, i.e. \documentclass[xcolor=table]{beamer}
-\begin{table}[H]
+\begin{table}[h]
 \centering
@@
 \multicolumn{3}{|c|}{\cellcolor[HTML]{C0C0C0}\textbf{Paxos RPC Counts}}                  \\ \hline
-                                 & \textbf{Single Proposer} & \textbf{Multiple Proposer} \\ \hline
+                                 & \textbf{Single Proposer} & \textbf{Multiple Proposers} \\ \hline
 \textbf{Normal Paxos}            & 30                       & 89            \\ \hline
@@
 \subsection{Bottlenecks}
-Although we made significant improvements performance-wise, we
-identified a number of bottlenecks.
+Although the above enhancements significantly improved performance, various were identified which could be addressed in the future.
 
-One notable bottleneck is disk writes. The database library we used,
-levedb, uses a cache and is quite fast. However, in the process of
-debugging networked RPCs, we implemented simple logging. This logging
-slowed down the operations by nearly a factor of ten. This was a good
-demonstration of how care must be taken in even the most mundane
-aspects, such as logging.
+One notable bottleneck is disk reading and writing.
+The database library wrapper, LeveDB, offers a cache as well as compression and is quite fast.
+However, benchmark tests revealed that scanning the database to compile data for shard transfers accounted for a significant portion of total transfer time.
+Additionally, in the process of debugging networked RPCs, simple logging was implemented which slowed down operations by nearly a factor of ten; this was a good demonstration of disk latency and how care must be taken in even mundane aspects of the system.
+In the future, the system topology could be somewhat altered to allow for more parallelizable recovery and transfer procedures to help minimize the effect of disk latency (for example, a topology inspired by FDS).
 
-Another bottleneck is network capacity and/or number of messages. As
-shown in the latency and thoughput section, the increasing the number
-of replicas has an adverse effect on performance since each operation
-requires communication with a majority of replicas.
+Another bottleneck is network capacity and/or number of messages.
+As shown in the latency and thoughput section, increasing the number of replicas has an adverse effect on performance since each operation requires communication with a majority of replicas.
+The paxos optimizations therefore helped to improve performance, and further optimizations could be made to the data transfer protocols to reduce the required number of messages.
 
-Increasing the number of replica groups increases the performance of
-the system for normal operation, especially under heavy load with
-evenly distributed key requests. This is especially noticable in the
-system throughput metric.  The number of groups roughly corresponds to
-how parallelizable the operations are.
+Increasing the number of replica groups increases the performance of the system for normal operation, especially under heavy load with evenly distributed key requests.
+This is especially noticable in the system throughput metric.
+This arises since number of groups roughly corresponds to how parallelizable the operations are.
 
-Using multipaxos reduces the amount of network communication required,
-which improves the throughput and latency by about a factor of
-two. Multipaxos also reduces the amount of paxos state information,
-since prepares are pre-determined. This saves some disk space.  We
-also suspect that using UDP would also improve throughput, as servers
-that do not recieve a response already re-transmit their request,
-making the TCP ack system redundant.
+Using multipaxos reduces the amount of network communication required, which improves the throughput and latency by about a factor of two.
+Multipaxos also saves some disk space by reducing the amount of stored paxos state information since prepares are pre-determined.
+
+It is also suspected that using UDP would improve throughput, as servers that do not recieve a response already re-transmit their request, making the TCP acknowledgement system redundant.
 
 \section{Conclusion}
-This project implements, tests, and benchmarks a persistent and
-performant sharded key/value store. While it is not the next Spanner,
-the end result is reasonable considering the time and manpower
-limitations.
+This project implements, tests, and benchmarks a persistent and performant sharded key/value store.
+While there are various improvements and further testing that were not yet implemented, the current system comprises a working model which demonstrates the effectiveness of its various protocols and optimizations while revealing the bottlenecks that can be addressed in the future.
 

### 2014-05-12 src/recoveryBench/server.go
added db benchmark

@@ -1,0 +1,227 @@
+package main
+
+import "fmt"
+import "net"
+import "net/rpc"
+import "log"
+import "strconv"
+import "io"
+import "bufio"
+import "github.com/jmhodges/levigo"
+import "encoding/binary"
+import "bytes"
+import "encoding/gob"
+import "math/rand"
+
+const port = 3222
+const fill = true
+var dbReadOptions  *levigo.ReadOptions
+var dbWriteOptions *levigo.WriteOptions
+var dbOpts         *levigo.Options
+var db             *levigo.DB
+
+type Args struct {}
+type Reply struct {
+	Port int
+}
+
+func call(srv string, rpcname string, args interface{},
+        reply interface{}) bool {
+        fmt.Printf("sending tpc rpc to %s\n", srv)
+        c, errx := rpc.Dial("tcp", srv)
+        if errx != nil {
+                fmt.Printf("there a connection error!")
+                return false
+        }
+        fmt.Printf("connection opened!")
+        defer c.Close()
+	
+        err := c.Call(rpcname, args, reply)
+	if err == nil {
+		fmt.Printf("call succeeded")
+		return true
+	}
+        fmt.Printf("call failed")
+        return false
+}
+
+func Xfer(args *Args, reply *Reply) error {
+	fmt.Printf("Xfer called")
+        reply.Port = port
+	
+	go func() {
+		l, e := net.Listen("tcp", ":"+strconv.Itoa(port))
+		if e != nil {
+			log.Fatal("listen error: ", e)
+		}
+		fmt.Println("listening")
+		c, err := l.Accept()
+		if err != nil {
+			log.Fatal("Could not accept connections!")
+		}
+		defer l.Close()
+		
+		finished := false
+		
+		reader := bufio.NewReader(c)
+		buffer := make([]byte, 1024)
+		
+	readloop:
+		for !finished {
+			command, err := reader.ReadByte()
+			if err != nil {
+				fmt.Println("Recovery read error")
+				c.Close()
+				break readloop
+			}
+			switch command {
+			case 0 : // done
+				c.Close()
+				finished = true
+			case 1 : // key/value pair
+				keylen, _ := binary.ReadVarint(reader)
+				if int64(len(buffer)) < keylen {
+					buffer = make([]byte, keylen)
+				}
[…]

### 2014-05-12 src/shardkv/common.go

@@ -66,9 +66,6 @@ type RecoverArgs struct {
 type RecoverReply struct {
 	MinSeq        int
 	CurrentConfig shardmaster.Config
-
-	Response      map[int64]string
-	Seen          map[int64]bool
 	Err           bool
 }
 

### 2014-05-12 src/shardkv/server.go

@@ -38,7 +38,6 @@ var logfile *os.File
 // for accurate testing!
 const persistent = true
 const recovery = true
-const writeToMemory = false                    // Whether responses/store should be written to memory (as well as disk / disk cache)
 const dbUseCompression = true                  // Whether database should compress entries
 const dbUseCache = true                        // Whether database should use a built-in cache
 const dbCacheSize = 20                         // Size of database cache in MB (ignored if dbUseCache is false)
@@ -97,9 +96,9 @@ type ShardKV struct {
 	px       *paxos.Paxos
 	gid      int64 // my replica group ID
 	config   shardmaster.Config
-	store    map[string]string // key/value store
-	response map[int64]string  // client responses, indexed by client ID
-	seen     map[int64]bool    // which ops have been seen, indexed by op ID
+//	store    map[string]string // key/value store
+//	response map[int64]string  // client responses, indexed by client ID
+//	seen     map[int64]bool    // which ops have been seen, indexed by op ID
 	minSeq   int
 
 	// Persistence stuff
@@ -116,63 +115,41 @@ type ShardKV struct {
 
 // Write the desired key/value to memory and/or disk
 func (kv *ShardKV) putValue(key string, value string) {
-	// Write to memory if using memory
-	if writeToMemory {
-		kv.store[key] = value
-	}
 	// Write to disk if persistent is enabled
 	kv.dbPut(key, value)
 }
 
 // Get the desired value, either from memory or disk
 func (kv *ShardKV) getValue(key string) (string, bool) {
-	value, exists := kv.store[key]
-	if !exists {
-		value, exists = kv.dbGet(key)
-	}
+	value, exists := kv.dbGet(key)
 	return value, exists
 }
 
 // Write the seen opID to memory and/or disk
 func (kv *ShardKV) putSeen(opID int64, seen bool) {
 	// Write to memory if using memory
-	if writeToMemory {
-		kv.seen[opID] = seen
-	}
+	// if writeToMemory {
+	// 	kv.seen[opID] = seen
+	// }
 	// Write to disk if persistent is enabled
 	kv.dbWriteSeen(opID, seen)
 }
 
[…]

### 2014-05-12 src/test/test_test.go

@@ -429,8 +429,16 @@ func TestDiskRecovery(t *testing.T) {
 		go func(c int) {
 			ck := shardkv.MakeClerk(smPorts, true)
 			for i := 0; i < nItems; i++ {
-				ck.Put(paddedRandIntString(keySize), paddedRandIntString(valSize))
-				counts[c]++
+				k := paddedRandIntString(keySize)
+				v := paddedRandIntString(valSize)
+				var buffer bytes.Buffer
+				enc := gob.NewEncoder(&buffer)
+				err := enc.Encode(value)
+
+				err := kv.db.Put(kv.dbWriteOptions, []byte(key), buffer.Bytes())
+				if err != nil {
+					fmt.Printf("error writing to database\n")
+				}
 			}
 		}(i)
 	}

### 2014-05-12 writeup/paper.tex
Added optimization

@@
 \subsection{Paxos Optimizations}
+In order to improve the performance of Paxos, we implemented the Multi-Paxos leader selection optimization.
+This optimization solved the dueling leaders problem, preventing us from hanging forever in a loop between conflicting leaders, and also allowed us to reduce the number of RPC calls that were necessary to come to agreement, by allowing us to skip the Prepare phase of the Paxos protocol for a leader that was constant between instances.
+The main issue with the Multi-Paxos leader selection is ensuring that leadership information is propogated through the system and ensuring that we never agree as part of two separate partitions when the Prepare phase is skipped.
+Unfortunately, we discovered a number of implementation bugs with our default Paxos implementation, which we ended up fixing to end up with a fairly reliable Multi-Paxos optimized system that had a selected leader and skipped unnecessary prepare phases.
+The results were that the number of required RPCs was reduced by about 33\%, and that there were reduced bandwidth requirements due to the singular leadership.
+There was an issue that a single machine became the primary proposer, potentially limiting it based on disk and processor speeds, however this doesn't appear to be an issue at the anticipated use cases.
 

### 2014-05-12 writeup/paper.tex
Added first-draft recovery times

@@
 \subsubsection{Disk Recovery}
-Raw time to write 1GB to database, HDD:  21.470750409s
+Various tests were performed to evaluate the recovery time for servers restarting.
+The AWS servers can either have hard disk drives or solid state drives, and both were evaluated for raw disk speed by writing 1GB of data to the database.
+A HDD averaged about 47.7 MB/s while the SSD averaged about 166.6 MB/s.
+The recovery was then tested by creating a one-group service, placing 100 MB worth of data in the database, and finally restarting a replica without its disk contents.
+Various tests were run with differently sized values (and therefore differing numbers of entries).
+In all tests, the memory limit was set to under 75 MB, so that data transfer was forced to take place using multiple messages.
 
-Raw time to write 1GB to database, SSD:  6.1470524260s
+\begin{table}
+\begin{tabular}{|l|l|l|l|}
+\hline
+\multicolumn{4}{|c|}{\cellcolor[HTML]{C0C0C0}{\color[HTML]{000000} \textbf{Recovery Performance for 100MB Database}}}                           
+\\ \hline \textbf{Value Size (KB)} &  \textbf{Number of Entries} & \textbf{Time Reading Database (s)} & \textbf{Recovery Time (s)} 
+\\ \hline 5  & 20352 & 464.7 & ?
+\\ \hline 50  & 2046 & 30.8 & 37.2
+\\ \hline 100  & 1023 & 15.9 & 21.1
+\\ \hline 150  & 682 & 10.2 & 15.8
+\\ \hline 250  & 409 & 6.3 & 12.4
+\\ \hline 500  & 204 & 3.5 & 10.6
+\\ \hline 1024  & 99 & 1.88 & 8.3
+\\ \hline 1500  & 68 & 1.4 & ?
+\\ \hline
+\end{tabular}
+\end{table}
 
-Recovery time of a 1GB database using our system: TBD
+\begin{figure}[h]
+\centering
+\makebox[\textwidth][c]{
+\includegraphics[scale=.4]{recovery.png}}
+\caption{Performance increases dramatically as the number of keys is reduced (and the size of entries is increased).}
+\label{fig:recovery}
+\end{figure}
+
+In these experiments, an average of about 5 messages was sent in order to complete the shard transfers.
+In the above table, the database read time is the total time that the sender spent scanning its database to collect the data needed to be sent.
+The remaining time is spent transferring the data over the network and by the receiver writing a copy to disk.
+It can be seen that this time accounts for the vast majority of total transfer time when there are many keys, and decreases dramatically as the number of entries is reduced.
+This indicates that the main bottleneck is iteration over the database at the sender.
+The current implementation uses the built-in LevelDB iterator - future research can be done into how this operates (and into whether the database is fragmented on the drive) to try and improve performance since the aforementioned measured disk speeds indicate that it should be able to read the data much faster.
+Meanwhile, the networking time remains approximately constant regardless of the number of entries, indicating that the network bandwidth itself is the limiting factor for RPC time.
+The fact that the number of messages remained approximately constant indicates that the recovery protocol does not have significant overhead which depends on the number of keys but rather is simply limited by memory capacity as desired.
+As the number of keys decreases, the network therefore becomes the limiting factor.
+
+%Raw time to write 1GB to database, HDD:  21.470750409s
+
+%Raw time to write 1GB to database, SSD:  6.1470524260s
+
+%Recovery time of a 1GB database using our system: TBD
 
@@
 The database library wrapper, LeveDB, offers a cache as well as compression and is quite fast.
-However, benchmark tests revealed that scanning the database to compile data for shard transfers accounted for a significant portion of total transfer time.
+However, as discussed previously, benchmark tests revealed that scanning the database to compile data for shard transfers accounted for a significant portion of total transfer time.
 Additionally, in the process of debugging networked RPCs, simple logging was implemented which slowed down operations by nearly a factor of ten; this was a good demonstration of disk latency and how care must be taken in even mundane aspects of the system.

### 2014-05-12 writeup/paper.tex
added to writeup

@@
 \section{Introduction}
-The presented project implements a persistent, fault-tolerant, high-performance key/value store.
-In order to achieve these goals, the system implements at-most-once semantics, relies upon Paxos for reaching consenus, shards the database across multiple replica groups, persistently stores values to disk, and implements protocols for automatically recovering and updating disk contents upon server reboot.
-In order to improve performance, the system features two paxos optimizations: a protocol similar to multi-paxos to avoid the dueling leaders problem, and the ability for leaders to send out prepare messages in advance to reduce the required number of RPCs for paxos agreeement.
+The presented project implements a persistent, fault-tolerant,
+high-performance key/value store. In order to achieve these goals, the
+system implements at-most-once semantics, relies upon Paxos for
+reaching consenus, shards the database across multiple replica groups,
+persistently stores values to disk, and implements protocols for
+automatically recovering and updating disk contents upon server
+reboot.  In order to improve performance, the system features two
+paxos optimizations: a protocol similar to multi-paxos to avoid the
+dueling leaders problem, and the ability for leaders to send out
+prepare messages in advance to reduce the required number of RPCs for
+paxos agreeement.
 
-Extensive testing was done to evaluate the system's persistence, recovery, and performance by extending existing tests and creating new ones.
-After modifying the local RPC system to work on a real network, the system was deployed on an Amazon Web Services (AWS) cluster to obtain realistic measurements of throughput, latency, and disk recovery speed.
+Extensive testing was done to evaluate the system's persistence,
+recovery, and performance by extending existing tests and creating new
+ones.  After modifying the local RPC system to work on a real network,
+the system was deployed on an Amazon Web Services (AWS) cluster to
+obtain realistic measurements of throughput, latency, and disk
+recovery speed.
 
-\section{Design} \label{sec:design}
-The design is based upon the code developed throughout the semester for lab 4.
-There are therefore three basic components running on each server: Paxos processes to reach consenus for instances, a shard master to determine which groups are responsible for various shards of the database, and a shard key/value storage process to execute client Put/Get requests.
-The Paxos instances are used to create a consistent log of operations for both shard reconfiguration and client operations.
+\section{Design} \label{sec:design} The design is based upon the code
+developed throughout the semester for lab 4.  There are therefore
+three basic components running on each server: Paxos processes to
+reach consenus for instances, a shard master to determine which groups
+are responsible for various shards of the database, and a shard
+key/value storage process to execute client Put/Get requests.  The
+Paxos instances are used to create a consistent log of operations for
+both shard reconfiguration and client operations.
 
-Battery tests were run to determine which combination of our prior labs created the fastest and most reliable starting point.
-Once a working codebase was compiled, modifications were made to address three  major areas: Paxos optimizations, persistence (with disk recovery), and testing over a real network on AWS.
+Battery tests were run to determine which combination of our prior
+labs created the fastest and most reliable starting point.  Once a
+working codebase was compiled, modifications were made to address
+three major areas: Paxos optimizations, persistence (with disk
+recovery), and testing over a real network on AWS.
 
 \subsection{Paxos Optimizations}
-In order to improve the performance of Paxos, we implemented the Multi-Paxos leader selection optimization.
-This optimization solved the dueling leaders problem, preventing us from hanging forever in a loop between conflicting leaders, and also allowed us to reduce the number of RPC calls that were necessary to come to agreement, by allowing us to skip the Prepare phase of the Paxos protocol for a leader that was constant between instances.
-The main issue with the Multi-Paxos leader selection is ensuring that leadership information is propogated through the system and ensuring that we never agree as part of two separate partitions when the Prepare phase is skipped.
-Unfortunately, we discovered a number of implementation bugs with our default Paxos implementation, which we ended up fixing to end up with a fairly reliable Multi-Paxos optimized system that had a selected leader and skipped unnecessary prepare phases.
-The results were that the number of required RPCs was reduced by about 33\%, and that there were reduced bandwidth requirements due to the singular leadership.
-There was an issue that a single machine became the primary proposer, potentially limiting it based on disk and processor speeds, however this doesn't appear to be an issue at the anticipated use cases.
+In order to improve the performance of Paxos, we implemented the
+Multi-Paxos leader selection optimization. This optimization solved
+the dueling leaders problem, preventing us from hanging forever in a
+loop between conflicting leaders, and also allowed us to reduce the
+number of RPC calls that were necessary to come to agreement, by
+allowing us to skip the Prepare phase of the Paxos protocol for a
+leader that was constant between instances. The main issue with the
+Multi-Paxos leader selection is ensuring that leadership information
+is propogated through the system and ensuring that we never agree as
+part of two separate partitions when the Prepare phase is
+skipped. Unfortunately, we discovered a number of implementation bugs
+with our default Paxos implementation, which we ended up fixing to end
+up with a fairly reliable Multi-Paxos optimized system that had a
+selected leader and skipped unnecessary prepare phases. The results
+were that the number of required RPCs was reduced by about 33\%, and
+that there were reduced bandwidth requirements due to the singular
+leadership. There was an issue that a single machine became the
+primary proposer, potentially limiting it based on disk and processor
+speeds, however this doesn't appear to be an issue at the anticipated
+use cases.
 
 \subsection{Persistent Storage}
-In order to tolerate servers restarting and to handle databases that cannot fit in memory, all three components of the system store state and data persistently on disk.
-To interface with the disk, a Go wrapper for the high-performance C++ key/value libray LevelDB was utilized.
-An outline of what each system component stores on disk is presented below.
+In order to tolerate servers restarting and to handle databases that
+cannot fit in memory, all three components of the system store state
+and data persistently on disk.  To interface with the disk, a Go
+wrapper for the high-performance C++ key/value libray LevelDB was
+utilized.  An outline of what each system component stores on disk is
+presented below.
 \begin{itemize}
-\item The Paxos processes write the highest accepted value, the highest accepted proposal number, and the highest acknowledged prepare request to disk for each instance.
-This ensures that the protocol will operate correctly even if a server restarts (for example, a decided value can never be changed).
-In addition to this state, the Paxos processes persistently store the done responses received from peers and the number of the maximum known instance (see the recovery section below).
+\item The Paxos processes write the highest accepted value, the
+  highest accepted proposal number, and the highest acknowledged
+  prepare request to disk for each instance.  This ensures that the
+  protocol will operate correctly even if a server restarts (for
+  example, a decided value can never be changed).  In addition to this
+  state, the Paxos processes persistently store the done responses
+  received from peers and the number of the maximum known instance
+  (see the recovery section below).
 
-\item The shard master writes every configuration to disk, ensuring that will be able to respond to historical queries even after restarting.
-In addition, it stores the highest executed Paxos log entry and the number of its highest known configuration.
+\item The shard master writes every configuration to disk, ensuring
+  that will be able to respond to historical queries even after
+  restarting.  In addition, it stores the highest executed Paxos log
+  entry and the number of its highest known configuration.
 
-\item The shard key/value service writes its database to disk by writing each value as each Put request is executed.
-In addition, it must preserve at-most-once semantics and consistency across reboots, so it persistently stores which operations have been seen as well as the most recent response sent to each client.
-It also writes the highest executed Paxos log entry and the current configuration number.
+\item The shard key/value service writes its database to disk by
+  writing each value as each Put request is executed.  In addition, it
+  must preserve at-most-once semantics and consistency across reboots,
+  so it persistently stores which operations have been seen as well as
+  the most recent response sent to each client.  It also writes the
+  highest executed Paxos log entry and the current configuration
+  number.
 \end{itemize}
 
-In all cases, the specified data is written to disk before returning to the caller, so any response that a caller receives from a peer will not be forgotten by the responder even if it restarts.
-It may be the case that the responder crashes before sending the response, but this is acceptable since the system offers at-most-once semantics (so the caller must be prepared to retry and the responder is prepared to filter duplicates).
+In all cases, the specified data is written to disk before returning
+to the caller, so any response that a caller receives from a peer will
+not be forgotten by the responder even if it restarts.  It may be the
+case that the responder crashes before sending the response, but this
+is acceptable since the system offers at-most-once semantics (so the
+caller must be prepared to retry and the responder is prepared to
+filter duplicates).
 
 \subsubsection{Recovery from Failures}
-In order to tolerate servers restarting with or without their disk contents, the system includes protocols for recovering and catching up on startup.
-These can be summarized by considering what happens when a server starts:
+In order to tolerate servers restarting with or without their disk
+contents, the system includes protocols for recovering and catching up
+on startup.  These can be summarized by considering what happens when
+a server starts:
 
 \begin{itemize}
-\item The Paxos process will load any saved done responses and maximum instance number from its disk.
-It will then contact peers to request their done responses and maximum instance number, determining which peer is the most updated out of any peers that respond.
-From these values, both the ones read from disk and the ones received, it can determine which instances it should know about but does not; it then requests these instances from its peers.
-\item The shard master process will load the highest known configuration and highest executed Paxos log instance from disk if it exists.
-It will then request the same information from its peers, determining who among the responders is the most updated.
-From this information it can determine if there are any missed configurations, and if so request them from a peer.
-\item The shard key/value service will load its last known configuration and the highest executed Paxos log instance from disk if it exists.
-It will then request the same information from peers, determining the most updated state that it should have.
-Finally, it can request the actual database key/value entries from a peer (see below for details of how this transfer is performed) along with any client responses and lists of seen operations.
+\item The Paxos process will load any saved done responses and maximum
+  instance number from its disk.  It will then contact peers to
+  request their done responses and maximum instance number,
+  determining which peer is the most updated out of any peers that
+  respond.  From these values, both the ones read from disk and the
+  ones received, it can determine which instances it should know about
+  but does not; it then requests these instances from its peers.
+\item The shard master process will load the highest known
+  configuration and highest executed Paxos log instance from disk if
+  it exists.  It will then request the same information from its
+  peers, determining who among the responders is the most updated.
+  From this information it can determine if there are any missed
+  configurations, and if so request them from a peer.
+\item The shard key/value service will load its last known
+  configuration and the highest executed Paxos log instance from disk
+  if it exists.  It will then request the same information from peers,
+  determining the most updated state that it should have.  Finally, it
+  can request the actual database key/value entries from a peer (see
+  below for details of how this transfer is performed) along with any
+  client responses and lists of seen operations.
 \end{itemize}
-These protocols allow any number of servers restarting with their disk contents to catch themselves up from any updated peer.
-They also allow any number of servers restarting without their disk contents to recover the needed data from any peer that has the data.
+These protocols allow any number of servers restarting with their disk
+contents to catch themselves up from any updated peer.  They also
+allow any number of servers restarting without their disk contents to
+recover the needed data from any peer that has the data.
 
-\textbf{Single server crashing and restarting, disk in-tact}: The
-recovery process contacts a random replica in the same replica group.
-The recovering server tells the up-to-date server its current state,
-and the up-to-date server responds with any state the recoverer is
-missing. This works because the server only discards old decisions if
-all participants have marked that sequence number as 'done'.
+\textbf{Single server crashing and restarting, disk in-tact or lost}:
+The server will contact its peers to figure out the state it is
+missing, and it will then send RPCs to collect the missing state. The
+server it is doing state transfer with cannot change disk state until
+the entire process is done. If the disk is lost, the request indicates
+that it has no state, which causes the entire database to be sent.
 
-\textbf{Single server crashing and disk lost}: The recovery process
-contacts a random replica in the same replica group, and tells the
-up-to-date server that it has no state. The two servers open an
-auxillary TCP connection to send over the relevant portions of the
-database. This avoids RPC overhead, since the database might be large.
-The database has an in-order iterator, in case the TCP connection
-fails, the recoverer can re-establish the connection and ask the
-sender to pick up where it left off. If the sending server fails, then
-the recovering server asks a new replica. This allows for a failure
-mode where the original server remians locked permanently. We have
-decided that this is preferable to sacrificing consistency, and the
-system administartor can resolve the problem by rebooting the hing
-server.
-
-\textbf{Complete crash of all servers}: If all servers lose their disk
-state, then all will restart and begin like a brand new deployment. If
-the disks are not lost, then the recovery process will read the paxos
-state written to disk and pick up where it left off. Because we write
-to disk before any operation commits, data is not lost.
+\textbf{Complete crash of all servers}: Servers read their state from
+disk, and then contact peers to figure out if they are out of date.
 
 %% \subsubsection{Memory and Performance Considerations}
-%% There are a number of design choices which were made in order to address both memory management and overall performance.
+%% There are a number of design choices which were made in order to
+%% address both memory management and overall performance.
 
-%% Firstly, the service must be able to handle database content which is too large to fit in volatile memory.
-Towards this end, values are written to disk immediately as described above so it does not need to be kept in memory.
-More importantly, however, servers that need to transfer shards (either because the shard master decided a shard needs to move or because a server is recovering) must be able to do so in multiple RPCs since memory is too small to create a reply with the entire database.
-Therefore, a shard transfer occurrs with the following protocol:
+%% Firstly, the service must be able to handle database content which
+%% is too large to fit in volatile memory.  Towards this end, values
+%% are written to disk immediately as described above so it does not
+%% need to be kept in memory.  More importantly, however, servers that
+%% need to transfer shards (either because the shard master decided a
+%% shard needs to move or because a server is recovering) must be able
+%% to do so in multiple RPCs since memory is too small to create a
+%% reply with the entire database.  Therefore, a shard transfer
+%% occurrs with the following protocol:
 %% \begin{itemize}
-%% \item The requester sends a request to a peer, including a request number, and the current configuration number.  If doing a shard transfer, the shard number is also included.
-%% \item If the receiver is currently at the appropriate configuration, it accepts the request.
-It will then begin populating a reply with client response data, a list of seen operations, and the desired key/value data.
-While doing this, it will periodically check memory usage; if it exceeds a configurable threshold, it will reply with whatever it currently has and mark the reply as incomplete.
-%% \item The requester copies the data from the reply to its disk.  If the reply was marked as incomplete, it sends another request to the same peer with an incremented request number.
-%% \item This continues until the reply is marked as complete, at which point the requester launches a goroutine to make sure the sender receives an acknowledgement of the receipt of the final data message.
-%% \item If at any point the requester decides that the contacted peer is dead, it launches an acknowledgement goroutine as previously described for that peer (to terminate the transfer in case the peer wakes up or is not dead) and moves on to try a different peer, resetting the request number to 0.
+%% \item The requester sends a request to a peer, including a request
+%%   number, and the current configuration number.  If doing a shard
+%%   transfer, the shard number is also included.
+%% \item If the receiver is currently at the appropriate
+%%   configuration, it accepts the request.  It will then begin
+%%   populating a reply with client response data, a list of seen
+%%   operations, and the desired key/value data.  While doing this, it
+%%   will periodically check memory usage; if it exceeds a
+%%   configurable threshold, it will reply with whatever it currently
+%%   has and mark the reply as incomplete.
+%% \item The requester copies the data from the reply to its disk.  If
+%%   the reply was marked as incomplete, it sends another request to
+%%   the same peer with an incremented request number.
+%% \item This continues until the reply is marked as complete, at
+%%   which point the requester launches a goroutine to make sure the
+%%   sender receives an acknowledgement of the receipt of the final
+%%   data message.
+%% \item If at any point the requester decides that the contacted peer
+%%   is dead, it launches an acknowledgement goroutine as previously
+%%   described for that peer (to terminate the transfer in case the
+%%   peer wakes up or is not dead) and moves on to try a different
+%%   peer, resetting the request number to 0.
 %% \end{itemize}
-%% Both the requester and the sender ignore client requests until the transfer is complete - this ensures that the requester will receive a consistent snapshot of the data (and is the reason the requester launches acknowledgement goroutines since otherwise a peer erroneously declared dead would be halted forever).
+%% Both the requester and the sender ignore client requests until the
+%% transfer is complete - this ensures that the requester will receive
+%% a consistent snapshot of the data (and is the reason the requester
+%% launches acknowledgement goroutines since otherwise a peer
+%% erroneously declared dead would be halted forever).
 
-%% One aspect which addresses performance is that progress can still be made while recovery is taking place as long as a majority of servers are not recovering.
-For example, if a replica group has three servers and a shard transfer is ocurring between two of them (perhaps because one just restarted), a client can still have its requests processed if it contacts the third peer.
-This is because the Paxos service runs independently of the shard key/value service; although the latter will ignore client requests on the transferring servers, their Paxos processes can still respond and make progress.
-Since the Paxos instances are stored persistently on disk, the two occupied servers can catch themselves up from the log when they complete the transfer.
+One aspect which addresses performance is that progress can still
+be made while recovery is taking place as long as a majority of
+servers are not recovering.  For example, if a replica group has
+three servers and a shard transfer is ocurring between two of them
+(perhaps because one just restarted), a client can still have its
+requests processed if it contacts the third peer.  This is because
+the Paxos service runs independently of the shard key/value
+service; although the latter will ignore client requests on the
+transferring servers, their Paxos processes can still respond and
+make progress.  Since the Paxos instances are stored persistently
+on disk, the two occupied servers can catch themselves up from the
+log when they complete the transfer.
 
-%% In addition, effort was made to utilize memory in addition to the disk in order to increase performance.
-The LevelDB database has the option to use a least-recently-used (LRU) cache to store recently used entries in memory.
-In addition, a configuration variable was added to the service to allow values to be written to memory as well as disk until the memory is full.
-These two options, combined with the native cache of the operating system, allow many needed values to be read quickly from memory rather than always consulting the disk.
+In addition, effort was made to utilize memory in addition to the
+disk in order to increase performance.  The LevelDB database has
+the option to use a least-recently-used (LRU) cache to store
+recently used entries in memory.  In addition, a configuration
+variable was added to the service to allow values to be written to
+memory as well as disk until the memory is full.  These two
+options, combined with the native cache of the operating system,
+allow many needed values to be read quickly from memory rather than
+always consulting the disk.
 
-%% When one server contacts another via the FetchRecovery RPC, both
-%% servers enter a recovering state. This state means that neither are
-%% allowed to modify their databases \emph{except} through the recovery
-%% process. Depending on how many replicas there are, this may render a
-%% replica group unable to process requests while recovery is underway.
-%% If there are enough replicas so that paxos majority can still be
-%% reached, then the recovering servers will catch up once they come back
-%% online, using the functionality already built in to sync lagging paxos
-%% servers.
+When one server contacts another via the FetchRecovery RPC, both
+servers enter a recovering state. This state means that neither are
+allowed to modify their databases \emph{except} through the
+recovery process. Depending on how many replicas there are, this
+may render a replica group unable to process requests while
+recovery is underway.  If there are enough replicas so that paxos
+majority can still be reached, then the recovering servers will
+catch up once they come back online, using the functionality
+already built in to sync lagging paxos servers.
 
-%% All recovery processes send over both the relevant portions of the k/v
-%% store, as well as paxos state and any state needed to detect
-%% duplicates.
+All recovery processes send over both the relevant portions of the
+k/v store, as well as paxos state and any state needed to detect
+duplicates.
 
-%% We use leveldb. To guarantee persistence, we enforce writing to disk
-%% all operations before returning. This slows performance down, but
-%% prevents loss of data that the client wants committed. Speed could be
-%% gained using a Harp-like set of pointers, but this would be
-%% complicated and the client would have to understand the modified
+%% We use leveldb. To guarantee persistence, we enforce writing to
+%% disk all operations before returning. This slows performance down,
+%% but prevents loss of data that the client wants committed. Speed
+%% could be gained using a Harp-like set of pointers, but this would
+%% be complicated and the client would have to understand the modified
 %% persistence model.
@@
 \makebox[\textwidth][c]{
-\includegraphics[scale=.4]{recovery.png}}
+\includegraphics[scale=.7]{recovery.png}}
 \caption{Performance increases dramatically as the number of keys is reduced (and the size of entries is increased).}
@@
 
-In these experiments, an average of about 5 messages was sent in order to complete the shard transfers.
-In the above table, the database read time is the total time that the sender spent scanning its database to collect the data needed to be sent.
-The remaining time is spent transferring the data over the network and by the receiver writing a copy to disk.
-It can be seen that this time accounts for the vast majority of total transfer time when there are many keys, and decreases dramatically as the number of entries is reduced.
-This indicates that the main bottleneck is iteration over the database at the sender.
-The current implementation uses the built-in LevelDB iterator - future research can be done into how this operates (and into whether the database is fragmented on the drive) to try and improve performance since the aforementioned measured disk speeds indicate that it should be able to read the data much faster.
-Meanwhile, the networking time remains approximately constant regardless of the number of entries, indicating that the network bandwidth itself is the limiting factor for RPC time.
-The fact that the number of messages remained approximately constant indicates that the recovery protocol does not have significant overhead which depends on the number of keys but rather is simply limited by memory capacity as desired.
-As the number of keys decreases, the network therefore becomes the limiting factor.
+In these experiments, an average of about 5 messages was sent in order
+to complete the shard transfers.  In the above table, the database
+read time is the total time that the sender spent scanning its
+database to collect the data needed to be sent.  The remaining time is
+spent transferring the data over the network and by the receiver
+writing a copy to disk.  It can be seen that this time accounts for
+the vast majority of total transfer time when there are many keys, and
+decreases dramatically as the number of entries is reduced.  This
+indicates that the main bottleneck is iteration over the database at
+the sender.  The current implementation uses the built-in LevelDB
+iterator - future research can be done into how this operates (and
+into whether the database is fragmented on the drive) to try and
+improve performance since the aforementioned measured disk speeds
+indicate that it should be able to read the data much faster.
+Meanwhile, the networking time remains approximately constant
+regardless of the number of entries, indicating that the network
+bandwidth itself is the limiting factor for RPC time.  The fact that
+the number of messages remained approximately constant indicates that
+the recovery protocol does not have significant overhead which depends
+on the number of keys but rather is simply limited by memory capacity
+as desired.  As the number of keys decreases, the network therefore
+becomes the limiting factor.
 
@@
 \subsection{Bottlenecks}
-Although the above enhancements significantly improved performance, various were identified which could be addressed in the future.
+Although the above enhancements significantly improved performance,
+various were identified which could be addressed in the future.
 
-One notable bottleneck is disk reading and writing.
-The database library wrapper, LeveDB, offers a cache as well as compression and is quite fast.
-However, as discussed previously, benchmark tests revealed that scanning the database to compile data for shard transfers accounted for a significant portion of total transfer time.
-Additionally, in the process of debugging networked RPCs, simple logging was implemented which slowed down operations by nearly a factor of ten; this was a good demonstration of disk latency and how care must be taken in even mundane aspects of the system.
-In the future, the system topology could be somewhat altered to allow for more parallelizable recovery and transfer procedures to help minimize the effect of disk latency (for example, a topology inspired by FDS).
+One notable bottleneck is disk reading and writing.  The database
+library wrapper, LeveDB, offers a cache as well as compression and is
+quite fast.  However, as discussed previously, benchmark tests
+revealed that scanning the database to compile data for shard
+transfers accounted for a significant portion of total transfer time.
+Additionally, in the process of debugging networked RPCs, simple
+logging was implemented which slowed down operations by nearly a
+factor of ten (it was disabled for benchmarks); this was a good
+demonstration of disk latency and how care must be taken in even
+mundane aspects of the system.  In the future, the system topology
+could be somewhat altered to allow for more parallelizable recovery
+and transfer procedures to help minimize the effect of disk latency
+(for example, a topology inspired by FDS).
 
-Another bottleneck is network capacity and/or number of messages.
-As shown in the latency and thoughput section, increasing the number of replicas has an adverse effect on performance since each operation requires communication with a majority of replicas.
-The paxos optimizations therefore helped to improve performance, and further optimizations could be made to the data transfer protocols to reduce the required number of messages.
+For disk recovery, we also examined the possibility of using a raw TCP
+stream to send the data, instead of using RPCs, in order to reduce
+overhead. The transfer rate for a 1GB database was about 4.5
+Mbps. This was approximately equivalent to the data rate at which the
+database itself was created locally. Thus, the overhead necessary to
+read and write values to the database is a key bottleneck.
 
-Increasing the number of replica groups increases the performance of the system for normal operation, especially under heavy load with evenly distributed key requests.
-This is especially noticable in the system throughput metric.
-This arises since number of groups roughly corresponds to how parallelizable the operations are.
+Another bottleneck is network capacity and/or number of messages.  As
+shown in the latency and thoughput section, increasing the number of
+replicas has an adverse effect on performance since each operation
+requires communication with a majority of replicas.  The paxos
+optimizations therefore helped to improve performance, and further
+optimizations could be made to the data transfer protocols to reduce
+the required number of messages.
 
-Using multipaxos reduces the amount of network communication required, which improves the throughput and latency by about a factor of two.
-Multipaxos also saves some disk space by reducing the amount of stored paxos state information since prepares are pre-determined.
+Increasing the number of replica groups increases the performance of
+the system for normal operation, especially under heavy load with
+evenly distributed key requests. This is especially noticable in the
+system throughput metric.  This arises since number of groups roughly
+corresponds to how parallelizable the operations are.
 
-It is also suspected that using UDP would improve throughput, as servers that do not recieve a response already re-transmit their request, making the TCP acknowledgement system redundant.
+Using multipaxos reduces the amount of network communication required,
+which improves the throughput and latency by about a factor of
+two. Multipaxos also saves some disk space by reducing the amount of
+stored paxos state information since prepares are pre-determined.
+
+It is also suspected that using UDP would improve throughput, as
+servers that do not recieve a response already re-transmit their
+request, making the TCP acknowledgement system redundant.
 
 \section{Conclusion}
-This project implements, tests, and benchmarks a persistent and performant sharded key/value store.
-While there are various improvements and further testing that were not yet implemented, the current system comprises a working model which demonstrates the effectiveness of its various protocols and optimizations while revealing the bottlenecks that can be addressed in the future.
+This project implements, tests, and benchmarks a persistent and
+performant sharded key/value store.  While there are various
+improvements and further testing that were not yet implemented, the
+current system comprises a working model which demonstrates the
+effectiveness of its various protocols and optimizations while
+revealing the bottlenecks that can be addressed in the future.
 

### 2014-05-12 src/recoveryBench/server.go

@@ -1,0 +1,236 @@
+package main
+
+import "fmt"
+import "net"
+import "net/rpc"
+import "log"
+import "strconv"
+import "io"
+import "bufio"
+import "github.com/jmhodges/levigo"
+import "encoding/binary"
+import "bytes"
+import "encoding/gob"
+import "math/rand"
+
+const port = 3222
+const fill = false
+var dbReadOptions  *levigo.ReadOptions
+var dbWriteOptions *levigo.WriteOptions
+var dbOpts         *levigo.Options
+var db             *levigo.DB
+
+type Args struct {}
+type Reply struct {
+	Port int
+}
+
+type Blank struct{}
+
+func call(srv string, rpcname string, args interface{},
+        reply interface{}) bool {
+        fmt.Printf("sending tpc rpc to %s\n", srv)
+        c, errx := rpc.Dial("tcp", srv)
+        if errx != nil {
+                fmt.Printf("there a connection error!")
+                return false
+        }
+        fmt.Printf("connection opened!")
+        defer c.Close()
+	
+        err := c.Call(rpcname, args, reply)
+	if err == nil {
+		fmt.Printf("call succeeded")
+		return true
+	}
+        fmt.Printf("call failed")
+        return false
+}
+
+func (bl *Blank) Xfer(args *Args, reply *Reply) error {
+	fmt.Printf("Xfer called\n")
+        reply.Port = port
+	
+	go func() {
+		l, e := net.Listen("tcp", ":"+strconv.Itoa(port))
+		if e != nil {
+			log.Fatal("listen error: ", e)
+		}
+
+		c, err := l.Accept()
+		if err != nil {
+			log.Fatal("Could not accept connections!")
+		}
+		fmt.Println("listening\n")
+
+		defer l.Close()
+		
+		finished := false
+		
+		reader := bufio.NewReader(c)
+		buffer := make([]byte, 1024)
+
+		log.Printf("Starting recovery\n")
+	readloop:
+		for !finished {
+			command, err := reader.ReadByte()
+			if err != nil {
+				fmt.Println("Recovery read error")
+				c.Close()
+				break readloop
+			}
+			switch command {
+			case 0 : // done
+				c.Close()
+				finished = true
+			case 1 : // key/value pair
+				keylen, _ := binary.ReadVarint(reader)
[…]

### 2014-05-12 writeup/paper.tex
Some paper edits, added some citations

--- 2014-05-12 writeup/paper.tex
@@
 The presented project implements a persistent, fault-tolerant, high-performance key/value store.
-In order to achieve these goals, the system implements at-most-once semantics, relies upon Paxos for reaching consenus, shards the database across multiple replica groups, persistently stores values to disk, and implements protocols for automatically recovering and updating disk contents upon server reboot.
-In order to improve performance, the system features two paxos optimizations: a protocol similar to multi-paxos to avoid the dueling leaders problem, and the ability for leaders to send out prepare messages in advance to reduce the required number of RPCs for paxos agreeement.
+In order to achieve these goals, the system implements at-most-once semantics, relies upon Paxos for reaching consensus, shards the database across multiple replica groups, persistently stores values to disk, and implements protocols for automatically recovering and updating disk contents upon server reboot.
+In order to improve performance, the system features two paxos optimizations: a protocol similar to multi-paxos to avoid the dueling leaders problem, and the ability for leaders to send out prepare messages in advance to reduce the required number of RPCs for paxos agreement.
 
@@
 The design is based upon the code developed throughout the semester for lab 4.
-There are therefore three basic components running on each server: Paxos processes to reach consenus for instances, a shard master to determine which groups are responsible for various shards of the database, and a shard key/value storage process to execute client Put/Get requests.
+There are therefore three basic components running on each server: Paxos processes to reach consensus for instances, a shard master to determine which groups are responsible for various shards of the database, and a shard key/value storage process to execute client Put/Get requests.
 The Paxos instances are used to create a consistent log of operations for both shard reconfiguration and client operations.
@@
 \subsection{Paxos Optimizations}
-In order to improve the performance of Paxos, we implemented the Multi-Paxos leader selection optimization.
-This optimization solved the dueling leaders problem, preventing us from hanging forever in a loop between conflicting leaders, and also allowed us to reduce the number of RPC calls that were necessary to come to agreement, by allowing us to skip the Prepare phase of the Paxos protocol for a leader that was constant between instances.
-The main issue with the Multi-Paxos leader selection is ensuring that leadership information is propogated through the system and ensuring that we never agree as part of two separate partitions when the Prepare phase is skipped.
-Unfortunately, we discovered a number of implementation bugs with our default Paxos implementation, which we ended up fixing to end up with a fairly reliable Multi-Paxos optimized system that had a selected leader and skipped unnecessary prepare phases.
-The results were that the number of required RPCs was reduced by about 33\%, and that there were reduced bandwidth requirements due to the singular leadership.
-There was an issue that a single machine became the primary proposer, potentially limiting it based on disk and processor speeds, however this doesn't appear to be an issue at the anticipated use cases.
+In order to improve the performance of Paxos, the system features the Multi-Paxos leader selection optimization.
+This optimization solved the dueling leaders problem, preventing infinite loops between conflicting leaders, and also reduced the number of RPC calls necessary to reach agreement by skipping the Prepare phase of the Paxos protocol for a leader that was constant between instances.
+The main issue with the Multi-Paxos leader selection is ensuring that leadership information is propagated through the system and ensuring that the protocol never agrees as part of two separate partitions when the Prepare phase is skipped.
+Unfortunately, a number of bugs were discovered with the pre-optimized Paxos implementation, which were fixed before producing a reliable Multi-Paxos optimized system that has a selected leader and skips unnecessary prepare phases.
+Once completed, the number of required RPCs was reduced by about 33\%, and bandwidth requirements were reduced due to the singular leadership.
+There was an issue that a single machine became the primary proposer, potentially limiting the system based on its disk and processor speeds, however this doesn't appear to be an issue at the anticipated use cases.
 
@@
 In order to tolerate servers restarting and to handle databases that cannot fit in memory, all three components of the system store state and data persistently on disk.
-To interface with the disk, a Go wrapper for the high-performance C++ key/value libray LevelDB was utilized.
+To interface with the disk, a Go wrapper \cite{levigo} for the high-performance C++ key/value library LevelDB \cite{leveldb} was utilized.
 An outline of what each system component stores on disk is presented below.
@@
 \item The Paxos processes write the highest accepted value, the highest accepted proposal number, and the highest acknowledged prepare request to disk for each instance.
-This ensures that the protocol will operate correctly even if a server restarts (for example, a decided value can never be changed).
+This ensures that the protocol will operate correctly even if a server restarts \cite{paxos} (for example, a decided value can never be changed).
 In addition to this state, the Paxos processes persistently store the done responses received from peers and the number of the maximum known instance (see the recovery section below).
@@
 
-\textbf{Single server crashing and restarting, disk in-tact}: The
+\textbf{Single server crashing and restarting, disk intact}: As described above, each
 recovery process contacts a random replica in the same replica group.
-The recovering server tells the up-to-date server its current state,
-and the up-to-date server responds with any state the recoverer is
-missing. This works because the server only discards old decisions if
-all participants have marked that sequence number as 'done'.
+The recovering server can then get any missing data by comparing states with the updated server (e.g.
+shard data, shard master configurations, and Paxos instances - note that Paxos only discards old decisions if all participants have marked that sequence number as 'done').
 
-\textbf{Single server crashing and disk lost}: The recovery process
-contacts a random replica in the same replica group, and tells the
-up-to-date server that it has no state. The two servers open an
-auxillary TCP connection to send over the relevant portions of the
-database. This avoids RPC overhead, since the database might be large.
-The database has an in-order iterator, in case the TCP connection
-fails, the recoverer can re-establish the connection and ask the
-sender to pick up where it left off. If the sending server fails, then
-the recovering server asks a new replica. This allows for a failure
-mode where the original server remians locked permanently. We have
-decided that this is preferable to sacrificing consistency, and the
-system administartor can resolve the problem by rebooting the hing
-server.
+\textbf{Single server crashing and disk lost}:  This is essentially the same as if the disk was intact, except that much more data will need to be transferred (the same protocol can be used).
+Note that when data is transferred, multiple messages may be used for a single shard if it is too large to fit in memory - the sender will periodically check memory usage while constructing the reply and mark replies as incomplete if needed.
 
-\textbf{Complete crash of all servers}: If all servers lose their disk
-state, then all will restart and begin like a brand new deployment. If
-the disks are not lost, then the recovery process will read the paxos
-state written to disk and pick up where it left off. Because we write
-to disk before any operation commits, data is not lost.
+% Is the below actually implemented?
+%The recovery process contacts a random replica in the same replica group, and tells the up-to-date server that it has no state.
+The two servers open an auxiliary TCP connection to send over the relevant portions of the database.
+This avoids RPC overhead, since the database might be large.
+The database has an in-order iterator, in case the TCP connection fails, the recoverer can re-establish the connection and ask the sender to pick up where it left off.
+If the sending server fails, then the recovering server asks a new replica.
+This allows for a failure mode where the original server remains locked permanently.
+We have decided that this is preferable to sacrificing consistency, and the system administrator can resolve the problem by rebooting the hung server.
+
+\textbf{Complete crash of all servers}: If all servers lose their disk state, then all will restart and begin like a brand new deployment (the above protocols will simply not find any updated data to transfer).
+If the disks are not lost, then the recovery process will read the various
+states written to disk and pick up where it left off.  Because disk writes take place before any operation commits, data is not lost.
+
+It should be noted that during a shard transfer (including recovery), the shard key/value service on both the sender and recoverer ignore client requests in order to ensure that consistent data is sent.
+Since their Paxos processes continue running independently, however, progress can still be made even if a majority of servers is involved in a data transfer (the decided operations will be persistently logged and the halted servers can catch up when transfer is complete).
 
@@
 More importantly, however, servers that need to transfer shards (either because the shard master decided a shard needs to move or because a server is recovering) must be able to do so in multiple RPCs since memory is too small to create a reply with the entire database.
-Therefore, a shard transfer occurrs with the following protocol:
+Therefore, a shard transfer occurs with the following protocol:
 %% \begin{itemize}
@@
 %% One aspect which addresses performance is that progress can still be made while recovery is taking place as long as a majority of servers are not recovering.
-For example, if a replica group has three servers and a shard transfer is ocurring between two of them (perhaps because one just restarted), a client can still have its requests processed if it contacts the third peer.
+For example, if a replica group has three servers and a shard transfer is occurring between two of them (perhaps because one just restarted), a client can still have its requests processed if it contacts the third peer.
 This is because the Paxos service runs independently of the shard key/value service; although the latter will ignore client requests on the transferring servers, their Paxos processes can still respond and make progress.
@@
 There is also a test which performs the previously written tests but with random reboots of random servers or entire groups.
-There are also benchmarks for measuring shard transfer speed.
+There are also benchmarks for measuring shard transfer speed and to ensure that the simulated memory limit is never exceeded.
 
@@
 Figure 2 shows that this trend continues, such that adding groups linearly improves the throughput of the system.
-The table shows that average throughput for a single client may suffer, probably because the system throughput benchmark creates many clients, while the client throughput test creates a single client; having multiple clients increases the liklihood that a client must wait to see its request served since the replicas may be busy serving another client's request.
+The table shows that average throughput for a single client may suffer, probably because the system throughput benchmark creates many clients, while the client throughput test creates a single client; having multiple clients increases the likelihood that a client must wait to see its request served since the replicas may be busy serving another client's request.
 
@@
 Additionally, in the process of debugging networked RPCs, simple logging was implemented which slowed down operations by nearly a factor of ten; this was a good demonstration of disk latency and how care must be taken in even mundane aspects of the system.
-In the future, the system topology could be somewhat altered to allow for more parallelizable recovery and transfer procedures to help minimize the effect of disk latency (for example, a topology inspired by FDS).
+It was considered adding a commit point marker (similar to the ideas presented in \cite{harp}), which would be straightforward in the current system by simply returning before the disk write and not advancing the Paxos min until the write completes, but the above measurements indicate that database scanning is much more significant than isolated accesses.
+In the future, the improvements can be made to avoid this iteration bottleneck.
+For example, the system topology could be altered to allow for more parallelizable recovery with a procedure inspired by FDS \cite{fds}.
+More simply, however, the servers can be altered to store a separate database for each shard; the entire database could then be transferred to the receiver using TCP/UDP.
+Preliminary tests on AWS showed that a 100MB database can be sent over TCP in less than one second.
+In addition, this would avoid the need for either the sender or the receiver to scan through a database, thus eliminating the major bottleneck identified previously.
 
 Another bottleneck is network capacity and/or number of messages.
-As shown in the latency and thoughput section, increasing the number of replicas has an adverse effect on performance since each operation requires communication with a majority of replicas.
+As shown in the latency and throughput section, increasing the number of replicas has an adverse effect on performance since each operation requires communication with a majority of replicas.
 The paxos optimizations therefore helped to improve performance, and further optimizations could be made to the data transfer protocols to reduce the required number of messages.
@@
 Increasing the number of replica groups increases the performance of the system for normal operation, especially under heavy load with evenly distributed key requests.
-This is especially noticable in the system throughput metric.
+This is especially noticeable in the system throughput metric.
 This arises since number of groups roughly corresponds to how parallelizable the operations are.
@@
 
-It is also suspected that using UDP would improve throughput, as servers that do not recieve a response already re-transmit their request, making the TCP acknowledgement system redundant.
+It is also suspected that using UDP would improve throughput, as servers that do not receive a response already re-transmit their request, making the TCP acknowledgement system redundant.
 
@@
 %\newpage
+\bibliographystyle{IEEEtran}
+\bibliography{IEEEabrv,paper}
 \end{document}

### 2014-05-12 writeup/paper.tex
add section about server config

--- 2014-05-12 writeup/paper.tex
@@
 
+\subsection{Server Configuration}
+
+Two Amazon AWS clusters were provisioned to test real-life performance of the system.
+The faster of the clusters used Amazon's m3.large instance type, where each server has 7.5 gigabytes of RAM, 2 Intel Xeon i5 Sandy Bridge CPU cores, and 1 32 gigabyte solid state drive.
+Real-world network performance, measured using iperf, was approximately 700 Mb/s.
+The slower of the clusters used Amazon's m1.small instance type, where each server has 1.7 gigabytes of RAM, 1 Intel Xeon Westmere core, and 1 160 gigabyte hard disk drive.
+Real-world network performance was approximately 150 Mb/s.
+Each cluster had one control node and twelve slave nodes.
+Instances were allocated on multi-tenant hardware, and nodes within a cluster were within the same region and availability zone.
+
 \subsection{Throughput and Latency}

### 2014-05-12 writeup/paper.tex
Merge with new server config section

@@
   prepare request to disk for each instance.  This ensures that the
-  protocol will operate correctly even if a server restarts (for
+  protocol will operate correctly even if a server restarts \cite{paxos} (for
   example, a decided value can never be changed).  In addition to this
@@
 \end{itemize}
-These protocols allow any number of servers restarting with their disk
-contents to catch themselves up from any updated peer.  They also
-allow any number of servers restarting without their disk contents to
-recover the needed data from any peer that has the data.
 
-\textbf{Single server crashing and restarting, disk in-tact or lost}:
-The server will contact its peers to figure out the state it is
-missing, and it will then send RPCs to collect the missing state. The
-server it is doing state transfer with cannot change disk state until
-the entire process is done. If the disk is lost, the request indicates
-that it has no state, which causes the entire database to be sent.
+These protocols allow any number of servers restarting with their disk contents to catch themselves up from any updated peer.
+They also allow any number of servers restarting without their disk contents to recover the needed data from any peer that has the data.
 
-\textbf{Complete crash of all servers}: Servers read their state from
-disk, and then contact peers to figure out if they are out of date.
+\textbf{Single server crashing and restarting, disk intact or lost}: As described above, each
+recovery process contacts a random replica in the same replica group.
+The recovering server can then get any missing data by comparing states with the updated server (e.g.
+shard data, shard master configurations, and Paxos instances - note that Paxos only discards old decisions if all participants have marked that sequence number as 'done').
+This will therefore work regardless of whether the disk was lost or simply out of date.
+Note that when data is transferred, multiple messages may be used for a single shard if it is too large to fit in memory - the sender will periodically check memory usage while constructing the reply and mark replies as incomplete if needed.
 
-%% \subsubsection{Memory and Performance Considerations}
-%% There are a number of design choices which were made in order to
-%% address both memory management and overall performance.
+% Is the below actually implemented?
+%The recovery process contacts a random replica in the same replica group, and tells the up-to-date server that it has no state.
+The two servers open an auxiliary TCP connection to send over the relevant portions of the database.
+This avoids RPC overhead, since the database might be large.
+The database has an in-order iterator, in case the TCP connection fails, the recoverer can re-establish the connection and ask the sender to pick up where it left off.
+If the sending server fails, then the recovering server asks a new replica.
+This allows for a failure mode where the original server remains locked permanently.
+We have decided that this is preferable to sacrificing consistency, and the system administrator can resolve the problem by rebooting the hung server.
 
-%% Firstly, the service must be able to handle database content which
-%% is too large to fit in volatile memory.  Towards this end, values
-%% are written to disk immediately as described above so it does not
-%% need to be kept in memory.  More importantly, however, servers that
-%% need to transfer shards (either because the shard master decided a
-%% shard needs to move or because a server is recovering) must be able
-%% to do so in multiple RPCs since memory is too small to create a
-%% reply with the entire database.  Therefore, a shard transfer
-%% occurrs with the following protocol:
-%% \begin{itemize}
-%% \item The requester sends a request to a peer, including a request
-%%   number, and the current configuration number.  If doing a shard
-%%   transfer, the shard number is also included.
-%% \item If the receiver is currently at the appropriate
-%%   configuration, it accepts the request.  It will then begin
-%%   populating a reply with client response data, a list of seen
-%%   operations, and the desired key/value data.  While doing this, it
-%%   will periodically check memory usage; if it exceeds a
-%%   configurable threshold, it will reply with whatever it currently
-%%   has and mark the reply as incomplete.
-%% \item The requester copies the data from the reply to its disk.  If
-%%   the reply was marked as incomplete, it sends another request to
-%%   the same peer with an incremented request number.
-%% \item This continues until the reply is marked as complete, at
-%%   which point the requester launches a goroutine to make sure the
-%%   sender receives an acknowledgement of the receipt of the final
-%%   data message.
-%% \item If at any point the requester decides that the contacted peer
-%%   is dead, it launches an acknowledgement goroutine as previously
-%%   described for that peer (to terminate the transfer in case the
-%%   peer wakes up or is not dead) and moves on to try a different
-%%   peer, resetting the request number to 0.
-%% \end{itemize}
-%% Both the requester and the sender ignore client requests until the
-%% transfer is complete - this ensures that the requester will receive
-%% a consistent snapshot of the data (and is the reason the requester
-%% launches acknowledgement goroutines since otherwise a peer
-%% erroneously declared dead would be halted forever).
+\textbf{Complete crash of all servers}: If all servers lose their disk state, then all will restart and begin like a brand new deployment (the above protocols will simply not find any updated data to transfer).
+If the disks are not lost, then the recovery process will read the various states written to disk and pick up where it left off.
+Because disk writes take place before any operation commits, data is not lost.
 
-One aspect which addresses performance is that progress can still
-be made while recovery is taking place as long as a majority of
-servers are not recovering.  For example, if a replica group has
-three servers and a shard transfer is ocurring between two of them
-(perhaps because one just restarted), a client can still have its
-requests processed if it contacts the third peer.  This is because
-the Paxos service runs independently of the shard key/value
-service; although the latter will ignore client requests on the
-transferring servers, their Paxos processes can still respond and
-make progress.  Since the Paxos instances are stored persistently
-on disk, the two occupied servers can catch themselves up from the
-log when they complete the transfer.
+It should be noted that during a shard transfer (including recovery), the shard key/value service on both the sender and recoverer ignore client requests in order to ensure that consistent data is sent (their databases cannot be modified except through the recovery process itself).
+Since their Paxos processes continue running independently, however, progress can still be made even if a majority of servers is involved in a data transfer (the decided operations will be persistently logged and the halted servers can catch up when transfer is complete).
+This therefore improves performance during recoveries or transfers.
 
@@
 always consulting the disk.
-
-When one server contacts another via the FetchRecovery RPC, both
-servers enter a recovering state. This state means that neither are
-allowed to modify their databases \emph{except} through the
-recovery process. Depending on how many replicas there are, this
-may render a replica group unable to process requests while
-recovery is underway.  If there are enough replicas so that paxos
-majority can still be reached, then the recovering servers will
-catch up once they come back online, using the functionality
-already built in to sync lagging paxos servers.
-
-All recovery processes send over both the relevant portions of the
-k/v store, as well as paxos state and any state needed to detect
-duplicates.
-
-%% We use leveldb. To guarantee persistence, we enforce writing to
-%% disk all operations before returning. This slows performance down,
-%% but prevents loss of data that the client wants committed. Speed
-%% could be gained using a Harp-like set of pointers, but this would
-%% be complicated and the client would have to understand the modified
-%% persistence model.
 
@@
 \subsection{Server Configuration}
-
 Two Amazon AWS clusters were provisioned to test real-life performance of the system.
@@
 \subsection{Bottlenecks}
-Although the above enhancements significantly improved performance,
-various were identified which could be addressed in the future.
 
-One notable bottleneck is disk reading and writing.  The database
-library wrapper, LeveDB, offers a cache as well as compression and is
-quite fast.  However, as discussed previously, benchmark tests
-revealed that scanning the database to compile data for shard
-transfers accounted for a significant portion of total transfer time.
-Additionally, in the process of debugging networked RPCs, simple
-logging was implemented which slowed down operations by nearly a
-factor of ten (it was disabled for benchmarks); this was a good
-demonstration of disk latency and how care must be taken in even
-mundane aspects of the system.  In the future, the system topology
-could be somewhat altered to allow for more parallelizable recovery
-and transfer procedures to help minimize the effect of disk latency
-(for example, a topology inspired by FDS).
+Although the above enhancements significantly improved performance, various were identified which could be addressed in the future.
 
-For disk recovery, we also examined the possibility of using a raw TCP
-stream to send the data, instead of using RPCs, in order to reduce
-overhead. The transfer rate for a 1GB database was about 4.5
-Mbps. This was approximately equivalent to the data rate at which the
-database itself was created locally. Thus, the overhead necessary to
-read and write values to the database is a key bottleneck.
+One notable bottleneck is disk reading and writing.
+The database library wrapper, LeveDB, offers a cache as well as compression and is quite fast.
+However, as discussed previously, benchmark tests revealed that scanning the database to compile data for shard transfers accounted for a significant portion of total transfer time.
+Additionally, in the process of debugging networked RPCs, simple logging was implemented which slowed down operations by nearly a factor of ten (this was disabled for the above benchmarks); this was a good demonstration of disk latency and how care must be taken in even mundane aspects of the system.
+It was considered adding a commit point marker (similar to the ideas presented in \cite{harp}), which would be straightforward in the current system by simply returning before the disk write and not advancing the Paxos min until the write completes, but the above measurements indicate that database scanning is much more significant than isolated accesses.
+In the future, the improvements can be made to avoid this iteration bottleneck.
+For example, the system topology could be altered to allow for more parallelizable recovery with a procedure inspired by FDS \cite{fds}.
+More simply, however, the servers can be altered to store a separate database for each shard; the entire database could then be transferred to the receiver using TCP/UDP.
+Preliminary tests on AWS showed that a 100MB database can be sent over TCP in less than one second.
+In addition, this would avoid the need for either the sender or the receiver to scan through a database, thus eliminating the major bottleneck identified previously.
 
-Another bottleneck is network capacity and/or number of messages.  As
-shown in the latency and thoughput section, increasing the number of
-replicas has an adverse effect on performance since each operation
-requires communication with a majority of replicas.  The paxos
-optimizations therefore helped to improve performance, and further
-optimizations could be made to the data transfer protocols to reduce
-the required number of messages.
+Another bottleneck is network capacity and/or number of messages.
+As shown in the latency and throughput section, increasing the number of replicas has an adverse effect on performance since each operation requires communication with a majority of replicas.
+The paxos optimizations therefore helped to improve performance, and further optimizations could be made to the data transfer protocols to reduce the required number of messages.
 
-Increasing the number of replica groups increases the performance of
-the system for normal operation, especially under heavy load with
-evenly distributed key requests. This is especially noticable in the
-system throughput metric.  This arises since number of groups roughly
-corresponds to how parallelizable the operations are.
+Increasing the number of replica groups increases the performance of the system for normal operation, especially under heavy load with evenly distributed key requests.
+This is especially noticeable in the system throughput metric.
+This arises since number of groups roughly corresponds to how parallelizable the operations are.
 
-Using multipaxos reduces the amount of network communication required,
-which improves the throughput and latency by about a factor of
-two. Multipaxos also saves some disk space by reducing the amount of
-stored paxos state information since prepares are pre-determined.
+Using multipaxos reduces the amount of network communication required, which improves the throughput and latency by about a factor of two.
+Multipaxos also saves some disk space by reducing the amount of stored paxos state information since prepares are pre-determined.
 
-It is also suspected that using UDP would improve throughput, as
-servers that do not recieve a response already re-transmit their
-request, making the TCP acknowledgement system redundant.
+It is also suspected that using UDP would improve throughput, as servers that do not receive a response already re-transmit their request, making the TCP acknowledgement system redundant.
 

### 2014-05-12 writeup/paper.tex
github link
Spell check

@@
 system implements at-most-once semantics, relies upon Paxos for
-reaching consenus, shards the database across multiple replica groups,
+reaching consensus, shards the database across multiple replica groups,
 persistently stores values to disk, and implements protocols for
@@
 prepare messages in advance to reduce the required number of RPCs for
-paxos agreeement.
+paxos agreement.
 
@@
 
+Source code for this project is accessible at \url{https://github.com/pranjalv123/mexos}.
+
 \section{Design} \label{sec:design} The design is based upon the code
@@
 three basic components running on each server: Paxos processes to
-reach consenus for instances, a shard master to determine which groups
+reach consensus for instances, a shard master to determine which groups
 are responsible for various shards of the database, and a shard
@@
 Multi-Paxos leader selection is ensuring that leadership information
-is propogated through the system and ensuring that we never agree as
+is propagated through the system and ensuring that we never agree as
 part of two separate partitions when the Prepare phase is
@@
 and data persistently on disk.  To interface with the disk, a Go
-wrapper for the high-performance C++ key/value libray LevelDB was
+wrapper for the high-performance C++ key/value library LevelDB was
 utilized.  An outline of what each system component stores on disk is
@@
 \\ \hline \textbf{Value Size (KB)} &  \textbf{Number of Entries} & \textbf{Time Reading Database (s)} & \textbf{Recovery Time (s)} 
-\\ \hline 5  & 20352 & 464.7 & ?
+\\ \hline 5  & 20352 & 464.7 & ~473
 \\ \hline 50  & 2046 & 30.8 & 37.2
@@
 \\ \hline 1024  & 99 & 1.88 & 8.3
-\\ \hline 1500  & 68 & 1.4 & ?
+\\ \hline 1500  & 68 & 1.4 & 
 \\ \hline
